---
title: "Modelling complex reporting processes"
order: 14.7
bibliography: ../nfidd.bib
---

::: {.callout-important}
## Draft session

This is a **draft** session and is still under review.
The structure, narrative, and example are in place, but the heavy model fits are not yet run as part of the course.
Every chunk that compiles or fits an `epinowcast` model is marked `#| eval: false`, so the code is shown but not executed when the site is built.
To run these chunks yourself you will need a working `cmdstanr` toolchain and the `epinowcast` package installed (see the setup callout below).
:::

# Introduction

In the [nowcasting concepts session](nowcasting) and the [joint nowcasting session](joint-nowcasting) we built our own nowcasting models in Stan.
We started from a known delay distribution, then learned to jointly estimate the delay and the nowcast from the reporting triangle, and finally added a renewal process so that the generative model of transmission informed the nowcast.

Those bespoke models were ideal for *learning* how nowcasting works because we could see every part of the model.
However, real surveillance data is usually messier than those models handle comfortably.
In this session we step back and ask what happens when the reporting process itself becomes complicated, and we introduce [`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] as a flexible, well-tested framework for exactly these situations.

## Slides

- There are no slides for this draft session yet.

## Objectives

This session aims to show why real reporting processes are harder to model than our bespoke Stan models assume, and to introduce `epinowcast` as a framework that generalises the joint nowcasting model we have already built.

::: {.callout-note collapse="true"}

# Setup

## Source file

The source file of this session is located at `sessions/complex-reporting-processes.qmd`.

## Libraries used

In this session we will use the `epinowcast` package to specify and fit nowcasting models, the `data.table` package that `epinowcast` uses for its data structures, and the `ggplot2` library for plotting.

```{r libraries, message = FALSE}
#| eval: false
library("epinowcast")
library("data.table")
library("ggplot2")
```

::: {.callout-tip}
`epinowcast` compiles models with `cmdstanr`, so you will need a working CmdStan toolchain (the same one used elsewhere in this course).
The package is on CRAN and is also available from the [epinowcast R-universe](https://epinowcast.r-universe.dev).
Installation and setup instructions are on the [package website](https://package.epinowcast.org/).
:::

## Initialisation

We set a random seed for reproducibility.

```{r initialise}
#| eval: false
set.seed(123)
```

:::

# Why real reporting is harder than our models assume

## The bespoke models we have built so far

In the [joint nowcasting session](joint-nowcasting) we modelled each cell of the reporting triangle as

$$
  n_{t,d} \mid \lambda_{t}, p_{t,d} \sim \text{Poisson}\left(\lambda_{t} \times p_{t,d}\right),
$$

where $\lambda_{t}$ is the expected number of onsets on day $t$ and $p_{t,d}$ is the probability that an onset on day $t$ is reported with a delay of $d$ days.
We modelled $\lambda_{t}$ with a geometric random walk (or a renewal process) and $p_{t,d}$ with a single delay distribution that was the same on every day.

This works well in our simulated examples because we simulated the data from almost exactly that model.
Real data rarely cooperate so neatly.

## Ways real reporting departs from this model

Consider some features of routine surveillance data that our bespoke models do not capture well.

- **Day-of-week reporting effects.**
Many reporting systems report far less at weekends and on public holidays, then catch up early in the following week.
This is a property of the *report date*, not of the underlying epidemic, and it breaks the assumption that the delay distribution depends only on the delay $d$.

- **Delays that vary over time.**
Reporting delays lengthen when a system is under strain and shorten when it has spare capacity.
A single fixed $p_{t,d}$ cannot represent a delay distribution that drifts across the outbreak.

- **Missing and late reporting.**
Some reports never arrive, or arrive far beyond the maximum delay we assumed, and weekends and holidays create gaps in the data.

- **The reporting triangle at scale.**
As soon as we stratify by age group, region, or pathogen, we have many reporting triangles at once, often sharing structure (for example, a common day-of-week effect) that we would like to model jointly.

::: {.callout-note}
## Take 5 minutes
Think about the joint nowcasting model you built.
What would you have to change in that Stan code to add a day-of-week reporting effect *and* let the delay distribution change over the course of the outbreak?
:::

::: {.callout-note collapse="true"}
## Solution

There is no single right answer, but the changes add up quickly.

- You would need to index the reporting probabilities by report date (or day of week) as well as by delay, turning the delay vector into a more complex structure.
- You would need a model for how the delay distribution changes over time, for example a random walk on the parameters of the delay distribution, with its own priors.
- You would need to keep the reporting probabilities on each day summing to one, which is awkward once they vary by date.
- Each addition makes the Stan model longer, slower to fit, and harder to debug and test.

This is the point at which writing and maintaining a bespoke model stops being the best use of our time.
A general, tested framework that already implements these features lets us focus on the modelling choices rather than the implementation.
:::

# Introducing `epinowcast`

[`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] is a Bayesian framework for real-time infectious disease surveillance.
It is a direct generalisation of the joint nowcasting model we built: it works from the reporting triangle and jointly estimates the expected counts and the reporting process, but it makes each component far more flexible and exposes them through a formula interface rather than hand-written Stan code.

`epinowcast` decomposes the problem into modular pieces, each of which mirrors something we have already seen.

- An **expectation** model for the expected counts $\lambda_{t}$ (the analogue of our geometric random walk or renewal process).
- A **reference date** model for the delay distribution by date of the epidemiological event (the analogue of our $p_{t,d}$, but allowed to vary over time and by covariates).
- A **report date** model for effects that act on the *report* date, such as day-of-week reporting effects.

Each piece is specified with an R formula, so adding a day-of-week effect or a time-varying delay is a one-line change rather than a rewrite of the model.
Under the hood `epinowcast` still builds and fits a Stan model, but the model is tested, optimised, and reused across problems.

::: {.callout-tip}
We are treating `epinowcast` as a tool here rather than dissecting its internals.
The mapping back to the bespoke model is deliberate: if you understand the joint nowcasting session, you already understand what `epinowcast` is doing, just with more flexible components.
:::

# A worked example: Germany COVID-19 hospitalisations

`epinowcast` ships with a real dataset, `germany_covid19_hosp`, of COVID-19 hospitalisation counts in Germany.
These data show exactly the complications described above, including a pronounced day-of-week reporting pattern, which makes them a good example for a more realistic reporting process.

## The data

We can load the data and look at its structure without fitting anything.

```{r load-data}
#| eval: false
library("epinowcast")
library("data.table")

germany_hosp <- germany_covid19_hosp
head(germany_hosp)
```

The dataset is in the long format that `epinowcast` expects.

- `reference_date`: the date the event refers to (here, the date of the positive test / hospitalisation).
- `report_date`: the date on which the count was reported.
- `confirm`: the cumulative count for that reference date as known on that report date.
- `location` and `age_group`: stratification variables.

For this example we take the national, all-ages series.

```{r filter-data}
#| eval: false
nat_germany_hosp <- germany_hosp[location == "DE"][age_group == "00+"]
```

::: {.callout-note}
## Take 2 minutes
How does this data structure relate to the reporting triangle from the [joint nowcasting session](joint-nowcasting)?
:::

::: {.callout-note collapse="true"}
## Solution

It is the same idea in a different layout.
Each `reference_date` is a row of the triangle, each `report_date` gives a column, and `confirm` is the cumulative version of the cell counts $n_{t,d}$ we summed over delays before.
`epinowcast` works with the cumulative counts and converts them into the reporting triangle internally during preprocessing.
:::

## Preparing a retrospective nowcast

To mimic a real-time nowcast we pretend we are standing at a particular date with only the data reported by then, and we keep the later data aside so we can check the nowcast against what was eventually reported.

```{r prepare-data}
#| eval: false
# Pretend we are standing at this date with only data reported by then
retro_germany <- nat_germany_hosp |>
  enw_filter_report_dates(latest_date = "2021-08-01") |>
  enw_filter_reference_dates(include_days = 40)

# The eventually-reported counts, kept aside to check the nowcast against
latest_obs <- nat_germany_hosp |>
  enw_filter_report_dates(latest_date = "2021-08-21") |>
  enw_latest_data()
```

`epinowcast` then preprocesses the data into its internal reporting-triangle representation, where we also choose the maximum delay we will model.

```{r preprocess}
#| eval: false
pobs <- enw_preprocess_data(retro_germany, max_delay = 30)
pobs
```

## Specifying and fitting the nowcast

Now we specify the model.
We will start from a model that captures the day-of-week reporting effect, which is the most visible complication in these data.

We build three modules.

- The **expectation** module models the expected hospitalisations over time with a random walk by week, the same multiplicative-growth idea as our geometric random walk.
- The **reference** module models the delay distribution; here we keep it simple with a single parametric delay.
- The **report** module adds a day-of-week effect on the report date, which our bespoke model could not easily express.

```{r modules}
#| eval: false
# Expected counts: a weekly random walk (cf. our geometric random walk)
expectation_module <- enw_expectation(~ rw(week), data = pobs)

# Delay distribution by reference date: a single parametric delay for now
reference_module <- enw_reference(~ 1, data = pobs)

# Report-date effects: a day-of-week reporting effect
report_module <- enw_report(~ (1 | day_of_week), data = pobs)
```

::: {.callout-note}
## Take 3 minutes
Compare these three formulas to the choices we made in the [joint nowcasting session](joint-nowcasting).
Which module corresponds to $\lambda_{t}$ and which to $p_{t,d}$, and what is genuinely new here?
:::

::: {.callout-note collapse="true"}
## Solution

- `enw_expectation()` plays the role of $\lambda_{t}$: `rw(week)` is a random walk on the expected counts, much like the geometric random walk we used for onsets.
- `enw_reference()` plays the role of the delay distribution $p_{t,d}$.
Here `~ 1` keeps it constant over time, exactly as in our bespoke model, but we could let it vary by writing, for example, `rw(week)` here too.
- The genuinely new piece is `enw_report(~ (1 | day_of_week))`: a reporting effect that depends on the *report* date rather than the delay, which is awkward to add to the hand-rolled model but is a one-line change here.
:::

We then fit the model.
This is the step that compiles and samples a Stan model, so it is slow; it is marked `eval: false` in this draft.

```{r fit}
#| eval: false
nowcast <- epinowcast(
  pobs,
  expectation = expectation_module,
  reference = reference_module,
  report = report_module,
  fit = enw_fit_opts(
    save_warmup = FALSE, pp = TRUE,
    chains = 2, iter_warmup = 500, iter_sampling = 500
  )
)
```

::: {.callout-tip}
The `enw_fit_opts()` call controls sampling in the same way the arguments to `nfidd_sample()` did in earlier sessions (number of chains, warmup and sampling iterations, and so on).
We use a small number of iterations here purely to keep the example quick; a real analysis would use more.
:::

## Inspecting and plotting the nowcast

Once fitted, the `epinowcast` object has plotting and summary methods.
Plotting against the eventually-reported data lets us see how well the nowcast filled in the recent, incomplete counts.

```{r plot-nowcast}
#| eval: false
plot(nowcast, latest_obs = latest_obs)
```

As in our bespoke nowcasts, the most recent dates are the most uncertain because they are the least complete, and the day-of-week reporting effect is now absorbed by the model rather than distorting the nowcast.

::: {.callout-note}
## Take 3 minutes
We modelled the delay distribution as constant over time with `enw_reference(~ 1, ...)`.
How would you let the delay distribution change over the outbreak, and why might that matter for these data?
:::

::: {.callout-note collapse="true"}
## Solution

You can put structure on the reference module just as on the expectation module, for example `enw_reference(~ rw(week), data = pobs)` to let the delay distribution drift week to week.
This matters when reporting speed changes over time, for instance when a system is overwhelmed during a wave and delays lengthen, or improves as processes are streamlined.
A fixed delay distribution would then mis-estimate how much of the recent data is still missing, biasing the nowcast.
This is exactly the time-varying-delay complication that was hard to add to our bespoke model and is a one-line change here.
:::

# Going further

This draft only scratches the surface of what `epinowcast` can do.
A few directions to explore once the basics are comfortable:

- **Stratified nowcasting.** Fit many reporting triangles at once (for example by age group or region) with partially pooled effects, so that sparse strata borrow strength from the others.
- **Richer delay and reporting models.** Combine time-varying delays, day-of-week effects, and parametric or non-parametric delay distributions in the same model.
- **Joint nowcasting and $R_t$ estimation.** Use a renewal-based expectation model so that the nowcast and the reproduction number are estimated together, generalising the joint model from the [joint nowcasting session](joint-nowcasting).
- **Evaluation.** Score nowcasts across many dates using proper scoring rules, as we did for forecasts in the [combining nowcasting and forecasting session](forecasting-nowcasting).

The [`epinowcast` package website](https://package.epinowcast.org/) [@epinowcastPackage2024] has vignettes working through each of these, including a fuller treatment of the Germany hospitalisations example.
For the methodological background to generative nowcasting models of this kind, see [@lisonGenerativeBayesianModeling2024], and for a comparison of nowcasting methods used in practice see [@wolfframCollaborativeNowcastingCOVID192023].

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
