---
title: "Modelling complex reporting processes"
order: 9
bibliography: ../nfidd.bib
---

# Introduction

In the [joint nowcasting session](joint-nowcasting) we built a nowcasting model in Stan that jointly estimated the delay distribution and the nowcast from the reporting triangle, and then added a renewal process so transmission informed the nowcast.
That model assumed a single delay distribution, the same on every day.

Bespoke models like that are ideal for *learning* how nowcasting works, and they remain a reasonable choice for real analyses when you understand the reporting process well.
But routine surveillance data is usually messier: the reporting process itself can be complicated in ways a single fixed delay cannot capture.
In this session we introduce [`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024], one flexible framework for these situations.
We reproduce the joint nowcasting model in `epinowcast`, then *extend* it to handle reporting complications the bespoke model cannot: a day-of-week reporting effect, a delay that drifts over time, and weekly (batched) reporting.
Throughout we simulate data with the feature, fit a model that handles it, and look at the result.

## Slides

- [Modelling complex reporting processes](slides/complex-reporting-processes)

## Objectives

This session shows why real reporting processes are harder to model than a single fixed delay assumes, reproduces the joint nowcasting model in `epinowcast`, and extends it to complex reporting features through worked examples.

::: {.callout-note collapse="true"}

# Setup

## Source file

The source file of this session is located at `sessions/complex-reporting-processes.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to load infection times and the simulation helpers we have used throughout, the `dplyr` and `tidyr` packages for data wrangling, and the `ggplot2` library for plotting.
The worked examples additionally use the `epinowcast` package and the `data.table` package that it uses for its data structures.

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
library("epinowcast")
library("data.table")
```

::: {.callout-tip}
This session uses `epinowcast` and `cmdstanr`, which we load with `library()` above.
The course's getting-started install already provides them, so there is no per-session install to run: `pak::pak("nfidd/sismid-nowcasting", dependencies = TRUE)` installs `epinowcast` and `cmdstanr` alongside the other course packages.

`epinowcast` compiles its models with `cmdstanr` the first time a model is fitted, which needs a working CmdStan toolchain (the same one used elsewhere in this course).
Full installation and setup instructions, including how to install CmdStan, are on the [package website](https://package.epinowcast.org/).
The fits in this session use the course-standard sampler settings but short, simulated data so the course builds in reasonable time.
:::

## Initialisation

We set a random seed for reproducibility.
We also tell `epinowcast` not to warn about partially specified initial conditions, which keeps the output tidy.

```{r initialise}
set.seed(123)
options(cmdstanr_warn_inits = FALSE)
```

:::

# Why real reporting is harder than our models assume

Recall the joint nowcasting model: we modelled each cell of the reporting triangle as

$$
  n_{t,d} \mid \lambda_{t}, p_{t,d} \sim \text{Poisson}\left(\lambda_{t} \times p_{t,d}\right),
$$

where $\lambda_{t}$ is the expected number of onsets on day $t$ and $p_{t,d}$ is the probability that an onset on day $t$ is reported with a delay of $d$ days.
We gave $\lambda_{t}$ a daily geometric random walk (or a renewal process) and $p_{t,d}$ a single delay distribution, the same on every day.

That worked in our simulated examples because we simulated from almost exactly that model.
Real data rarely cooperate so neatly.
Routine surveillance data often shows features a single fixed delay does not capture: day-of-week reporting effects, reporting delays that vary over time or by strata such as age or region, weekly or batched reporting, and missing or late reports.
Rather than list these and tell you they are hard, we *show* several: simulate the feature, fit a model that handles it, and look at the result.

::: {.callout-note}
## Take 5 minutes
Think about the joint nowcasting model you built.
What would you have to change in that Stan code to add a day-of-week reporting effect *and* let the delay distribution change over the course of the outbreak?
:::

::: {.callout-note collapse="true"}
## Solution

There is no single right answer, but the concrete changes add up quickly.

- **A day-of-week reporting effect.** The reporting probability would have to depend on the calendar day the report lands on, not just the delay $d$. In practice you would turn the delay vector into a matrix indexed by reference date *and* delay, $p_{t,d}$, and multiply each cell by a day-of-week factor $w_{(t+d)\,\bmod\,7}$ keyed on the *report* date $t+d$. That means seven new parameters (a `simplex` or sum-to-zero vector for identifiability) and an extra loop over cells in `transformed parameters`.
- **A time-varying delay.** You would put a random walk on the parameters of the delay distribution, for example `meanlog[t] = meanlog[t-1] + sigma * eta[t]` (and similarly for `sdlog`), each with its own prior, then recompute the delay PMF for every reference date $t$ rather than once. That is a new vector of innovations `eta`, a new standard-deviation parameter `sigma` with a prior, and a per-date PMF calculation.
- **Keeping it a valid distribution.** The probabilities for each reference date must still sum to one over delays after both changes, so you would re-impose the per-date normalisation (a simplex or an explicit renormalisation) every day instead of reusing one shared, pre-normalised vector.
- **The downside.** Each addition means more parameters, a larger `transformed parameters` block, slower gradient evaluations, and more code to unit-test, all before you have fitted anything.

None of this is impossible, and writing it yourself remains a reasonable option.
But it is the point at which reaching for a general framework that already implements these features lets us focus on the modelling choices rather than the implementation.
:::

# Introducing `epinowcast`

[`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] is a Bayesian framework for real-time infectious disease surveillance.
It is *one* of several tools for this kind of modelling (we point to others at the end of the session), and it is a direct generalisation of the joint nowcasting model we built: it works from the reporting triangle and jointly estimates the expected counts and the reporting process, but it makes each component far more flexible and exposes them through a formula interface rather than hand-written Stan code.

`epinowcast` decomposes the problem into modular pieces. Most mirror something we have already seen; the **report date** and **missing data** modules are new — they handle effects that act on the report date, such as day-of-week reporting, and reports whose reference date is unknown.

- An **expectation** model for the expected counts $\lambda_{t}$ (the analogue of our geometric random walk or renewal process).
- A **reference date** model for the delay distribution by date of the epidemiological event (the analogue of our $p_{t,d}$, but allowed to vary over time and by covariates).
- A **report date** model for effects that act on the *report* date, such as day-of-week reporting effects.
- A **missing data** model for reports whose reference date is missing or not yet known (we had no analogue in the bespoke model).
- An **observation** model for the likelihood linking expected to observed counts (the analogue of our Poisson likelihood for $n_{t,d}$).

Each piece is specified with an R formula, so adding a day-of-week effect or a time-varying delay is a one-line change rather than a rewrite of the model.
Internally, `epinowcast` still builds and fits a Stan model.

::: {.callout-tip}
We are treating `epinowcast` as a tool here rather than dissecting its internals.
The mapping back to the bespoke model is deliberate: if you understand the joint nowcasting session, you already understand what `epinowcast` is doing, just with more flexible components.
:::

# The course model, in `epinowcast`

Before adding anything new, let us reproduce the joint nowcasting model we already built, but in `epinowcast`.
We simulate exactly as in the [joint nowcasting session](joint-nowcasting): the same onset time series, the same fixed reporting delay, and the same reporting triangle.

```{r simulate-onsets}
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), ip_pmf
)
cutoff <- 71

reporting_delay_pmf <- censored_delay_pmf(
  rlnorm, max = 15, meanlog = 1, sdlog = 0.5
)

reporting_triangle <- onset_df |>
  filter(day < cutoff) |>
  mutate(
    reporting_delay = list(
      tibble(d = 0:15, reporting_delay = reporting_delay_pmf)
    )
  ) |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    reported_onsets = rpois(n(), onsets * reporting_delay)
  )
```

## Putting the data into `epinowcast` format

`epinowcast` expects data in a "long" format with one row per (reference date, report date) pair, where `confirm` is the **cumulative** count for that reference date as known on that report date.
This is the same reporting triangle we have been using, just with calendar dates instead of integer days and cumulative rather than incremental counts.

We attach arbitrary calendar dates (so that, later, weekends line up with a day-of-week effect) and let `epinowcast`'s own `enw_add_cumulative()` form the cumulative counts from the incident `new_confirm` column, rather than computing the cumulative sum by hand.
Because we reuse this transformation in each example below, we wrap it in a small helper.

Our simulation already generates every report an onset will ever receive (all delays $d = 0, \ldots, 15$), so `to_enw_format()` formats the *complete* eventually-reported data.
From that one complete data set we derive the two things we need with `epinowcast`'s own functions rather than truncating by hand:

- the data available **as of the nowcast**, which we fit to, using `enw_filter_report_dates()` to keep only reports that had arrived by the nowcast date;
- the **eventually-reported** data, which we judge the nowcast against, using `enw_latest_data()` to take the most recent report for each reference date.

After filtering to the as-of-nowcast reports we call `enw_complete_dates()` to make the reporting triangle explicitly complete, so any reference-date and delay cell that never received a report is filled with a zero rather than left missing.
Our simulation is already complete so this changes nothing here, but on real data it guards against gaps in the triangle before preprocessing.

```{r to-enw-helper}
# Day 0 of the simulation is a Monday
ref_date0 <- as.Date("2023-01-02")
# The nowcast is made once reports up to this date are in
nowcast_date <- ref_date0 + (cutoff - 1)
# Only the most recent dates are nowcast, so we zoom the result plots to a
# recent window around the nowcast horizon rather than the whole series
plot_from <- nowcast_date - 35

# Format the complete (eventually-reported) data in epinowcast's long format
to_enw_format <- function(triangle) {
  triangle |>
    transmute(
      reference_date = ref_date0 + day,
      report_date = ref_date0 + reported_day,
      new_confirm = reported_onsets
    ) |>
    arrange(reference_date, report_date) |>
    as.data.table() |>
    enw_add_cumulative()
}

# The as-of-nowcast snapshot we fit to
enw_long <- to_enw_format(reporting_triangle) |>
  enw_filter_report_dates(latest_date = nowcast_date) |>
  enw_complete_dates(max_delay = 16, missing_reference = FALSE)
head(enw_long)
```

::: {.callout-tip}
`enw_filter_report_dates()` plus `enw_latest_data()` makes the two data sets explicit, which is why we use them here.
`enw_retrospective()` is a one-shot helper that produces an as-of-nowcast snapshot and its eventual-data comparison together, and is worth knowing about for real analyses.
:::

::: {.callout-note}
## Take 2 minutes
How does this data structure relate to the reporting triangle from the [joint nowcasting session](joint-nowcasting)?
:::

::: {.callout-note collapse="true"}
## Solution

It is the same idea in a different layout.
Each `reference_date` is a row of the triangle, each `report_date` gives a column, `new_confirm` is the cell count $n_{t,d}$, and `confirm` is its cumulative version over delays.
`epinowcast` works with the cumulative `confirm` and converts it back into the reporting triangle internally during preprocessing.
:::

## Preprocessing and fitting

`epinowcast` preprocesses the data into its internal reporting-triangle representation, where we also choose the maximum delay we will model.
Note that `max_delay` counts delay *classes*, and delays are zero-indexed: `max_delay = 16` models $d = 0, \ldots, 15$, which is exactly the support we simulated from.
Setting it to 15 would quietly drop the last delay class rather than model it.

```{r preprocess}
pobs <- enw_preprocess_data(enw_long, max_delay = 16)
```

The preprocessed object has its own `plot()` method.
By default it shows the latest observed notifications by reference date (`type = "obs"`), which is the incomplete picture the nowcast will correct.

```{r preprocess-plot}
plot(pobs)
```

The same method also exposes `epinowcast`'s delay diagnostics through the `type` argument, which are worth inspecting before fitting.
`type = "delay_cumulative"` shows the empirical cumulative proportion reported by delay, so we can see how quickly reports accumulate and check that our `max_delay` is long enough to capture most of the distribution.

```{r preprocess-delay-cumulative}
plot(pobs, type = "delay_cumulative")
```

`type = "delay_fraction"` shows the same information as a heatmap of the fraction reported by delay and reference date, which makes any drift in the delay over time easier to spot.

```{r preprocess-delay-fraction}
plot(pobs, type = "delay_fraction")
```

Now we specify the model.
We define each module explicitly so the mapping back to the bespoke model is visible: a daily random walk for the expected counts (matching the daily geometric random walk we built), a single delay distribution held constant over time, and a Poisson observation model.

```{r course-modules}
# Expected counts lambda_t: a daily random walk (cf. our geometric random walk)
expectation_module <- enw_expectation(~ rw(day), data = pobs)

# Delay distribution p_{t,d} by reference date: a single parametric lognormal
# delay, constant over time (epinowcast's default reference model)
reference_module <- enw_reference(~ 1, data = pobs)

# Observation model: Poisson, as in n_{t,d} ~ Poisson(lambda_t * p_{t,d})
obs_module <- enw_obs(family = "poisson", data = pobs)
```

::: {.callout-tip .optional collapse="true"}
## Reproducing the bespoke model exactly

To match the bespoke joint-nowcast model exactly, estimate the delay nonparametrically as a free per-delay distribution (a simplex over delays) instead of the parametric lognormal:

```{r course-reference-nonparametric, eval = FALSE}
reference_module <- enw_reference(
  parametric = ~ 0, distribution = "none", non_parametric = ~ 1, data = pobs
)
```

This makes the delay a free per-delay distribution matching the bespoke model, but fits noticeably more slowly, which is why the session defaults to the parametric lognormal above.
:::

We collect the fitting options once and reuse them across the session.
These play the same role as the arguments to `nfidd_sample()` did in earlier sessions, and we set them to match: four chains run in parallel with 500 warmup and 500 sampling iterations each.

```{r fit-opts}
fit_opts <- enw_fit_opts(
  save_warmup = FALSE, pp = TRUE,
  chains = 4, parallel_chains = 4,
  iter_warmup = 500, iter_sampling = 500,
  show_messages = FALSE, refresh = 0
)
```

We then fit the model, passing each module explicitly.
The first fit compiles the Stan model, so it is the slowest step; subsequent fits in this session reuse the compiled model.

```{r course-fit, message = FALSE, results = "hide"}
course_nowcast <- epinowcast(
  pobs,
  expectation = expectation_module,
  reference = reference_module,
  obs = obs_module,
  fit = fit_opts
)
```

::: {.callout-note}
## Take 3 minutes
Compare these modules to the choices we made in the [joint nowcasting session](joint-nowcasting).
Which module corresponds to $\lambda_{t}$, which to $p_{t,d}$, and which to the likelihood?
:::

::: {.callout-note collapse="true"}
## Solution

- `enw_expectation()` plays the role of $\lambda_{t}$: `rw(day)` is a daily random walk on the expected counts, like the geometric random walk we used for onsets.
- `enw_reference(~ 1)` plays the role of the delay distribution $p_{t,d}$: a single parametric lognormal delay held constant over time. (The optional callout above shows how to recover the nonparametric free-PMF delay used in the bespoke joint-nowcast model.)
- `enw_obs(family = "poisson")` plays the role of the likelihood $n_{t,d} \sim \text{Poisson}(\lambda_{t} p_{t,d})$.

This is the *same* joint nowcasting model we built by hand, now expressed as a handful of modules.
Everything that follows is an extension of one of these modules, or the addition of a report-date module for report-date effects.
:::

## Plotting the nowcast

`epinowcast` objects come with `plot()` and `summary()` methods, so we do not have to extract draws by hand as we did with our bespoke fits.
To judge the nowcast we want to compare it against what was *eventually* reported, so we take the latest report for each reference date from the complete data with `enw_latest_data()` and pass it to `plot()` via `latest_obs`.
Building the comparison data from the complete data matters: if we built it from the truncated snapshot we fit to (`enw_long`), the "eventually observed" points would just reproduce the incomplete nowcast-era counts and the plot would not tell us whether the nowcast was any good.

Only the most recent dates are actually being nowcast, so we zoom the plot to a recent window around the nowcast horizon with `coord_cartesian()`, which restricts the view without dropping any data, rather than filtering `latest_obs`.

```{r course-plot}
latest_obs <- enw_latest_data(to_enw_format(reporting_triangle))
plot(course_nowcast, latest_obs = latest_obs) +
  coord_cartesian(xlim = c(plot_from, nowcast_date))
```

As in our bespoke nowcasts, the most recent dates are the most uncertain because they are the least complete: the nowcast (median and credible intervals) fills in the gap up to the eventually-observed counts (the points).
This gap is the right truncation the nowcast corrects for.
This default plot is the nowcast by *reference* date.
`epinowcast` can also plot the model's posterior predictions against the data that were reported at the time with `type = "posterior_prediction"`, which we facet by reference date to check the fit within each date's accumulating reports.
Faceting by *every* reference date makes each panel too small to read, so we restrict it to the most recent reference dates near the nowcast horizon, which are the ones actually being nowcast.

```{r course-plot-pp, fig.height = 7}
# Keep only the recent reference dates so each facet is legible
pp_recent <- plot(course_nowcast, type = "posterior_prediction")
pp_recent$data <- pp_recent$data[reference_date > nowcast_date - 9]
pp_recent +
  facet_wrap(vars(reference_date), scales = "free")
```

# Extending the model: day-of-week effects

Now we add a complication the bespoke model could not handle comfortably.
A day-of-week pattern can enter the data in *three* distinct places, and it helps to keep them separate because each belongs in a different module.

- A **reference-date effect on the expected counts**: events truly happen at different rates by day of week. For onset data, for example, fewer onsets may be recorded at weekends. This belongs in `enw_expectation()`.
- A **reference-date effect on the delay**: the reporting *delay distribution* itself differs by the day of week an event occurred, for example weekend onsets take longer to be reported than weekday onsets. This belongs in `enw_reference()`.
- A **report-date effect**: the reporting system processes reports at different rates by day of week, so fewer reports land at weekends regardless of when the onset occurred. This belongs in `enw_report()`.

We build a simulation with all three, one step at a time, so it is clear how each effect enters the data.
Assume day 0 of the simulation is a Monday, so day of week is `day %% 7` (0 = Monday).

We start from the underlying onsets and apply a **reference-date** day-of-week effect on the expected counts, thinning the expected onsets by a weekly weight on the onset day.

```{r dow-reference}
# Multiplicative weekly weights for Mon, Tue, ..., Sun
ref_dow_effect <- c(1, 1, 1, 1, 0.95, 0.7, 0.7) # fewer onsets at weekends
rep_dow_effect <- c(1, 1, 1, 1, 0.9, 0.4, 0.5)  # fewer reports at weekends

dow_onsets <- onset_df |>
  filter(day < cutoff) |>
  mutate(
    ref_dow = day %% 7,
    onsets = onsets * ref_dow_effect[ref_dow + 1]
  )
```

Next we apply the **reporting delay**, but now we let the delay depend on the onset day of week: weekend onsets are reported with a longer delay (a larger `meanlog`) than weekday onsets.
This is a reference-date effect on the *delay* rather than on the counts.

```{r dow-delay}
# Longer delay for weekend onsets (Sat, Sun): a reference-date delay effect
delay_meanlog_dow <- c(1, 1, 1, 1, 1, 1.4, 1.4)

dow_delayed <- dow_onsets |>
  rowwise() |>
  mutate(
    reporting_delay = list(
      tibble(
        d = 0:15,
        reporting_delay = censored_delay_pmf(
          rlnorm, max = 15,
          meanlog = delay_meanlog_dow[ref_dow + 1], sdlog = 0.5
        )
      )
    )
  ) |>
  ungroup() |>
  unnest(reporting_delay) |>
  mutate(reported_day = day + d)
```

Finally we apply a **report-date** day-of-week effect, thinning by a weekly weight on the *report* day, and draw the observed counts.

```{r dow-report}
dow_triangle <- dow_delayed |>
  mutate(
    rep_dow = reported_day %% 7,
    rep_weight = rep_dow_effect[rep_dow + 1],
    reported_onsets = rpois(n(), onsets * reporting_delay * rep_weight)
  )
```

We can see the report-date effect clearly if we sum reports by report day: the totals dip on a regular weekly cycle.

```{r dow-plot-data}
reports_by_day <- dow_triangle |>
  filter(reported_day <= max(day)) |>
  summarise(reports = sum(reported_onsets), .by = reported_day)

ggplot(reports_by_day, aes(x = reported_day, y = reports)) +
  geom_col() +
  labs(x = "Report day", y = "Reports", title = "Reports by report day")
```

::: {.callout-note}
## Take 5 minutes
Look back at the three simulation chunks (`dow-reference`, `dow-delay`, `dow-report`) and answer without running any code.

1. Every weekly weight is indexed as `[... + 1]`, for example `ref_dow_effect[ref_dow + 1]`. Why the `+ 1`?
2. In `dow-delay` the day of week enters through `meanlog`, not through a multiplicative weight. What does giving weekend onsets a larger `meanlog` do to their reporting delay?
3. The reference-date effect keys off `day %% 7` but the report-date effect keys off `reported_day %% 7`. Why does each effect use a different day?
:::

::: {.callout-note collapse="true"}
## Solution

1. `ref_dow` is `day %% 7`, which runs from 0 (Monday) to 6 (Sunday), but R vectors are indexed from 1.
   Adding 1 maps day-of-week 0 to the first weight and day-of-week 6 to the seventh.
2. `meanlog` is the log-scale mean of the lognormal delay, so a larger value shifts the whole delay distribution towards longer delays.
   Weekend onsets are therefore reported more slowly than weekday onsets, rather than simply being fewer in number.
3. The reference-date effects describe what happens on the day an event *occurred*, so they key off the onset day `day`.
   The report-date effect describes how the reporting system behaves on the day a report is *processed*, which is the onset day plus its delay, so it keys off `reported_day`.
   The same onset contributes to different report-day weights depending on how long its delay was.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: how day-of-week effects combine

```{=html}
<iframe class="nfidd-embed" src="../interactive/complex-reporting-processes.html?embed=combine" title="how day-of-week effects combine interactive" loading="lazy" style="width:100%;height:820px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Start from a smooth underlying onset curve and switch each of the three effects on and off to see how the observed reported series builds up its weekly structure.
:::

The three effects act on different things.
The reference-date effect on the counts changes *when events happen* (the onset day), so it belongs in the model for the expected counts.
The reference-date effect on the delay changes *how long reporting takes* for events on a given onset day, so it belongs in the delay (reference) module.
The report-date effect changes *when reports are processed* (the report day), so it belongs in a separate report-date module, which our bespoke $p_{t,d}$ (a function of delay $d$ alone) had no way to express.
`epinowcast` lets us add all three in one model.

These extended models have a slightly harder posterior, so for them we add `adapt_delta = 0.95` to the fitting options (just as we did for the harder joint model in the [joint nowcasting session](joint-nowcasting)).

```{r fit-opts-hard}
fit_opts_hard <- enw_fit_opts(
  save_warmup = FALSE, pp = TRUE,
  chains = 4, parallel_chains = 4,
  iter_warmup = 500, iter_sampling = 500,
  adapt_delta = 0.95,
  show_messages = FALSE, refresh = 0
)
```

```{r dow-fit, message = FALSE, results = "hide"}
dow_long <- enw_filter_report_dates(
  to_enw_format(dow_triangle), latest_date = nowcast_date
)
dow_pobs <- enw_preprocess_data(dow_long, max_delay = 16)

dow_nowcast <- epinowcast(
  dow_pobs,
  # Reference-side day-of-week effect on the expected onsets
  expectation = enw_expectation(~ rw(day) + day_of_week, data = dow_pobs),
  # Reference-side day-of-week effect on the delay distribution
  reference = enw_reference(~ day_of_week, data = dow_pobs),
  # Report-side day-of-week reporting effect
  report = enw_report(~ (1 | day_of_week), data = dow_pobs),
  obs = enw_obs(family = "poisson", data = dow_pobs),
  fit = fit_opts_hard
)
```

```{r dow-result-plot}
dow_latest <- enw_latest_data(to_enw_format(dow_triangle))
plot(dow_nowcast, latest_obs = dow_latest) +
  coord_cartesian(xlim = c(plot_from, nowcast_date))
```

As with the course model, we can also check the fit with the posterior-prediction plot, faceted by the most recent reference dates.

```{r dow-plot-pp, fig.height = 7}
pp_recent <- plot(dow_nowcast, type = "posterior_prediction")
pp_recent$data <- pp_recent$data[reference_date > nowcast_date - 9]
pp_recent +
  facet_wrap(vars(reference_date), scales = "free")
```

::: {.callout-note}
## Take 5 minutes
Read the two result plots rather than the model code.

1. In the nowcast plot (`dow-result-plot`), does the model recover the underlying onsets, and what regular pattern do you see in the nowcast median across reference dates?
2. In the posterior-prediction plot (`dow-plot-pp`), each panel is one reference (onset) date, with its reports accumulating over report dates. Compare weekday and weekend reference dates. Where does the day-of-week signal show up, and does the model track it?
:::

::: {.callout-note collapse="true"}
## Solution

1. The nowcast median follows the eventually-observed points, filling in the recent, still-incomplete dates.
   It is not a smooth curve: the median rises and falls on a regular seven-day cycle, higher mid-week and lower at weekends.
   That zigzag is the model recovering the day-of-week structure in the *onsets* after stripping out the separate day-of-week effect in *reporting*, rather than smearing the two together.
2. Within a panel the accumulation is not a straight climb; it steps up faster on some report days than others, which is the report-date effect (fewer reports processed at weekends).
   Across panels the weekend reference dates sit lower and, because weekend onsets were simulated with a longer delay, fill in more slowly than weekday reference dates.
   The posterior-prediction band covers the observed accumulation in each panel, so the three day-of-week terms together are tracking the data rather than fighting each other.

Each of these effects is one extra formula on the relevant module.
The report-date effect in particular was awkward in the bespoke Stan model, where the reporting probabilities depended only on the delay $d$ and had to sum to one for each reference date, so making them also depend on the calendar day of the report meant restructuring the whole delay component.
:::

# Extending the model: a time-varying delay

For our second worked example we relax the assumption that the delay distribution is the same on every day.
Reporting delays may lengthen when a system is under strain and shorten when it has spare capacity, so a single fixed $p_{t,d}$ cannot represent a delay that drifts across the outbreak.

We simulate this by letting the mean of the reporting delay grow slowly with the onset day, so later onsets may be reported with longer delays.

```{r tvd-simulate}
tvd_triangle <- onset_df |>
  filter(day < cutoff) |>
  rowwise() |>
  mutate(
    reporting_delay = list({
      # meanlog drifts upwards over the outbreak: delays may lengthen over time
      meanlog_t <- 1 + 0.5 * day / cutoff
      tibble(
        d = 0:15,
        reporting_delay = censored_delay_pmf(
          rlnorm, max = 15, meanlog = meanlog_t, sdlog = 0.5
        )
      )
    })
  ) |>
  ungroup() |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    reported_onsets = rpois(n(), onsets * reporting_delay)
  )
```

Before fitting anything, it is worth looking at what this drift actually does to the data.
For each reference date the mean of the reporting delay distribution is $\sum_d d\, p_{t,d}$, so we can compute it from the simulated per-date delay PMF and average it by reference week to see the delay lengthening over the outbreak.

```{r tvd-show-delay}
tvd_mean_delay <- tvd_triangle |>
  summarise(
    mean_delay = sum(d * reporting_delay) / sum(reporting_delay),
    .by = c(day)
  ) |>
  mutate(reference_week = day %/% 7) |>
  summarise(mean_delay = mean(mean_delay), .by = reference_week)

ggplot(tvd_mean_delay, aes(x = reference_week, y = mean_delay)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Reference week", y = "Mean reporting delay (days)",
    title = "The reporting delay drifts upwards over the outbreak"
  )
```

The mean delay climbs week on week, which is the feature a single fixed delay cannot represent: later onsets are reported more slowly than earlier ones, so the model has to let the delay change over time to nowcast the recent dates correctly.

::: {.callout-note}
## Take 5 minutes
The whole drift comes from one line in `tvd-simulate`: `meanlog_t <- 1 + 0.5 * day / cutoff`.

1. `meanlog` is the log-scale mean of the lognormal delay. What does letting it grow with `day` do to the most recent reference dates, and why does that make them harder to nowcast?
2. What else could you simulate by changing this one line? Sketch how you would make (a) the delay *shorten* over the outbreak, and (b) a sudden step change in the delay rather than a smooth drift. Would the `~ rw(week)` delay model still cope?
:::

::: {.callout-note collapse="true"}
## Solution

1. A larger `meanlog` shifts the whole lognormal towards longer delays, so recent onsets take longer to be fully reported.
   The most recent reference dates are therefore less complete than a fixed-delay model would assume, and it is exactly those dates the nowcast has to fill in, so getting the drift right matters most where the data are thinnest.
2. (a) A shortening delay is just a decreasing trend, for example `1.5 - 0.5 * day / cutoff`.
   (b) A step change is a branch on `day`, for example `if (day < cutoff / 2) 1 else 1.5`.
   The random walk makes no assumption about direction or shape, so it tracks a shortening delay as readily as a lengthening one and follows a smooth drift closely.
   It would lag a sharp step by a week or so, because the weekly random walk smooths, but it would still adapt rather than break.
   Changing the *spread* (`sdlog`) instead of the mean is a genuinely different feature, and you would want to check whether the delay model is flexible enough to follow a drift in the spread rather than only the average.
:::

This is a property of the *reference* (onset) date, so we put structure on the **reference module** rather than on the report module.
Where the course model held the delay constant over time, we now let the delay distribution drift week to week with `enw_reference(~ rw(week), ...)`.

::: {.callout-warning}
We simulated a roughly *linear* drift, so we could fit a model with a matching linear trend on the delay parameters and recover it almost perfectly.
That would be cheating: in a real analysis we do not know the functional form of the drift, because we did not generate the data.
We therefore deliberately choose a *flexible* model, a random walk on the delay parameters, that can track a drift without us baking in the exact form we happened to simulate.
:::

::: {.callout-tip}
## Time-varying delay versus a rolling window
A common alternative to a time-varying delay is to estimate a single delay from only the most recent stretch of data, a rolling window.
The random-walk approach here uses *all* the data with a smoothness prior, so it borrows strength across the whole series while still letting the delay change from week to week.
A rolling window instead throws away every report outside the window, so it has less data to estimate the delay from and cannot use the earlier reports at all, even though they still say something about how the delay is changing.
:::

```{r tvd-fit, message = FALSE, results = "hide"}
tvd_long <- enw_filter_report_dates(
  to_enw_format(tvd_triangle), latest_date = nowcast_date
)
tvd_pobs <- enw_preprocess_data(tvd_long, max_delay = 16)

tvd_nowcast <- epinowcast(
  tvd_pobs,
  expectation = enw_expectation(~ rw(day), data = tvd_pobs),
  # Delay distribution now drifts week to week (this is the new part)
  reference = enw_reference(~ rw(week), data = tvd_pobs),
  obs = enw_obs(family = "poisson", data = tvd_pobs),
  fit = fit_opts_hard
)
```

```{r tvd-result-plot}
tvd_latest <- enw_latest_data(to_enw_format(tvd_triangle))
plot(tvd_nowcast, latest_obs = tvd_latest) +
  coord_cartesian(xlim = c(plot_from, nowcast_date))
```

As with the course model, we can also check the fit with the posterior-prediction plot, faceted by the most recent reference dates.

```{r tvd-plot-pp, fig.height = 7}
pp_recent <- plot(tvd_nowcast, type = "posterior_prediction")
pp_recent$data <- pp_recent$data[reference_date > nowcast_date - 9]
pp_recent +
  facet_wrap(vars(reference_date), scales = "free")
```

::: {.callout-note}
## Take 3 minutes
Look at the nowcast plot (`tvd-result-plot`), where we fit `enw_reference(~ rw(week), ...)`.

1. Does the nowcast median track the eventually-observed points at the recent, still-incomplete dates?
2. What would the plot look like if we had kept a *fixed* delay when the true delay was lengthening?
:::

::: {.callout-note collapse="true"}
## Solution

1. The median follows the eventual points up to the nowcast date, and the credible interval widens for the most recent dates where fewer reports have arrived.
   `rw(week)` lets the parameters of the delay distribution change from week to week, so the model learns that recent onsets are being reported more slowly and fills the gap up to the eventual counts.
2. A fixed delay would assume recent dates are more complete than they really are, so the nowcast median would sit *below* the eventual points at the recent end and *under*-nowcast exactly where it matters most.
   This is the same flexibility we used for the expectation module, applied to the delay instead.
:::

# Extending the model: weekly (batched) reporting

For our third worked example the data are *reported* less often than they occur.
Many surveillance systems release updates weekly rather than daily, so counts arrive in weekly batches even though events happen every day.
There are two ways to handle this in `epinowcast`, and they trade resolution against simplicity.

- **Weekly reporting of weekly data ("weekly-weekly").** Aggregate everything to a weekly timestep and model a weekly process. This is the simplest and cheapest option, but it can only ever produce weekly nowcasts.
- **Weekly reporting of daily data ("weekly-daily").** Keep a daily process but tell the model that reports only land once a week, using a *structural* reporting schedule. This recovers daily resolution from weekly reports at a higher computational cost.

We show both here on the same simulated series.
`epinowcast`'s [temporal aggregation guide](https://package.epinowcast.org/articles/temporal-aggregation.html) works through these patterns in full, including a daily benchmark and a scoring comparison; we give the compact version and link there rather than reproducing all of it.

The two approaches differ most when the epidemic is changing rapidly near the nowcast date.
When the most recent, still-incomplete week is rising or falling steeply, collapsing it to a single weekly value discards the within-week trend, and the weekly-weekly model has to extrapolate that fast-moving week from a short weekly random walk with only a handful of points.
A daily process keeps the resolution and can still track the change.

To make this visible we deliberately place the nowcast in a steep decline, keeping data up to a later date than the rest of the session so the most recent week is falling fast.
This is a favourable choice made to show where the daily approach helps; we come back to how conditional the benefit is at the end of the section.
We re-simulate the same reporting process at this later cutoff and leave everything else unchanged.

```{r weekly-changepoint}
# Position the nowcast in a rapid decline, so the most recent,
# still-incomplete week is changing fast rather than flat
weekly_cutoff <- 106
weekly_nowcast_date <- ref_date0 + (weekly_cutoff - 1)
weekly_plot_from <- weekly_nowcast_date - 63

# Re-seed so this weekly reporting triangle is reproducible on its own
set.seed(123)
weekly_reporting_triangle <- onset_df |>
  filter(day < weekly_cutoff) |>
  mutate(
    reporting_delay = list(
      tibble(d = 0:15, reporting_delay = reporting_delay_pmf)
    )
  ) |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    reported_onsets = rpois(n(), onsets * reporting_delay)
  )

weekly_enw_long <- enw_filter_report_dates(
  to_enw_format(weekly_reporting_triangle),
  latest_date = weekly_nowcast_date
)
```

## Weekly-weekly

We take the change-point reporting triangle and aggregate it to a weekly reporting schedule with `enw_aggregate_cumulative()`, which sums both reference and report dates into weekly steps.

```{r weekly-aggregate}
weekly_long <- enw_aggregate_cumulative(weekly_enw_long, timestep = "week")
head(weekly_long)
```

We then preprocess with a weekly timestep, so `epinowcast` knows the reporting schedule is weekly rather than daily.
With a weekly timestep the maximum delay is now counted in weeks, so we model up to three weeks of delay.

```{r weekly-preprocess}
weekly_pobs <- enw_preprocess_data(
  weekly_long, max_delay = 3, timestep = "week"
)
plot(weekly_pobs)
```

The model is the course model again; only the timestep of the data has changed.

```{r weekly-fit, message = FALSE, results = "hide"}
weekly_nowcast <- epinowcast(
  weekly_pobs,
  expectation = enw_expectation(~ rw(week), data = weekly_pobs),
  reference = enw_reference(~ 1, data = weekly_pobs),
  obs = enw_obs(family = "poisson", data = weekly_pobs),
  fit = fit_opts_hard
)
```

As before we compare against the eventually-complete data, aggregated to the same weekly timestep.

```{r weekly-result-plot}
weekly_latest <- enw_latest_data(
  enw_aggregate_cumulative(
    to_enw_format(weekly_reporting_triangle), timestep = "week"
  )
)
plot(weekly_nowcast, latest_obs = weekly_latest) +
  coord_cartesian(xlim = c(weekly_plot_from, weekly_nowcast_date))
```

## Weekly-daily (structural reporting)

Now we keep the *daily* process but model the fact that reports only arrive once a week.
The idea is to build a daily grid where the cumulative count is only observed on the reporting day and carried forward on the other days, then tell the model which cells are real observations.
We start from the daily reporting triangle we fit to, blank out the cumulative counts on every other day, and let `enw_impute_na_observations()` carry the last reported value forward; `enw_flag_observed_observations()` records which cells are real in a `.observed` column.

Which day we choose as the reporting day matters, and not only for realism.
`enw_aggregate_cumulative()` anchors its weekly grid on the report dates in the data, so the weekly-weekly model already has a reporting day baked in.
If we picked a different day here, the two models would see their last report on different dates, and the one with the older final report would be nowcasting further ahead with less information.
Any difference between the two nowcasts would then just be that information gap rather than the modelling choice we want to compare.
So we read the reporting day off the aggregated data rather than hard-coding it, which keeps the comparison like-for-like.

```{r weekly-daily-scaffold}
# Both models must see their last report on the same date, or the comparison
# below is confounded by one model simply having fresher data
report_dow <- weekdays(max(weekly_long$report_date))
report_dow

weekly_rep_data <- enw_complete_dates(weekly_enw_long, timestep = "day")
weekly_rep_data[, day_of_week := weekdays(report_date)]
weekly_rep_data[
  , confirm := fifelse(day_of_week == report_dow, confirm, NA_real_)
]
weekly_rep_data <- weekly_rep_data |>
  enw_flag_observed_observations() |>
  enw_impute_na_observations() |>
  enw_filter_reference_dates_by_report_start() |>
  enw_add_incidence()

weekly_rep_pobs <- weekly_rep_data |>
  enw_complete_dates(timestep = "day") |>
  enw_preprocess_data(max_delay = 16, timestep = "day")
```

We describe the known once-a-week reporting cycle with `enw_dayofweek_structural_reporting()` and pass it to `enw_report()` through its `structural` argument instead of a formula.
Because most daily cells are imputed rather than observed, we also point the observation model at the `.observed` indicator so the likelihood only uses the real reports.

```{r weekly-daily-fit, message = FALSE, results = "hide"}
# Structural variants need a little less adapt_delta and more tree depth
fit_opts_structural <- enw_fit_opts(
  save_warmup = FALSE, pp = TRUE,
  chains = 4, parallel_chains = 4,
  iter_warmup = 500, iter_sampling = 500,
  adapt_delta = 0.9, max_treedepth = 12,
  show_messages = FALSE, refresh = 0
)

structural <- enw_dayofweek_structural_reporting(
  weekly_rep_pobs, day_of_week = report_dow
)

weekly_daily_nowcast <- epinowcast(
  weekly_rep_pobs,
  expectation = enw_expectation(~ rw(day), data = weekly_rep_pobs),
  report = enw_report(structural = structural, data = weekly_rep_pobs),
  obs = enw_obs(
    family = "poisson", observation_indicator = ".observed",
    data = weekly_rep_pobs
  ),
  fit = fit_opts_structural
)
```

::: {.callout-note}
## Take 5 minutes
This model makes three configuration choices that differ from the course model.
For each, say what it does and what would go wrong if you left it out.

1. `report = enw_report(structural = structural, ...)` instead of a report formula.
2. `observation_indicator = ".observed"` in `enw_obs()`.
3. `expectation = enw_expectation(~ rw(day), ...)` rather than a weekly random walk.
:::

::: {.callout-note collapse="true"}
## Solution

1. The structural report module encodes the known once-a-week reporting schedule.
   Without it the model would treat the carried-forward daily values as ordinary reports and try to explain them with a smooth daily delay, misestimating both the delay and how complete the recent dates are.
2. The observation indicator tells the likelihood to use only the cells flagged as real reports.
   Without it the model would treat every imputed cell as a genuine observation, counting each carried-forward value as data and hugely over-weighting the reports.
3. The daily random walk lets the expected counts change from one day to the next, which is what lets the model track a trend developing *within* the most recent week.
   A weekly random walk would let them change only once a week, so the within-week trend that the weekly-weekly model misses would be largely lost again.
   `epinowcast`'s [temporal aggregation guide](https://package.epinowcast.org/articles/temporal-aggregation.html) instead pairs a weekly random walk with a day-of-week effect, which is a good choice when the within-week shape is a repeating weekly pattern rather than a trend.
:::

The nowcast is now on the daily scale even though reports only arrived weekly, so we compare it against the eventually-complete daily data.

```{r weekly-daily-plot}
weekly_daily_latest <- enw_latest_data(
  to_enw_format(weekly_reporting_triangle)
)
plot(weekly_daily_nowcast, latest_obs = weekly_daily_latest) +
  coord_cartesian(xlim = c(weekly_plot_from, weekly_nowcast_date))
```

This structural nowcast is on the daily scale, so we can read it against the eventually-observed daily points directly.
Its payoff over the weekly-weekly model is the finer daily granularity: a within-week nowcast recovered from once-a-week reports, which the weekly-weekly model cannot give.
It will not match a nowcast fitted to truly daily reports, because once-a-week reporting leaves the model less information about within-week timing than daily data would.

Note that we are asking a lot of the daily random walk here: it has one latent value per day but only one report per week to learn from.
That is deliberate, because keeping the daily process is the whole point of this approach, but it is why the daily nowcast is visibly less certain than the daily nowcasts earlier in this session, which were fitted to daily reports.

To compare the two directly we can aggregate the daily structural nowcast up to the same weekly buckets as the weekly-weekly model and plot them against the eventually-observed weekly totals.
A week can only be rebuilt from the daily model where its whole seven days fall inside the recent nowcast window, so the comparison covers the most recent complete weeks.

```{r weekly-compare}
# Bin each daily reference date onto the weekly grid, keeping only full weeks
weekly_daily_samples <- summary(weekly_daily_nowcast, type = "nowcast_samples")
weekly_daily_samples[
  , week := as.IDate(
      ref_date0 + 7L * ceiling(as.integer(reference_date - ref_date0) / 7)
    )
]
full_weeks <- weekly_daily_samples[
  , .(n = uniqueN(reference_date)), by = week
][n == 7, week]

wd_weekly <- weekly_daily_samples[week %in% full_weeks][
  , .(total = sum(sample)), by = .(week, .draw)
][
  , .(median = median(total), lower = quantile(total, 0.05),
      upper = quantile(total, 0.95)), by = week
][, model := "Weekly-daily (aggregated)"]

ww_weekly <- summary(weekly_nowcast, type = "nowcast")[
  reference_date %in% full_weeks
][, .(week = reference_date, median = q50, lower = q5, upper = q95)][
  , model := "Weekly-weekly"
]

obs_weekly <- enw_latest_data(
  enw_aggregate_cumulative(
    to_enw_format(weekly_reporting_triangle), timestep = "week"
  )
)[reference_date %in% full_weeks, .(week = reference_date, observed = confirm)]

ggplot(rbind(wd_weekly, ww_weekly), aes(week, colour = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  geom_line(aes(y = median)) +
  geom_point(aes(y = median)) +
  geom_point(data = obs_weekly, aes(week, observed), inherit.aes = FALSE) +
  labs(x = "Week (reference date)", y = "Weekly onsets",
       colour = NULL, fill = NULL)
```

Because the nowcast sits in a steep decline, the two models disagree about the most recent, still-incomplete week (the black points are the eventually-observed totals).
The weekly-weekly model has only a handful of weekly points to learn from, so its weekly random walk carries the recent level forward and overshoots the falling week.
The weekly-daily model keeps a daily random walk, so it can see the decline happening *within* the week and its central estimate lands closer to the eventual total.
On the earlier week, which is complete and already fully reported, the two agree closely.

Both intervals do contain the eventual total here, so this is a difference in central estimate and in how much of the week's shape each model can represent, not a case of one model being flatly wrong.
The weekly-weekly interval is the more overconfident of the two: the eventual total sits close to its lower edge, so a slightly sharper turn would have taken it outside.

The weekly-weekly model gave us a weekly nowcast from a small, cheap model, while the weekly-daily model recovers a daily nowcast from the same weekly reports at the price of a larger reporting triangle and a harder fit.
The [temporal aggregation guide](https://package.epinowcast.org/articles/temporal-aggregation.html) scores these approaches against a fully daily benchmark if you want to see how much each one gives up.

::: {.callout-warning}
## The daily advantage is conditional
The daily model helps here because the epidemic is changing rapidly right at the nowcast date; when the recent weeks are flat or already reported the two approaches agree, and weekly-weekly is then just as good for far less compute.
The daily model is also harder to fit and its weekly totals carry more uncertainty, so on the weekly total alone it will not always beat weekly-weekly.
What it reliably buys is resolution: a daily nowcast that keeps the within-week shape weekly aggregation discards, which matters most exactly when things are changing fast.

The comparison is also only fair because we gave both models the same final report date.
It is worth being alert to this in your own work: a nowcast that appears to lose to a simpler model is often just working from staler data, and that shows up as a wider interval and a slower response to a turning point rather than as an obvious error.
:::

::: {.callout-note}
## Take 5 minutes
Read the comparison plot (`weekly-compare`), where both weekly nowcasts are shown against the eventually-observed weekly totals (the black points).

1. On the earlier, complete week the two models agree; at the most recent, still-incomplete week they diverge. Which model's central estimate lands closer to the eventual total, and in which direction does the other one err?
2. Why can the weekly-daily model see the decline that the weekly-weekly model misses, given both are built from exactly the same weekly reports?
:::

::: {.callout-note collapse="true"}
## Solution

1. On the complete week both models sit on the eventual total, so they are indistinguishable there.
   At the recent falling week the weekly-daily median is lower and closer to the eventual total, while the weekly-weekly median sits above it: with only a handful of weekly points, its weekly random walk carries the recent level forward and overshoots a week that is falling fast.
   Both intervals still cover the eventual total, but the weekly-weekly one only just does, with the truth near its lower edge.
2. The two models see the same *reports*, but not at the same resolution.
   Within one weekly release, each day of the preceding week has had a different amount of time to be reported, so the same delay distribution implies a different degree of completeness for each day.
   The daily model uses that to reconstruct the shape *within* the week and can see the decline under way; the weekly model collapses the week to a single number and has nothing left to read the within-week trend from.
:::

# Other reporting complications

We have worked through three complications end to end: a day-of-week effect, a time-varying delay, and weekly reporting.
Other complications are handled by the same modular idea, each a change to one formula rather than a new model.
Two worth knowing about are delays that vary by strata (for example age or region), which add the stratifying variable to a module formula such as `enw_reference(~ (1 | age_group), data = pobs)` so each stratum has its own delay while borrowing strength across strata, and missing reference dates, which `epinowcast` models with a dedicated module.
We link the relevant documentation for both in *Going further* below rather than fitting them here.

::: {.callout-note}
## Take 5 minutes
Suppose you also had the data split by two age groups, and you suspected reporting was slower in one of them.
Which module would you change, and what would be the advantage of fitting both age groups in one `epinowcast` model rather than two separate nowcasts?
:::

::: {.callout-note collapse="true"}
## Solution

You would add age group to the *reference* module so the delay distribution can differ by group.
Fitting both groups jointly lets a sparse group borrow strength from the other through shared (partially pooled) structure, which is more stable than fitting each group on its own, especially when counts are low.
:::

# Going further

This session only scratches the surface of what `epinowcast` can do, and `epinowcast` is only one of several tools for this kind of modelling.
In fact `epinowcast` can express essentially all of the modelling in this course, from delay estimation and nowcasting through renewal-based $R_t$ estimation to the reporting effects above.

`epinowcast` has extensive documentation and worked case studies that go well beyond this session; rather than try to cover everything here, the best next step is to read them:

- The [`epinowcast` package website](https://package.epinowcast.org/) [@epinowcastPackage2024], including its [model description](https://package.epinowcast.org/articles/model.html) and the [getting started vignette](https://package.epinowcast.org/articles/epinowcast.html).
- The [model features summary](https://package.epinowcast.org/articles/features.html), a table of everything `epinowcast` currently supports with links to the relevant documentation for each feature.

To go deeper on the topics we covered, the relevant reference pages are:

- **Reporting schedule (weekly or batched reporting):** the [temporal aggregation guide](https://package.epinowcast.org/articles/temporal-aggregation.html), [`enw_aggregate_cumulative()`](https://package.epinowcast.org/reference/enw_aggregate_cumulative.html) and the `timestep` argument of [`enw_preprocess_data()`](https://package.epinowcast.org/reference/enw_preprocess_data.html).
- **Day-of-week and other report-date effects:** the [`enw_report()`](https://package.epinowcast.org/reference/enw_report.html) report-date module.
- **Time-varying delay (reference-date effects):** the [`enw_reference()`](https://package.epinowcast.org/reference/enw_reference.html) reference-date module.
- **Delays that vary by strata:** the [hierarchical age-stratified nowcasting case study](https://package.epinowcast.org/articles/germany-age-stratified-nowcasting.html), which fits many reporting triangles at once with partially pooled effects.
- **Missing reference dates:** the [`enw_missing()`](https://package.epinowcast.org/reference/enw_missing.html) module.
- The [plotting reference](https://package.epinowcast.org/reference/plot.epinowcast.html) for more ways to visualise a fit than the default `plot()` we used here.

Its [case studies](https://package.epinowcast.org/articles/package-use-cases.html) and vignettes also go beyond this session.
The [real-time $R_t$ estimation vignette](https://package.epinowcast.org/articles/single-timeseries-rt-estimation.html) jointly estimates transmission and reporting delays with a renewal-based expectation model, generalising the joint model from the [joint nowcasting session](joint-nowcasting).
Others work through evaluation of nowcasts with proper scoring rules, as we did for forecasts in the [combining nowcasting and forecasting session](forecasting-nowcasting).

::: {.callout-note}
## Challenge: fit the wrong model on purpose
A good way to build intuition for these models is to misspecify one deliberately and see how the misspecification shows up.
Reusing the models and data already defined above, fit the wrong model to each worked example and compare it to the correct one:

- Fit the constant-delay course model (the reference module with no time variation) to the *time-varying-delay* data, and compare its nowcast to the `~ rw(week)` model that lets the delay drift.
- Fit a model that omits the day-of-week terms to the *day-of-week* data, and compare it to the model that includes them.
- Fit the model to the weekly data as if it were ordinary daily data, and compare it to the weekly-aware fits.

For each, compare the nowcasts against the eventually-observed data and use a posterior-predictive check (`type = "posterior_prediction"`) to see where the fit breaks down.
Which misspecifications bite hardest, and how would you detect them from the fit alone if you did not have the eventual data to check against?
:::

`epinowcast` is not the only option, and the right tool depends on the problem.

- The bespoke Stan models we built in this course remain a perfectly good choice when you understand the reporting process and want full control.
- The [`baselinenowcast`](https://baselinenowcast.epinowcast.org) package implements simple multiplicative nowcasting methods that make useful benchmarks.
- A range of other methods and frameworks are used in practice; the collaborative nowcasting comparison in [@wolfframCollaborativeNowcastingCOVID192023] surveys several and discusses what drives differences in their performance.

For the methodological background to generative nowcasting models of this kind, see [@lisonGenerativeBayesianModeling2024].

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
