---
title: "Modelling complex reporting processes"
order: 14.5
bibliography: ../nfidd.bib
---

::: {.callout-important}
## Draft session

This is a **draft** session and is still under review.
The simulation and data-preparation code runs as part of the course, but the `epinowcast` model fits do not.
Every chunk that compiles or fits an `epinowcast` model is marked `#| eval: false`, so the code is shown but not executed when the site is built.
This keeps the site build fast and means `epinowcast` is *not* a dependency of the course.
To run those chunks yourself you will need a working `cmdstanr` toolchain and the `epinowcast` package installed (see the setup callout below).
:::

# Introduction

In the [nowcasting concepts session](nowcasting) and the [joint nowcasting session](joint-nowcasting) we built our own nowcasting models in Stan.
We started from a known delay distribution, then learned to jointly estimate the delay and the nowcast from the reporting triangle, and finally added a renewal process so that the generative model of transmission informed the nowcast.

Those bespoke models were ideal for *learning* how nowcasting works because we could see every part of the model.
They are also a perfectly valid choice for real analyses: there is nothing wrong with writing your own model when you understand the reporting process well.
However, real surveillance data is usually messier than those models handle comfortably.
In this session we step back and ask what happens when the reporting process itself becomes complicated.
We then introduce [`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] as one flexible, well-tested framework for exactly these situations, and frame it as a direct generalisation of the joint nowcasting model we already built.

## Slides

- There are no slides for this draft session yet.

## Objectives

This session aims to show why real reporting processes are harder to model than our bespoke Stan models assume, and to introduce `epinowcast` as a framework that generalises the joint nowcasting model we have already built.

::: {.callout-note collapse="true"}

# Setup

## Source file

The source file of this session is located at `sessions/complex-reporting-processes.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to load infection times and the simulation helpers we have used throughout, the `dplyr` and `tidyr` packages for data wrangling, and the `ggplot2` library for plotting.
The worked `epinowcast` example additionally uses the `epinowcast` package and the `data.table` package that it uses for its data structures.

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
```

::: {.callout-tip}
`epinowcast` is not a dependency of this course, because it compiles its models with `cmdstanr` and would slow the course build down considerably.
The `epinowcast` chunks below are shown but not run.
To run them yourself, install `epinowcast` from its R-universe (it is also on CRAN) along with `cmdstanr`:

```{r install-epinowcast}
#| eval: false
install.packages(
  "epinowcast",
  repos = c("https://epinowcast.r-universe.dev", getOption("repos"))
)
```

You will also need a working CmdStan toolchain (the same one used elsewhere in this course).
Installation and setup instructions are on the [package website](https://package.epinowcast.org/).
:::

## Initialisation

We set a random seed for reproducibility.

```{r initialise}
set.seed(123)
```

:::

# Why real reporting is harder than our models assume

## The model we have built so far

In the [joint nowcasting session](joint-nowcasting) we modelled each cell of the reporting triangle as

$$
  n_{t,d} \mid \lambda_{t}, p_{t,d} \sim \text{Poisson}\left(\lambda_{t} \times p_{t,d}\right),
$$

where $\lambda_{t}$ is the expected number of onsets on day $t$ and $p_{t,d}$ is the probability that an onset on day $t$ is reported with a delay of $d$ days.
We modelled $\lambda_{t}$ with a geometric random walk (or a renewal process) and $p_{t,d}$ with a single delay distribution that was the same on every day.

This works well in our simulated examples because we simulated the data from almost exactly that model.
Real data rarely cooperate so neatly.

## Ways real reporting departs from this model

Consider some features of routine surveillance data that our bespoke models do not capture well.

- **Day-of-week reporting effects.**
Many reporting systems report far less at weekends and on public holidays, then catch up early in the following week.
This is a property of the *report date*, not of the underlying epidemic, and it breaks the assumption that the delay distribution depends only on the delay $d$.

- **Delays that vary over time.**
Reporting delays lengthen when a system is under strain and shorten when it has spare capacity.
A single fixed $p_{t,d}$ cannot represent a delay distribution that drifts across the outbreak.

- **Delays that vary by strata.**
Reporting speed often differs by age group, region, or pathogen, so a single shared delay can be wrong for some strata.

- **Weekly or otherwise batched reporting.**
Some systems report only once a week, or in irregular batches, so that counts arrive in lumps rather than smoothly.

- **Missing and late reporting.**
Some reports never arrive, or arrive far beyond the maximum delay we assumed, and weekends and holidays create gaps in the data.

::: {.callout-note}
## Take 5 minutes
Think about the joint nowcasting model you built.
What would you have to change in that Stan code to add a day-of-week reporting effect *and* let the delay distribution change over the course of the outbreak?
:::

::: {.callout-note collapse="true"}
## Solution

There is no single right answer, but the changes add up quickly.

- You would need to index the reporting probabilities by report date (or day of week) as well as by delay, turning the delay vector into a more complex structure.
- You would need a model for how the delay distribution changes over time, for example a random walk on the parameters of the delay distribution, with its own priors.
- You would need to keep the reporting probabilities on each day summing to one, which is awkward once they vary by date.
- Each addition makes the Stan model longer, slower to fit, and harder to debug and test.

None of this is impossible, and writing it yourself remains a reasonable option.
But it is the point at which reaching for a general, tested framework that already implements these features lets us focus on the modelling choices rather than the implementation.
:::

# Introducing `epinowcast`

[`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] is a Bayesian framework for real-time infectious disease surveillance.
It is *one* of several tools for this kind of modelling (we point to others at the end of the session), and it is a direct generalisation of the joint nowcasting model we built: it works from the reporting triangle and jointly estimates the expected counts and the reporting process, but it makes each component far more flexible and exposes them through a formula interface rather than hand-written Stan code.

`epinowcast` decomposes the problem into modular pieces, each of which mirrors something we have already seen.

- An **expectation** model for the expected counts $\lambda_{t}$ (the analogue of our geometric random walk or renewal process).
- A **reference date** model for the delay distribution by date of the epidemiological event (the analogue of our $p_{t,d}$, but allowed to vary over time and by covariates).
- A **report date** model for effects that act on the *report* date, such as day-of-week reporting effects.

Each piece is specified with an R formula, so adding a day-of-week effect or a time-varying delay is a one-line change rather than a rewrite of the model.
Under the hood `epinowcast` still builds and fits a Stan model, but the model is tested, optimised, and reused across problems.

::: {.callout-tip}
We are treating `epinowcast` as a tool here rather than dissecting its internals.
The mapping back to the bespoke model is deliberate: if you understand the joint nowcasting session, you already understand what `epinowcast` is doing, just with more flexible components.
:::

# Simulating data with a complex reporting process

Rather than reach for a real dataset, we simulate data the same way we have throughout the course, so that we know the truth we are trying to recover.
We build directly on the reporting-triangle simulation from the [joint nowcasting session](joint-nowcasting), then add a realistic reporting complication on top: a **day-of-week reporting effect**.

First we generate the symptom-onset time series exactly as before, from the shared infection process.

```{r simulate-onsets}
gen_time_pmf <- make_gen_time_pmf()
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), ip_pmf
)
cutoff <- 71
head(onset_df)
```

Next we define the reporting delay distribution, as in the joint nowcasting session.

```{r reporting-delay}
reporting_delay_pmf <- censored_delay_pmf(
  rlnorm, max = 15, meanlog = 1, sdlog = 0.5
)
```

Now we add the complication.
Real systems often report less at weekends, so the number of reports landing on a given day depends on the **day of the week of the report date**, not on the delay alone.
We capture this with a multiplicative weekly weight, assuming day 0 of our simulation is a Monday.

```{r dow-effect}
# Multiplicative reporting weight for Mon, Tue, ..., Sun
# Reporting is suppressed at the weekend (Sat/Sun)
dow_effect <- c(1, 1, 1, 1, 0.9, 0.4, 0.5)
```

We then build the reporting triangle just as in the joint nowcasting session, but multiply the reporting probability by the day-of-week weight for each report day before adding Poisson noise.

```{r simulate-triangle}
reporting_triangle <- onset_df |>
  filter(day < cutoff) |>
  mutate(
    reporting_delay = list(
      tibble(d = 0:15, reporting_delay = reporting_delay_pmf)
    )
  ) |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    # day of week of the *report* day (0 = Monday)
    report_dow = reported_day %% 7,
    dow_weight = dow_effect[report_dow + 1],
    reported_onsets = rpois(n(), onsets * reporting_delay * dow_weight)
  )
tail(reporting_triangle)
```

As we only observe reports up to the present, we keep just the cells that have been reported by our cutoff.

```{r filter-triangle}
filtered_triangle <- reporting_triangle |>
  filter(reported_day <= max(day))
```

We can see the day-of-week effect clearly if we sum the reports by report day: the totals dip on a regular weekly cycle.

```{r plot-dow}
reports_by_day <- filtered_triangle |>
  summarise(reports = sum(reported_onsets), .by = reported_day)

ggplot(reports_by_day, aes(x = reported_day, y = reports)) +
  geom_col() +
  labs(x = "Report day", y = "Reports", title = "Reports by report day")
```

::: {.callout-note}
## Take 2 minutes
Look at the plot of reports by report day.
Where does the weekly dip come from, and why would a model with a delay distribution that depends only on the delay $d$ struggle to reproduce it?
:::

::: {.callout-note collapse="true"}
## Solution

The dip is a property of the *report date*: fewer reports are processed at weekends, regardless of when the onset occurred.
Our bespoke $p_{t,d}$ only depends on the delay $d$ since onset, so it has no way to know which calendar day a report lands on.
To reproduce the weekly pattern the model needs a term that depends on the report date itself, which is exactly the report-date module in `epinowcast`.
:::

## Putting the data into `epinowcast` format

`epinowcast` expects data in a "long" format with one row per (reference date, report date) pair, where `confirm` is the **cumulative** count for that reference date as known on that report date.
This is the same reporting triangle we have been using, just with calendar dates instead of integer days and cumulative rather than incremental counts.

We attach arbitrary calendar dates (so that weekends line up with our day-of-week effect) and form the cumulative counts.

```{r to-enw-format}
# Day 0 of the simulation is a Monday
ref_date0 <- as.Date("2023-01-02")

enw_long <- filtered_triangle |>
  transmute(
    reference_date = ref_date0 + day,
    report_date = ref_date0 + reported_day,
    count = reported_onsets
  ) |>
  arrange(reference_date, report_date) |>
  group_by(reference_date) |>
  mutate(confirm = cumsum(count)) |>
  ungroup() |>
  select(reference_date, report_date, confirm)

head(enw_long)
```

::: {.callout-note}
## Take 2 minutes
How does this data structure relate to the reporting triangle from the [joint nowcasting session](joint-nowcasting)?
:::

::: {.callout-note collapse="true"}
## Solution

It is the same idea in a different layout.
Each `reference_date` is a row of the triangle, each `report_date` gives a column, and `confirm` is the cumulative version of the cell counts $n_{t,d}$ we summed over delays before.
`epinowcast` works with the cumulative counts and converts them into the reporting triangle internally during preprocessing.
:::

# A worked example: nowcasting with a day-of-week effect

We now hand this simulated data to `epinowcast`.
The remaining chunks in this section use `epinowcast` and so are not run when the site is built; install the package (see the setup callout) to run them yourself.

`epinowcast` preprocesses the data into its internal reporting-triangle representation, where we also choose the maximum delay we will model.

```{r preprocess}
#| eval: false
library("epinowcast")
library("data.table")

pobs <- enw_preprocess_data(as.data.table(enw_long), max_delay = 15)
pobs
```

Now we specify the model as three modules, each mapping back to a choice we made in the joint nowcasting session.

```{r modules}
#| eval: false
# Expected counts: a weekly random walk (cf. our geometric random walk)
expectation_module <- enw_expectation(~ rw(week), data = pobs)

# Delay distribution by reference date: a single parametric delay for now
reference_module <- enw_reference(~ 1, data = pobs)

# Report-date effects: a day-of-week reporting effect
report_module <- enw_report(~ (1 | day_of_week), data = pobs)
```

::: {.callout-note}
## Take 3 minutes
Compare these three formulas to the choices we made in the [joint nowcasting session](joint-nowcasting).
Which module corresponds to $\lambda_{t}$ and which to $p_{t,d}$, and what is genuinely new here?
:::

::: {.callout-note collapse="true"}
## Solution

- `enw_expectation()` plays the role of $\lambda_{t}$: `rw(week)` is a random walk on the expected counts, much like the geometric random walk we used for onsets.
- `enw_reference()` plays the role of the delay distribution $p_{t,d}$.
Here `~ 1` keeps it constant over time, exactly as in our bespoke model, but we could let it vary by writing, for example, `rw(week)` here too.
- The genuinely new piece is `enw_report(~ (1 | day_of_week))`: a reporting effect that depends on the *report* date rather than the delay, which is awkward to add to the hand-rolled model but is a one-line change here.
This is the term that absorbs the weekly dip we built into the simulation.
:::

We then fit the model.
This is the step that compiles and samples a Stan model, so it is slow; it is marked `eval: false` here.

```{r fit}
#| eval: false
nowcast <- epinowcast(
  pobs,
  expectation = expectation_module,
  reference = reference_module,
  report = report_module,
  fit = enw_fit_opts(
    save_warmup = FALSE, pp = TRUE,
    chains = 2, parallel_chains = 2,
    iter_warmup = 500, iter_sampling = 500
  )
)
```

::: {.callout-tip}
The `enw_fit_opts()` call controls sampling in the same way the arguments to `nfidd_sample()` did in earlier sessions (number of chains, warmup and sampling iterations, and so on).
We use a small number of iterations here purely to keep the example quick; a real analysis would use more.
:::

Once fitted, the `epinowcast` object has plotting and summary methods.
Plotting shows the nowcast for the recent, incomplete reference dates alongside the counts available at the nowcast date.

```{r plot-nowcast}
#| eval: false
plot(nowcast)
```

As in our bespoke nowcasts, the most recent dates are the most uncertain because they are the least complete.
The difference here is that the day-of-week reporting effect is now absorbed by the model rather than distorting the nowcast.

# Other reporting complications

We have worked through a day-of-week effect end to end.
The other complications from the start of the session are handled by the same modular idea: each is a change to one formula, not a new model.
Rather than code every one, here is how you would approach them and a couple of short exercises.

- **Time-varying delays.**
Put structure on the *reference* module so the delay distribution can drift over the outbreak, for example `enw_reference(~ rw(week), data = pobs)`.
- **Delays that vary by strata.**
Add the stratifying variable to the relevant module formula, for example a fixed or random effect of age group on the reference module, so each stratum can have its own delay while borrowing strength across strata.
- **Weekly or batched reporting.**
This is another report-date pattern: counts arrive in lumps on particular report days, which a report-date model can represent.
- **Missing and late reporting.**
`epinowcast` has dedicated support for missing reference dates and can model the proportion of reports that are eventually observed, rather than assuming every report arrives within the maximum delay.

::: {.callout-note}
## Take 5 minutes
We modelled the delay distribution as constant over time with `enw_reference(~ 1, ...)`.
Sketch how you would let the delay distribution change over the outbreak, and say why that might matter.
What would you change in our *simulation* above to generate data with a time-varying delay to test it on?
:::

::: {.callout-note collapse="true"}
## Solution

You can put structure on the reference module just as on the expectation module, for example `enw_reference(~ rw(week), data = pobs)` to let the delay distribution drift week to week.
This matters when reporting speed changes over time, for instance when a system is overwhelmed during a wave and delays lengthen, or improves as processes are streamlined.
A fixed delay distribution would then mis-estimate how much of the recent data is still missing, biasing the nowcast.

To simulate it, you would let the parameters of `reporting_delay_pmf` depend on the onset `day` (for example, slowly increasing `meanlog`) instead of using one fixed PMF for every day, then rebuild the triangle as before.
:::

::: {.callout-note}
## Take 5 minutes
Suppose you also had the data split by two age groups, and you suspected reporting was slower in one of them.
Which module would you change, and what would be the advantage of fitting both age groups in one `epinowcast` model rather than two separate nowcasts?
:::

::: {.callout-note collapse="true"}
## Solution

You would add age group to the *reference* module so the delay distribution can differ by group.
Fitting both groups jointly lets a sparse group borrow strength from the other through shared (partially pooled) structure, which is more stable than fitting each group on its own, especially when counts are low.
:::

# Going further

This session only scratches the surface of what `epinowcast` can do, and `epinowcast` is only one of several tools for this kind of modelling.

`epinowcast` has extensive documentation and worked case studies that go well beyond this session; rather than try to cover everything here, the best next step is to read them:

- The [`epinowcast` package website](https://package.epinowcast.org/) [@epinowcastPackage2024], including its [model description](https://package.epinowcast.org/articles/model.html) and the [getting started vignette](https://package.epinowcast.org/articles/epinowcast.html).
- Its case studies and vignettes work through stratified nowcasting (many reporting triangles at once with partially pooled effects), richer delay and reporting models, joint nowcasting with $R_t$ estimation (a renewal-based expectation model, generalising the joint model from the [joint nowcasting session](joint-nowcasting)), and evaluation of nowcasts with proper scoring rules (as we did for forecasts in the [combining nowcasting and forecasting session](forecasting-nowcasting)).

`epinowcast` is not the only option, and the right tool depends on the problem.

- The bespoke Stan models we built in this course remain a perfectly good choice when you understand the reporting process and want full control.
- The [`baselinenowcast`](https://baselinenowcast.epinowcast.org) package implements simple multiplicative nowcasting methods that make useful benchmarks.
- A range of other methods and frameworks are used in practice; the collaborative nowcasting comparison in [@wolfframCollaborativeNowcastingCOVID192023] surveys several and discusses what drives differences in their performance.

For the methodological background to generative nowcasting models of this kind, see [@lisonGenerativeBayesianModeling2024].

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
