---
title: "Modelling complex reporting processes"
order: 14.5
bibliography: ../nfidd.bib
---

# Introduction

In the [joint nowcasting session](joint-nowcasting) we built a nowcasting model in Stan that jointly estimated the delay distribution and the nowcast from the reporting triangle, and then added a renewal process so transmission informed the nowcast.
That model assumed a single delay distribution, the same on every day.

Bespoke models like that are ideal for *learning* how nowcasting works, and they remain a reasonable choice for real analyses when you understand the reporting process well.
But routine surveillance data is usually messier: the reporting process itself can be complicated in ways a single fixed delay cannot capture.
In this session we introduce [`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024], one flexible framework for these situations.
We reproduce the joint nowcasting model in `epinowcast`, then *extend* it to handle reporting complications the bespoke model cannot: a day-of-week reporting effect, a delay that drifts over time, and weekly (batched) reporting.
Throughout we simulate data with the feature, fit a model that handles it, and look at the result.

## Slides

- [Modelling complex reporting processes](slides/complex-reporting-processes)

## Objectives

This session shows why real reporting processes are harder to model than a single fixed delay assumes, reproduces the joint nowcasting model in `epinowcast`, and extends it to complex reporting features through worked examples.

::: {.callout-note collapse="true"}

# Setup

## Source file

The source file of this session is located at `sessions/complex-reporting-processes.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to load infection times and the simulation helpers we have used throughout, the `dplyr` and `tidyr` packages for data wrangling, and the `ggplot2` library for plotting.
The worked examples additionally use the `epinowcast` package and the `data.table` package that it uses for its data structures.

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
library("epinowcast")
library("data.table")
```

::: {.callout-tip}
This session uses the development version of `epinowcast` from GitHub, as some of the data and preprocessing helpers used here are newer than the latest CRAN release.
You can install it together with `cmdstanr` with:

```{r install, eval = FALSE}
install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
)
remotes::install_github("epinowcast/epinowcast")
```

`epinowcast` compiles its models with `cmdstanr` the first time a model is fitted, which needs a working CmdStan toolchain (the same one used elsewhere in this course).
Full installation and setup instructions, including how to install CmdStan, are on the [package website](https://package.epinowcast.org/).
The fits in this session use the course-standard sampler settings but short, simulated data so the course builds in reasonable time.
:::

## Initialisation

We set a random seed for reproducibility.
We also tell `epinowcast` not to warn about partially specified initial conditions, which keeps the output tidy.

```{r initialise}
set.seed(123)
options(cmdstanr_warn_inits = FALSE)
```

:::

# Why real reporting is harder than our models assume

Recall the joint nowcasting model: we modelled each cell of the reporting triangle as

$$
  n_{t,d} \mid \lambda_{t}, p_{t,d} \sim \text{Poisson}\left(\lambda_{t} \times p_{t,d}\right),
$$

where $\lambda_{t}$ is the expected number of onsets on day $t$ and $p_{t,d}$ is the probability that an onset on day $t$ is reported with a delay of $d$ days.
We gave $\lambda_{t}$ a daily geometric random walk (or a renewal process) and $p_{t,d}$ a single delay distribution, the same on every day.

That worked in our simulated examples because we simulated from almost exactly that model.
Real data rarely cooperate so neatly.
Routine surveillance data often shows features a single fixed delay does not capture: day-of-week reporting effects, reporting delays that vary over time or by strata such as age or region, weekly or batched reporting, and missing or late reports.
Rather than list these and tell you they are hard, we *show* several: simulate the feature, fit a model that handles it, and look at the result.

::: {.callout-note}
## Take 5 minutes
Think about the joint nowcasting model you built.
What would you have to change in that Stan code to add a day-of-week reporting effect *and* let the delay distribution change over the course of the outbreak?
:::

::: {.callout-note collapse="true"}
## Solution

There is no single right answer, but the concrete changes add up quickly.

- **A day-of-week reporting effect.** The reporting probability would have to depend on the calendar day the report lands on, not just the delay $d$. In practice you would turn the delay vector into a matrix indexed by reference date *and* delay, $p_{t,d}$, and multiply each cell by a day-of-week factor $w_{(t+d)\,\bmod\,7}$ keyed on the *report* date $t+d$. That means seven new parameters (a `simplex` or sum-to-zero vector for identifiability) and an extra loop over cells in `transformed parameters`.
- **A time-varying delay.** You would put a random walk on the parameters of the delay distribution, for example `meanlog[t] = meanlog[t-1] + sigma * eta[t]` (and similarly for `sdlog`), each with its own prior, then recompute the delay PMF for every reference date $t$ rather than once. That is a new vector of innovations `eta`, a new standard-deviation parameter `sigma` with a prior, and a per-date PMF calculation.
- **Keeping it a valid distribution.** The probabilities for each reference date must still sum to one over delays after both changes, so you would re-impose the per-date normalisation (a simplex or an explicit renormalisation) every day instead of reusing one shared, pre-normalised vector.
- **The cost.** Each addition means more parameters, a larger `transformed parameters` block, slower gradient evaluations, and more code to unit-test, all before you have fitted anything.

None of this is impossible, and writing it yourself remains a reasonable option.
But it is the point at which reaching for a general framework that already implements these features lets us focus on the modelling choices rather than the implementation.
:::

# Introducing `epinowcast`

[`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] is a Bayesian framework for real-time infectious disease surveillance.
It is *one* of several tools for this kind of modelling (we point to others at the end of the session), and it is a direct generalisation of the joint nowcasting model we built: it works from the reporting triangle and jointly estimates the expected counts and the reporting process, but it makes each component far more flexible and exposes them through a formula interface rather than hand-written Stan code.

`epinowcast` decomposes the problem into modular pieces, each of which mirrors something we have already seen.

- An **expectation** model for the expected counts $\lambda_{t}$ (the analogue of our geometric random walk or renewal process).
- A **reference date** model for the delay distribution by date of the epidemiological event (the analogue of our $p_{t,d}$, but allowed to vary over time and by covariates).
- A **report date** model for effects that act on the *report* date, such as day-of-week reporting effects.
- An **observation** model for the likelihood linking expected to observed counts (the analogue of our Poisson likelihood for $n_{t,d}$).

Each piece is specified with an R formula, so adding a day-of-week effect or a time-varying delay is a one-line change rather than a rewrite of the model.
Internally, `epinowcast` still builds and fits a Stan model.

::: {.callout-tip}
We are treating `epinowcast` as a tool here rather than dissecting its internals.
The mapping back to the bespoke model is deliberate: if you understand the joint nowcasting session, you already understand what `epinowcast` is doing, just with more flexible components.
:::

# The course model, in `epinowcast`

Before adding anything new, let us reproduce the joint nowcasting model we already built, but in `epinowcast`.
We simulate exactly as in the [joint nowcasting session](joint-nowcasting): the same onset time series, the same fixed reporting delay, and the same reporting triangle.

```{r simulate-onsets}
gen_time_pmf <- make_gen_time_pmf()
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), ip_pmf
)
cutoff <- 71

reporting_delay_pmf <- censored_delay_pmf(
  rlnorm, max = 15, meanlog = 1, sdlog = 0.5
)

reporting_triangle <- onset_df |>
  filter(day < cutoff) |>
  mutate(
    reporting_delay = list(
      tibble(d = 0:15, reporting_delay = reporting_delay_pmf)
    )
  ) |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    reported_onsets = rpois(n(), onsets * reporting_delay)
  )
```

## Putting the data into `epinowcast` format

`epinowcast` expects data in a "long" format with one row per (reference date, report date) pair, where `confirm` is the **cumulative** count for that reference date as known on that report date.
This is the same reporting triangle we have been using, just with calendar dates instead of integer days and cumulative rather than incremental counts.

We attach arbitrary calendar dates (so that, later, weekends line up with a day-of-week effect) and let `epinowcast`'s own `enw_add_cumulative()` form the cumulative counts from the incident `new_confirm` column, rather than computing the cumulative sum by hand.
Because we reuse this transformation in each example below, we wrap it in a small helper.

Our simulation already generates every report an onset will ever receive (all delays $d = 0, \ldots, 15$), so `to_enw_format()` formats the *complete* eventually-reported data.
From that one complete data set we derive the two things we need with `epinowcast`'s own functions rather than truncating by hand:

- the data available **as of the nowcast**, which we fit to, using `enw_filter_report_dates()` to keep only reports that had arrived by the nowcast date;
- the **eventually-reported** data, which we judge the nowcast against, using `enw_latest_data()` to take the most recent report for each reference date.

```{r to-enw-helper}
# Day 0 of the simulation is a Monday
ref_date0 <- as.Date("2023-01-02")
# The nowcast is made once reports up to this date are in
nowcast_date <- ref_date0 + (cutoff - 1)

# Format the complete (eventually-reported) data in epinowcast's long format
to_enw_format <- function(triangle) {
  triangle |>
    transmute(
      reference_date = ref_date0 + day,
      report_date = ref_date0 + reported_day,
      new_confirm = reported_onsets
    ) |>
    arrange(reference_date, report_date) |>
    as.data.table() |>
    enw_add_cumulative()
}

# The as-of-nowcast snapshot we fit to: drop reports that arrived later
enw_long <- enw_filter_report_dates(
  to_enw_format(reporting_triangle), latest_date = nowcast_date
)
head(enw_long)
```

::: {.callout-tip}
`enw_filter_report_dates()` plus `enw_latest_data()` makes the two data sets explicit, which is why we use them here.
`enw_retrospective()` is a one-shot helper that produces an as-of-nowcast snapshot and its eventual-data comparison together, and is worth knowing about for real analyses.
:::

::: {.callout-note}
## Take 2 minutes
How does this data structure relate to the reporting triangle from the [joint nowcasting session](joint-nowcasting)?
:::

::: {.callout-note collapse="true"}
## Solution

It is the same idea in a different layout.
Each `reference_date` is a row of the triangle, each `report_date` gives a column, `new_confirm` is the cell count $n_{t,d}$, and `confirm` is its cumulative version over delays.
`epinowcast` works with the cumulative `confirm` and converts it back into the reporting triangle internally during preprocessing.
:::

## Preprocessing and fitting

`epinowcast` preprocesses the data into its internal reporting-triangle representation, where we also choose the maximum delay we will model.

```{r preprocess}
pobs <- enw_preprocess_data(enw_long, max_delay = 15)
```

The preprocessed object has its own `plot()` method.
By default it shows the latest observed notifications by reference date (`type = "obs"`), which is the incomplete picture the nowcast will correct.

```{r preprocess-plot}
plot(pobs)
```

The same method also exposes `epinowcast`'s delay diagnostics through the `type` argument, which are worth inspecting before fitting.
`type = "delay_cumulative"` shows the empirical cumulative proportion reported by delay, so we can see how quickly reports accumulate and check that our `max_delay` is long enough to capture most of the distribution.

```{r preprocess-delay-cumulative}
plot(pobs, type = "delay_cumulative")
```

`type = "delay_fraction"` shows the same information as a heatmap of the fraction reported by delay and reference date, which makes any drift in the delay over time easier to spot.

```{r preprocess-delay-fraction}
plot(pobs, type = "delay_fraction")
```

Now we specify the model.
We define each module explicitly so the mapping back to the bespoke model is visible: a daily random walk for the expected counts (matching the daily geometric random walk we built), a single constant delay distribution, and a Poisson observation model.

```{r course-modules}
# Expected counts lambda_t: a daily random walk (cf. our geometric random walk)
expectation_module <- enw_expectation(~ rw(day), data = pobs)

# Delay distribution p_{t,d} by reference date: constant over time (cf. fixed p)
reference_module <- enw_reference(~ 1, data = pobs)

# Observation model: Poisson, as in n_{t,d} ~ Poisson(lambda_t * p_{t,d})
obs_module <- enw_obs(family = "poisson", data = pobs)
```

We collect the fitting options once and reuse them across the session.
These play the same role as the arguments to `nfidd_sample()` did in earlier sessions, and we set them to match: four chains run in parallel with 500 warmup and 500 sampling iterations each.

```{r fit-opts}
fit_opts <- enw_fit_opts(
  save_warmup = FALSE, pp = TRUE,
  chains = 4, parallel_chains = 4,
  iter_warmup = 500, iter_sampling = 500,
  show_messages = FALSE, refresh = 0
)
```

We then fit the model, passing each module explicitly.
The first fit compiles the Stan model, so it is the slowest step; subsequent fits in this session reuse the compiled model.

```{r course-fit, message = FALSE, results = "hide"}
course_nowcast <- epinowcast(
  pobs,
  expectation = expectation_module,
  reference = reference_module,
  obs = obs_module,
  fit = fit_opts
)
```

::: {.callout-note}
## Take 3 minutes
Compare these modules to the choices we made in the [joint nowcasting session](joint-nowcasting).
Which module corresponds to $\lambda_{t}$, which to $p_{t,d}$, and which to the likelihood?
:::

::: {.callout-note collapse="true"}
## Solution

- `enw_expectation()` plays the role of $\lambda_{t}$: `rw(day)` is a daily random walk on the expected counts, like the geometric random walk we used for onsets.
- `enw_reference()` plays the role of the delay distribution $p_{t,d}$: `~ 1` keeps it constant over time, exactly as in our bespoke model.
- `enw_obs(family = "poisson")` plays the role of the likelihood $n_{t,d} \sim \text{Poisson}(\lambda_{t} p_{t,d})$.

This is the *same* joint nowcasting model we built by hand, now expressed as a handful of modules.
Everything that follows is an extension of one of these modules, or the addition of a report-date module for report-date effects.
:::

## Plotting the nowcast

`epinowcast` objects come with `plot()` and `summary()` methods, so we do not have to extract draws by hand as we did with our bespoke fits.
To judge the nowcast we want to compare it against what was *eventually* reported, so we take the latest report for each reference date from the complete data with `enw_latest_data()` and pass it to `plot()` via `latest_obs`.
Building the comparison data from the complete data matters: if we built it from the truncated snapshot we fit to (`enw_long`), the "eventually observed" points would just reproduce the incomplete nowcast-era counts and the plot would not tell us whether the nowcast was any good.

```{r course-plot}
latest_obs <- enw_latest_data(to_enw_format(reporting_triangle))
plot(course_nowcast, latest_obs = latest_obs)
```

As in our bespoke nowcasts, the most recent dates are the most uncertain because they are the least complete: the nowcast (median and credible intervals) fills in the gap up to the eventually-observed counts (the points), which is exactly the right truncation the nowcast corrects for.
This default plot is the nowcast by *reference* date.
`epinowcast` can also plot the model's posterior predictions against the data that were reported at the time with `type = "posterior_prediction"`, which we facet by reference date to check the fit within each date's accumulating reports.

```{r course-plot-pp}
plot(course_nowcast, type = "posterior_prediction") +
  facet_wrap(vars(reference_date), scales = "free")
```

The plotting methods are more flexible still; see the [`epinowcast` plotting documentation](https://package.epinowcast.org/reference/plot.epinowcast.html) for further options.

# Extending the model: day-of-week effects

Now we add a complication the bespoke model could not handle comfortably.
A day-of-week pattern can enter the data in *three* distinct places, and it helps to keep them separate because each belongs in a different module.

- A **reference-date effect on the expected counts**: events genuinely happen at different rates by day of week. For onset data, for example, fewer onsets may be recorded at weekends. This belongs in `enw_expectation()`.
- A **reference-date effect on the delay**: the reporting *delay distribution* itself differs by the day of week an event occurred, for example weekend onsets take longer to be reported than weekday onsets. This belongs in `enw_reference()`.
- A **report-date effect**: the reporting system processes reports at different rates by day of week, so fewer reports land at weekends regardless of when the onset occurred. This belongs in `enw_report()`.

We build a simulation with all three, one step at a time, so it is clear how each effect enters the data.
Assume day 0 of the simulation is a Monday, so day of week is `day %% 7` (0 = Monday).

We start from the underlying onsets and apply a **reference-date** day-of-week effect on the expected counts, thinning the expected onsets by a weekly weight on the onset day.

```{r dow-reference}
# Multiplicative weekly weights for Mon, Tue, ..., Sun
ref_dow_effect <- c(1, 1, 1, 1, 0.95, 0.7, 0.7) # fewer onsets at weekends
rep_dow_effect <- c(1, 1, 1, 1, 0.9, 0.4, 0.5)  # fewer reports at weekends

dow_onsets <- onset_df |>
  filter(day < cutoff) |>
  mutate(
    ref_dow = day %% 7,
    onsets = onsets * ref_dow_effect[ref_dow + 1]
  )
```

Next we apply the **reporting delay**, but now we let the delay depend on the onset day of week: weekend onsets are reported with a longer delay (a larger `meanlog`) than weekday onsets.
This is a reference-date effect on the *delay* rather than on the counts.

```{r dow-delay}
# Longer delay for weekend onsets (Sat, Sun): a reference-date delay effect
delay_meanlog_dow <- c(1, 1, 1, 1, 1, 1.4, 1.4)

dow_delayed <- dow_onsets |>
  rowwise() |>
  mutate(
    reporting_delay = list(
      tibble(
        d = 0:15,
        reporting_delay = censored_delay_pmf(
          rlnorm, max = 15,
          meanlog = delay_meanlog_dow[ref_dow + 1], sdlog = 0.5
        )
      )
    )
  ) |>
  ungroup() |>
  unnest(reporting_delay) |>
  mutate(reported_day = day + d)
```

Finally we apply a **report-date** day-of-week effect, thinning by a weekly weight on the *report* day, and draw the observed counts.

```{r dow-report}
dow_triangle <- dow_delayed |>
  mutate(
    rep_dow = reported_day %% 7,
    rep_weight = rep_dow_effect[rep_dow + 1],
    reported_onsets = rpois(n(), onsets * reporting_delay * rep_weight)
  )
```

We can see the report-date effect clearly if we sum reports by report day: the totals dip on a regular weekly cycle.

```{r dow-plot-data}
reports_by_day <- dow_triangle |>
  filter(reported_day <= max(day)) |>
  summarise(reports = sum(reported_onsets), .by = reported_day)

ggplot(reports_by_day, aes(x = reported_day, y = reports)) +
  geom_col() +
  labs(x = "Report day", y = "Reports", title = "Reports by report day")
```

::: {.callout-tip .optional collapse="true"}
## Interactive: the day-of-week sawtooth

```{=html}
<iframe class="nfidd-embed" src="../interactive/complex-reporting-processes.html?embed=dow" title="the day-of-week reporting effect interactive" loading="lazy" style="width:100%;height:780px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Set the weekly reporting weights to distort a smooth epidemic into a weekly sawtooth, then divide the normalised pattern back out to recover the underlying signal.
:::

The three effects act on different things.
The reference-date effect on the counts changes *when events happen* (the onset day), so it belongs in the model for the expected counts.
The reference-date effect on the delay changes *how long reporting takes* for events on a given onset day, so it belongs in the delay (reference) module.
The report-date effect changes *when reports are processed* (the report day), so it belongs in a separate report-date module, which our bespoke $p_{t,d}$ (a function of delay $d$ alone) had no way to express.
`epinowcast` lets us add all three in one model.

These extended models have a slightly harder posterior, so for them we add `adapt_delta = 0.95` to the fitting options (just as we did for the harder joint model in the [joint nowcasting session](joint-nowcasting)).

```{r fit-opts-hard}
fit_opts_hard <- enw_fit_opts(
  save_warmup = FALSE, pp = TRUE,
  chains = 4, parallel_chains = 4,
  iter_warmup = 500, iter_sampling = 500,
  adapt_delta = 0.95,
  show_messages = FALSE, refresh = 0
)
```

```{r dow-fit, message = FALSE, results = "hide"}
dow_long <- enw_filter_report_dates(
  to_enw_format(dow_triangle), latest_date = nowcast_date
)
dow_pobs <- enw_preprocess_data(dow_long, max_delay = 15)

dow_nowcast <- epinowcast(
  dow_pobs,
  # Reference-side day-of-week effect on the expected onsets
  expectation = enw_expectation(~ rw(day) + day_of_week, data = dow_pobs),
  # Reference-side day-of-week effect on the delay distribution
  reference = enw_reference(~ day_of_week, data = dow_pobs),
  # Report-side day-of-week reporting effect
  report = enw_report(~ (1 | day_of_week), data = dow_pobs),
  obs = enw_obs(family = "poisson", data = dow_pobs),
  fit = fit_opts_hard
)
```

```{r dow-result-plot}
dow_latest <- enw_latest_data(to_enw_format(dow_triangle))
plot(dow_nowcast, latest_obs = dow_latest)
```

::: {.callout-note}
## Take 3 minutes
This model carries three day-of-week terms: `day_of_week` in the expectation module, `day_of_week` in the reference (delay) module, and `(1 | day_of_week)` in the report module.
Why are these different effects, and why was the report-date one awkward to add to the hand-rolled Stan model?
:::

::: {.callout-note collapse="true"}
## Solution

The three terms act on different parts of the process.
`day_of_week` in `enw_expectation()` models genuine variation in *when events occur* by reference (onset) day of week.
`day_of_week` in `enw_reference()` lets the *delay distribution* differ by reference day of week, capturing that (here) weekend onsets are reported more slowly; it changes the shape of the delay for those onset days, not how many onsets there are.
`(1 | day_of_week)` in `enw_report()` is a random effect on the *report* date, one level per day of the week, scaling how many reports are processed on each calendar day; it absorbs the weekly reporting dip so it does not distort the nowcast of the underlying onsets.
The two reference-date terms both key on the onset day of week but are still separable because one shifts the level of onsets and the other reshapes their delay; fitted together on this simulation the sampler converges without trouble.
In the bespoke model the reporting probabilities depended only on the delay $d$ and had to sum to one for each reference date, so making them also depend on the calendar day of the report (while staying a valid distribution) meant restructuring the whole delay component.
Here each effect is one extra formula on the relevant module.
:::

# Extending the model: a time-varying delay

For our second worked example we relax the assumption that the delay distribution is the same on every day.
Reporting delays may lengthen when a system is under strain and shorten when it has spare capacity, so a single fixed $p_{t,d}$ cannot represent a delay that drifts across the outbreak.

We simulate this by letting the mean of the reporting delay grow slowly with the onset day, so later onsets may be reported with longer delays.

```{r tvd-simulate}
tvd_triangle <- onset_df |>
  filter(day < cutoff) |>
  rowwise() |>
  mutate(
    reporting_delay = list({
      # meanlog drifts upwards over the outbreak: delays may lengthen over time
      meanlog_t <- 1 + 0.5 * day / cutoff
      tibble(
        d = 0:15,
        reporting_delay = censored_delay_pmf(
          rlnorm, max = 15, meanlog = meanlog_t, sdlog = 0.5
        )
      )
    })
  ) |>
  ungroup() |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    reported_onsets = rpois(n(), onsets * reporting_delay)
  )
```

This is a property of the *reference* (onset) date, so we put structure on the **reference module** rather than on the report module.
Where the course model used `enw_reference(~ 1, ...)` for a constant delay, we now let the delay distribution drift week to week with `enw_reference(~ rw(week), ...)`.

::: {.callout-warning}
We simulated a roughly *linear* drift, so we could fit a model with a matching linear trend on the delay parameters and recover it almost perfectly.
That would be cheating: in a real analysis we do not know the functional form of the drift, because we did not generate the data.
We therefore deliberately choose a *flexible* model, a random walk on the delay parameters, that can track a drift without us baking in the exact form we happened to simulate.
:::

```{r tvd-fit, message = FALSE, results = "hide"}
tvd_long <- enw_filter_report_dates(
  to_enw_format(tvd_triangle), latest_date = nowcast_date
)
tvd_pobs <- enw_preprocess_data(tvd_long, max_delay = 15)

tvd_nowcast <- epinowcast(
  tvd_pobs,
  expectation = enw_expectation(~ rw(day), data = tvd_pobs),
  # Delay distribution now drifts week to week (this is the new part)
  reference = enw_reference(~ rw(week), data = tvd_pobs),
  obs = enw_obs(family = "poisson", data = tvd_pobs),
  fit = fit_opts_hard
)
```

```{r tvd-result-plot}
tvd_latest <- enw_latest_data(to_enw_format(tvd_triangle))
plot(tvd_nowcast, latest_obs = tvd_latest)
```

::: {.callout-note}
## Take 3 minutes
We changed the delay from `enw_reference(~ 1, ...)` to `enw_reference(~ rw(week), ...)`.
Why does letting the delay vary over time matter for the nowcast, and what would happen if we kept a fixed delay when the true delay was lengthening?
:::

::: {.callout-note collapse="true"}
## Solution

`rw(week)` lets the parameters of the delay distribution change from week to week, so the model can learn that recent onsets are being reported more slowly.
If we forced a fixed delay when the true delay was lengthening, the model would assume recent dates are more complete than they really are and so would *under*-nowcast the most recent counts, biasing the picture of the epidemic exactly where it matters most.
This is the same flexibility we used for the expectation module, applied to the delay instead.
:::

# Extending the model: weekly (batched) reporting

For our third worked example the data are *reported* less often than they occur.
Many surveillance systems release updates weekly rather than daily, so counts arrive in weekly batches even though events happen every day.
There are two ways to handle this in `epinowcast`, and they trade resolution against simplicity.

- **Weekly reporting of weekly data ("weekly-weekly").** Aggregate everything to a weekly timestep and model a weekly process. This is the simplest and cheapest option, but it can only ever produce weekly nowcasts.
- **Weekly reporting of daily data ("weekly-daily").** Keep a daily process but tell the model that reports only land once a week, using a *structural* reporting schedule. This recovers daily resolution from weekly reports at a higher computational cost.

We show both here on the same simulated series.
`epinowcast`'s [temporal aggregation guide](https://package.epinowcast.org/articles/temporal-aggregation.html) works through these patterns in full, including a daily benchmark and a scoring comparison; we give the compact version and link there rather than reproducing all of it.

## Weekly-weekly

We reuse the course reporting triangle and aggregate it to a weekly reporting schedule with `enw_aggregate_cumulative()`, which sums both reference and report dates into weekly steps.

```{r weekly-aggregate}
weekly_long <- enw_aggregate_cumulative(enw_long, timestep = "week")
head(weekly_long)
```

We then preprocess with a weekly timestep, so `epinowcast` knows the reporting schedule is weekly rather than daily.
With a weekly timestep the maximum delay is now counted in weeks, so we model up to three weeks of delay.

```{r weekly-preprocess}
weekly_pobs <- enw_preprocess_data(
  weekly_long, max_delay = 3, timestep = "week"
)
plot(weekly_pobs)
```

The model is the course model again; only the timestep of the data has changed.

```{r weekly-fit, message = FALSE, results = "hide"}
weekly_nowcast <- epinowcast(
  weekly_pobs,
  expectation = enw_expectation(~ rw(week), data = weekly_pobs),
  reference = enw_reference(~ 1, data = weekly_pobs),
  obs = enw_obs(family = "poisson", data = weekly_pobs),
  fit = fit_opts_hard
)
```

As before we compare against the eventually-complete data, aggregated to the same weekly timestep.

```{r weekly-result-plot}
weekly_latest <- enw_latest_data(
  enw_aggregate_cumulative(
    to_enw_format(reporting_triangle), timestep = "week"
  )
)
plot(weekly_nowcast, latest_obs = weekly_latest)
```

## Weekly-daily (structural reporting)

Now we keep the *daily* process but model the fact that reports only arrive once a week.
The idea is to build a daily grid where the cumulative count is only observed on the reporting day (here Wednesday) and carried forward on the other days, then tell the model which cells are genuine observations.
We start from the daily reporting triangle we fit to, blank out the non-Wednesday cumulative counts, and let `enw_impute_na_observations()` carry the last Wednesday value forward; `enw_flag_observed_observations()` records which cells are real in a `.observed` column.

```{r weekly-daily-scaffold}
weekly_rep_data <- enw_complete_dates(enw_long, timestep = "day")
weekly_rep_data[, day_of_week := weekdays(report_date)]
weekly_rep_data[
  , confirm := fifelse(day_of_week == "Wednesday", confirm, NA_real_)
]
weekly_rep_data <- weekly_rep_data |>
  enw_flag_observed_observations() |>
  enw_impute_na_observations() |>
  enw_filter_reference_dates_by_report_start() |>
  enw_add_incidence()

weekly_rep_pobs <- weekly_rep_data |>
  enw_complete_dates(timestep = "day") |>
  enw_preprocess_data(max_delay = 15, timestep = "day")
```

We describe the known Wednesday-only reporting cycle with `enw_dayofweek_structural_reporting()` and pass it to `enw_report()` through its `structural` argument instead of a formula.
Because most daily cells are imputed rather than observed, we also point the observation model at the `.observed` indicator so the likelihood only uses the real Wednesday reports.

```{r weekly-daily-fit, message = FALSE, results = "hide"}
# Structural variants need a little less adapt_delta and more tree depth
fit_opts_structural <- enw_fit_opts(
  save_warmup = FALSE, pp = TRUE,
  chains = 4, parallel_chains = 4,
  iter_warmup = 500, iter_sampling = 500,
  adapt_delta = 0.9, max_treedepth = 12,
  show_messages = FALSE, refresh = 0
)

structural <- enw_dayofweek_structural_reporting(
  weekly_rep_pobs, day_of_week = "Wednesday"
)

weekly_daily_nowcast <- epinowcast(
  weekly_rep_pobs,
  expectation = enw_expectation(
    ~ rw(week) + (1 | day_of_week), data = weekly_rep_pobs
  ),
  report = enw_report(structural = structural, data = weekly_rep_pobs),
  obs = enw_obs(
    family = "poisson", observation_indicator = ".observed",
    data = weekly_rep_pobs
  ),
  fit = fit_opts_structural
)
```

The nowcast is now on the daily scale even though reports only arrived weekly, so we compare it against the eventually-complete daily data.

```{r weekly-daily-plot}
weekly_daily_latest <- enw_latest_data(to_enw_format(reporting_triangle))
plot(weekly_daily_nowcast, latest_obs = weekly_daily_latest)
```

The weekly-weekly model gave us a weekly nowcast from a small, cheap model, while the weekly-daily model recovers a daily nowcast from the same weekly reports at the price of a larger reporting triangle and a harder fit.
The [temporal aggregation guide](https://package.epinowcast.org/articles/temporal-aggregation.html) scores these approaches against a fully daily benchmark if you want to see how much each one gives up.

::: {.callout-note}
## Take 3 minutes
The weekly-weekly model only changed `timestep = "week"` in preprocessing, while the weekly-daily model kept a daily timestep and instead described the reporting schedule with a structural report module.
Why is the reporting schedule a property of the *data* rather than of the model, and what does the weekly-daily approach buy you over aggregating to weeks?
:::

::: {.callout-note collapse="true"}
## Solution

The reporting schedule describes *when reports arrive*, which is a feature of how the data were collected, so it belongs to the data and its timestep.
The weekly-weekly approach maps both the process and the delay onto a weekly grid, which is simple and cheap but can only produce weekly nowcasts and gives the process model few points to learn from.
The weekly-daily approach keeps a daily latent process and encodes the once-a-week reporting through the structural report module, so it recovers daily resolution from weekly reports; the cost is a larger daily reporting triangle and a harder fit.
If we instead treated weekly batches as if they were ordinary daily data we would tell the model that almost every day had zero reports with occasional spikes, badly misrepresenting the delay distribution and the completeness of recent data.
:::

# Other reporting complications

We have worked through three complications end to end: a day-of-week effect, a time-varying delay, and weekly reporting.
Other complications are handled by the same modular idea, each a change to one formula rather than a new model.
Two worth knowing about are delays that vary by strata (for example age or region), which add the stratifying variable to a module formula such as `enw_reference(~ (1 | age_group), data = pobs)` so each stratum has its own delay while borrowing strength across strata, and missing reference dates, which `epinowcast` models with a dedicated module.
We link the relevant documentation for both in *Going further* below rather than fitting them here.

::: {.callout-note}
## Take 5 minutes
Suppose you also had the data split by two age groups, and you suspected reporting was slower in one of them.
Which module would you change, and what would be the advantage of fitting both age groups in one `epinowcast` model rather than two separate nowcasts?
:::

::: {.callout-note collapse="true"}
## Solution

You would add age group to the *reference* module so the delay distribution can differ by group.
Fitting both groups jointly lets a sparse group borrow strength from the other through shared (partially pooled) structure, which is more stable than fitting each group on its own, especially when counts are low.
:::

# Going further

This session only scratches the surface of what `epinowcast` can do, and `epinowcast` is only one of several tools for this kind of modelling.

`epinowcast` has extensive documentation and worked case studies that go well beyond this session; rather than try to cover everything here, the best next step is to read them:

- The [`epinowcast` package website](https://package.epinowcast.org/) [@epinowcastPackage2024], including its [model description](https://package.epinowcast.org/articles/model.html) and the [getting started vignette](https://package.epinowcast.org/articles/epinowcast.html).
- The [model features summary](https://package.epinowcast.org/articles/features.html), a table of everything `epinowcast` currently supports with links to the relevant documentation for each feature.

To go deeper on the topics we covered, the relevant reference pages are:

- **Reporting schedule (weekly or batched reporting):** the [temporal aggregation guide](https://package.epinowcast.org/articles/temporal-aggregation.html), [`enw_aggregate_cumulative()`](https://package.epinowcast.org/reference/enw_aggregate_cumulative.html) and the `timestep` argument of [`enw_preprocess_data()`](https://package.epinowcast.org/reference/enw_preprocess_data.html).
- **Day-of-week and other report-date effects:** the [`enw_report()`](https://package.epinowcast.org/reference/enw_report.html) report-date module.
- **Time-varying delay (reference-date effects):** the [`enw_reference()`](https://package.epinowcast.org/reference/enw_reference.html) reference-date module.
- **Delays that vary by strata:** the [hierarchical age-stratified nowcasting case study](https://package.epinowcast.org/articles/germany-age-stratified-nowcasting.html), which fits many reporting triangles at once with partially pooled effects.
- **Missing reference dates:** the [`enw_missing()`](https://package.epinowcast.org/reference/enw_missing.html) module.
- The [plotting reference](https://package.epinowcast.org/reference/plot.epinowcast.html) for more ways to visualise a fit than the default `plot()` we used here.

Its case studies and vignettes also work through joint nowcasting with $R_t$ estimation (a renewal-based expectation model, generalising the joint model from the [joint nowcasting session](joint-nowcasting)), and evaluation of nowcasts with proper scoring rules (as we did for forecasts in the [combining nowcasting and forecasting session](forecasting-nowcasting)).

`epinowcast` is not the only option, and the right tool depends on the problem.

- The bespoke Stan models we built in this course remain a perfectly good choice when you understand the reporting process and want full control.
- The [`baselinenowcast`](https://baselinenowcast.epinowcast.org) package implements simple multiplicative nowcasting methods that make useful benchmarks.
- A range of other methods and frameworks are used in practice; the collaborative nowcasting comparison in [@wolfframCollaborativeNowcastingCOVID192023] surveys several and discusses what drives differences in their performance.

For the methodological background to generative nowcasting models of this kind, see [@lisonGenerativeBayesianModeling2024].

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
