---
title: "Modelling complex reporting processes"
order: 14.5
bibliography: ../nfidd.bib
---

# Introduction

In the [nowcasting concepts session](nowcasting) and the [joint nowcasting session](joint-nowcasting) we built our own nowcasting models in Stan.
We started from a known delay distribution, then learned to jointly estimate the delay and the nowcast from the reporting triangle, and finally added a renewal process so that the generative model of transmission informed the nowcast.

Those bespoke models were ideal for *learning* how nowcasting works because we could see every part of the model.
They are also a perfectly valid choice for real analyses: there is nothing wrong with writing your own model when you understand the reporting process well.
However, real surveillance data is usually messier than those models handle comfortably.
In this session we step back and ask what happens when the reporting process itself becomes complicated.

We introduce [`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] as one flexible, well-tested framework for exactly these situations.
We first reproduce the joint nowcasting model we already built, but expressed in `epinowcast`, so that the familiar model appears in the new tool.
We then *extend* it to handle two reporting complications that the bespoke model cannot capture comfortably: a day-of-week reporting effect and a reporting delay that drifts over time.
Throughout, we simulate data with the feature, fit a model that handles it, and look at the result.

## Slides

- There are no slides for this draft session yet.

## Objectives

This session aims to show why real reporting processes are harder to model than our bespoke Stan models assume, to reproduce the joint nowcasting model in `epinowcast`, and to extend it to handle complex reporting features through worked examples.

::: {.callout-note collapse="true"}

# Setup

## Source file

The source file of this session is located at `sessions/complex-reporting-processes.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to load infection times and the simulation helpers we have used throughout, the `dplyr` and `tidyr` packages for data wrangling, and the `ggplot2` library for plotting.
The worked examples additionally use the `epinowcast` package and the `data.table` package that it uses for its data structures.

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
library("epinowcast")
library("data.table")
```

::: {.callout-tip}
`epinowcast` compiles its models with `cmdstanr` the first time a model is fitted, which needs a working CmdStan toolchain (the same one used elsewhere in this course).
Installation and setup instructions, including how to install CmdStan, are on the [package website](https://package.epinowcast.org/).
The fits in this session are deliberately small and fast so that the course builds quickly; a real analysis would use more iterations and more chains.
:::

## Initialisation

We set a random seed for reproducibility.
We also tell `epinowcast` not to warn about partially specified initial conditions, which keeps the output tidy.

```{r initialise}
set.seed(123)
options(cmdstanr_warn_inits = FALSE)
```

:::

# Why real reporting is harder than our models assume

## The model we have built so far

In the [joint nowcasting session](joint-nowcasting) we modelled each cell of the reporting triangle as

$$
  n_{t,d} \mid \lambda_{t}, p_{t,d} \sim \text{Poisson}\left(\lambda_{t} \times p_{t,d}\right),
$$

where $\lambda_{t}$ is the expected number of onsets on day $t$ and $p_{t,d}$ is the probability that an onset on day $t$ is reported with a delay of $d$ days.
We modelled $\lambda_{t}$ with a geometric random walk (or a renewal process) and $p_{t,d}$ with a single delay distribution that was the same on every day.

This works well in our simulated examples because we simulated the data from almost exactly that model.
Real data rarely cooperate so neatly.
Routine surveillance data often shows features our bespoke model does not capture well, including day-of-week reporting effects, reporting delays that vary over time or by strata such as age or region, weekly or batched reporting, and missing or late reports.
Rather than list these problems and tell you they are hard, in this session we *show* a couple of them: we simulate the feature, fit a model that handles it, and look at the result.

::: {.callout-note}
## Take 5 minutes
Think about the joint nowcasting model you built.
What would you have to change in that Stan code to add a day-of-week reporting effect *and* let the delay distribution change over the course of the outbreak?
:::

::: {.callout-note collapse="true"}
## Solution

There is no single right answer, but the changes add up quickly.

- You would need to index the reporting probabilities by report date (or day of week) as well as by delay, turning the delay vector into a more complex structure.
- You would need a model for how the delay distribution changes over time, for example a random walk on the parameters of the delay distribution, with its own priors.
- You would need to keep the reporting probabilities on each day summing to one, which is awkward once they vary by date.
- Each addition makes the Stan model longer, slower to fit, and harder to debug and test.

None of this is impossible, and writing it yourself remains a reasonable option.
But it is the point at which reaching for a general, tested framework that already implements these features lets us focus on the modelling choices rather than the implementation.
:::

# Introducing `epinowcast`

[`epinowcast`](https://package.epinowcast.org/) [@epinowcastPackage2024] is a Bayesian framework for real-time infectious disease surveillance.
It is *one* of several tools for this kind of modelling (we point to others at the end of the session), and it is a direct generalisation of the joint nowcasting model we built: it works from the reporting triangle and jointly estimates the expected counts and the reporting process, but it makes each component far more flexible and exposes them through a formula interface rather than hand-written Stan code.

`epinowcast` decomposes the problem into modular pieces, each of which mirrors something we have already seen.

- An **expectation** model for the expected counts $\lambda_{t}$ (the analogue of our geometric random walk or renewal process).
- A **reference date** model for the delay distribution by date of the epidemiological event (the analogue of our $p_{t,d}$, but allowed to vary over time and by covariates).
- A **report date** model for effects that act on the *report* date, such as day-of-week reporting effects.

Each piece is specified with an R formula, so adding a day-of-week effect or a time-varying delay is a one-line change rather than a rewrite of the model.
Under the hood `epinowcast` still builds and fits a Stan model, but the model is tested, optimised, and reused across problems.

::: {.callout-tip}
We are treating `epinowcast` as a tool here rather than dissecting its internals.
The mapping back to the bespoke model is deliberate: if you understand the joint nowcasting session, you already understand what `epinowcast` is doing, just with more flexible components.
:::

# The course model, in `epinowcast`

Before adding anything new, let us reproduce the joint nowcasting model we already built, but in `epinowcast`.
We simulate exactly as in the [joint nowcasting session](joint-nowcasting): the same onset time series, the same fixed reporting delay, and the same reporting triangle.

```{r simulate-onsets}
gen_time_pmf <- make_gen_time_pmf()
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), ip_pmf
)
cutoff <- 71

reporting_delay_pmf <- censored_delay_pmf(
  rlnorm, max = 15, meanlog = 1, sdlog = 0.5
)

reporting_triangle <- onset_df |>
  filter(day < cutoff) |>
  mutate(
    reporting_delay = list(
      tibble(d = 0:15, reporting_delay = reporting_delay_pmf)
    )
  ) |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    reported_onsets = rpois(n(), onsets * reporting_delay)
  )
```

As we only observe reports up to the present, we keep just the cells that have been reported by our cutoff.

```{r filter-triangle}
filtered_triangle <- reporting_triangle |>
  filter(reported_day <= max(day))
```

## Putting the data into `epinowcast` format

`epinowcast` expects data in a "long" format with one row per (reference date, report date) pair, where `confirm` is the **cumulative** count for that reference date as known on that report date.
This is the same reporting triangle we have been using, just with calendar dates instead of integer days and cumulative rather than incremental counts.

We attach arbitrary calendar dates (so that, later, weekends line up with a day-of-week effect) and form the cumulative counts.
Because we use the same transformation in each example below, we wrap it in a small helper.

```{r to-enw-helper}
# Day 0 of the simulation is a Monday
ref_date0 <- as.Date("2023-01-02")

to_enw_format <- function(triangle) {
  triangle |>
    filter(reported_day <= max(day)) |>
    transmute(
      reference_date = ref_date0 + day,
      report_date = ref_date0 + reported_day,
      count = reported_onsets
    ) |>
    arrange(reference_date, report_date) |>
    group_by(reference_date) |>
    mutate(confirm = cumsum(count)) |>
    ungroup() |>
    select(reference_date, report_date, confirm) |>
    as.data.table()
}

enw_long <- to_enw_format(reporting_triangle)
head(enw_long)
```

::: {.callout-note}
## Take 2 minutes
How does this data structure relate to the reporting triangle from the [joint nowcasting session](joint-nowcasting)?
:::

::: {.callout-note collapse="true"}
## Solution

It is the same idea in a different layout.
Each `reference_date` is a row of the triangle, each `report_date` gives a column, and `confirm` is the cumulative version of the cell counts $n_{t,d}$ we summed over delays before.
`epinowcast` works with the cumulative counts and converts them into the reporting triangle internally during preprocessing.
:::

## Preprocessing and fitting

`epinowcast` preprocesses the data into its internal reporting-triangle representation, where we also choose the maximum delay we will model.

```{r preprocess}
pobs <- enw_preprocess_data(enw_long, max_delay = 15)
```

Now we specify the model.
For the course model we need just two modules, each mapping directly onto a choice from the joint nowcasting session: a weekly random walk for the expected counts and a single, constant delay distribution.

```{r course-modules}
# Expected counts: a weekly random walk (cf. our geometric random walk)
expectation_module <- enw_expectation(~ rw(week), data = pobs)

# Delay distribution by reference date: a single parametric delay, constant
# over time (cf. our fixed p_{t,d})
reference_module <- enw_reference(~ 1, data = pobs)
```

We collect a small set of fitting options that we reuse across the session.
These play the same role as the arguments to `nfidd_sample()` did in earlier sessions: number of chains, warmup and sampling iterations, and so on.
We use one chain and few iterations purely to keep the example quick.

```{r fit-opts}
fit_opts <- enw_fit_opts(
  save_warmup = FALSE, pp = TRUE,
  chains = 1, iter_warmup = 500, iter_sampling = 500,
  show_messages = FALSE, refresh = 0
)
```

We then fit the model.
The first fit compiles the Stan model, so it is the slowest step; subsequent fits in this session reuse the compiled model.

```{r course-fit, message = FALSE, results = "hide"}
course_nowcast <- epinowcast(
  pobs,
  expectation = expectation_module,
  reference = reference_module,
  fit = fit_opts
)
```

::: {.callout-note}
## Take 3 minutes
Compare these two formulas to the choices we made in the [joint nowcasting session](joint-nowcasting).
Which module corresponds to $\lambda_{t}$ and which to $p_{t,d}$?
:::

::: {.callout-note collapse="true"}
## Solution

- `enw_expectation()` plays the role of $\lambda_{t}$: `rw(week)` is a random walk on the expected counts, much like the geometric random walk we used for onsets.
- `enw_reference()` plays the role of the delay distribution $p_{t,d}$: `~ 1` keeps it constant over time, exactly as in our bespoke model.

This is the *same* joint nowcasting model we built by hand, now expressed in two lines.
Everything that follows is an extension of one of these two modules, or the addition of a third one for report-date effects.
:::

## Plotting the nowcast

`epinowcast` objects come with `plot()` and `summary()` methods, so we do not have to extract draws by hand as we did with our bespoke fits.
Plotting the fit shows the nowcast for the recent, incomplete reference dates (the median and credible intervals) against the counts that were actually available at the nowcast date.

```{r course-plot}
plot(course_nowcast)
```

As in our bespoke nowcasts, the most recent dates are the most uncertain because they are the least complete: the points (data available at the nowcast date) fall well below the nowcast for recent dates, which is exactly the right-truncation the nowcast corrects for.
The plotting methods are much more flexible than this; see the [`epinowcast` plotting documentation](https://package.epinowcast.org/reference/plot.epinowcast.html) for options such as plotting particular reference dates or the posterior predictive fit.

# Extending the model: a day-of-week reporting effect

Now we add a complication the bespoke model could not handle comfortably.
Real systems often report less at weekends, so the number of reports landing on a given day depends on the **day of the week of the report date**, not on the delay alone.
We capture this in the simulation with a multiplicative weekly weight, assuming day 0 of our simulation is a Monday.

```{r dow-simulate}
# Multiplicative reporting weight for Mon, Tue, ..., Sun
# Reporting is suppressed at the weekend (Sat/Sun)
dow_effect <- c(1, 1, 1, 1, 0.9, 0.4, 0.5)

dow_triangle <- onset_df |>
  filter(day < cutoff) |>
  mutate(
    reporting_delay = list(
      tibble(d = 0:15, reporting_delay = reporting_delay_pmf)
    )
  ) |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    # day of week of the *report* day (0 = Monday)
    report_dow = reported_day %% 7,
    dow_weight = dow_effect[report_dow + 1],
    reported_onsets = rpois(n(), onsets * reporting_delay * dow_weight)
  )
```

We can see the effect clearly if we sum the reports by report day: the totals dip on a regular weekly cycle.

```{r dow-plot-data}
reports_by_day <- dow_triangle |>
  filter(reported_day <= max(day)) |>
  summarise(reports = sum(reported_onsets), .by = reported_day)

ggplot(reports_by_day, aes(x = reported_day, y = reports)) +
  geom_col() +
  labs(x = "Report day", y = "Reports", title = "Reports by report day")
```

This is a property of the *report date*: fewer reports are processed at weekends, regardless of when the onset occurred.
Our bespoke $p_{t,d}$ only depends on the delay $d$ since onset, so it has no way to know which calendar day a report lands on.
To reproduce the weekly pattern the model needs a term that depends on the report date itself: the **report-date module** in `epinowcast`.

We preprocess the new data and add a report-date module to the two modules from the course model.
The day-of-week effect is a single extra formula.

```{r dow-fit, message = FALSE, results = "hide"}
dow_pobs <- enw_preprocess_data(to_enw_format(dow_triangle), max_delay = 15)

dow_nowcast <- epinowcast(
  dow_pobs,
  expectation = enw_expectation(~ rw(week), data = dow_pobs),
  reference = enw_reference(~ 1, data = dow_pobs),
  # Report-date effects: a day-of-week reporting effect (this is the new part)
  report = enw_report(~ (1 | day_of_week), data = dow_pobs),
  fit = fit_opts
)
```

```{r dow-result-plot}
plot(dow_nowcast)
```

::: {.callout-note}
## Take 3 minutes
The only change from the course model is the `enw_report(~ (1 | day_of_week))` line.
What is this term doing, and why was it awkward to add to the hand-rolled Stan model?
:::

::: {.callout-note collapse="true"}
## Solution

`enw_report(~ (1 | day_of_week))` adds a random effect on the *report* date, one level per day of the week, that scales how many reports land on each calendar day.
It absorbs the weekly dip we built into the simulation so that it does not distort the nowcast of the underlying onsets.
In the bespoke model the reporting probabilities depended only on the delay $d$ and had to sum to one for each reference date, so making them also depend on the calendar day of the report (while staying a valid distribution) meant restructuring the whole delay component.
Here it is one extra formula on a separate module.
:::

# Extending the model: a time-varying delay

For our second worked example we relax the assumption that the delay distribution is the same on every day.
Reporting delays lengthen when a system is under strain and shorten when it has spare capacity, so a single fixed $p_{t,d}$ cannot represent a delay that drifts across the outbreak.

We simulate this by letting the mean of the reporting delay grow slowly with the onset day, so later onsets are reported with longer delays.

```{r tvd-simulate}
tvd_triangle <- onset_df |>
  filter(day < cutoff) |>
  rowwise() |>
  mutate(
    reporting_delay = list({
      # meanlog drifts upwards over the outbreak: delays lengthen over time
      meanlog_t <- 1 + 0.5 * day / cutoff
      tibble(
        d = 0:15,
        reporting_delay = censored_delay_pmf(
          rlnorm, max = 15, meanlog = meanlog_t, sdlog = 0.5
        )
      )
    })
  ) |>
  ungroup() |>
  unnest(reporting_delay) |>
  mutate(
    reported_day = day + d,
    reported_onsets = rpois(n(), onsets * reporting_delay)
  )
```

This is a property of the *reference* (onset) date, so we put structure on the **reference module** rather than on the report module.
Where the course model used `enw_reference(~ 1, ...)` for a constant delay, we now let the delay distribution drift week to week with `enw_reference(~ rw(week), ...)`.

```{r tvd-fit, message = FALSE, results = "hide"}
tvd_pobs <- enw_preprocess_data(to_enw_format(tvd_triangle), max_delay = 15)

tvd_nowcast <- epinowcast(
  tvd_pobs,
  expectation = enw_expectation(~ rw(week), data = tvd_pobs),
  # Delay distribution now drifts week to week (this is the new part)
  reference = enw_reference(~ rw(week), data = tvd_pobs),
  fit = fit_opts
)
```

```{r tvd-result-plot}
plot(tvd_nowcast)
```

::: {.callout-note}
## Take 3 minutes
We changed the delay from `enw_reference(~ 1, ...)` to `enw_reference(~ rw(week), ...)`.
Why does letting the delay vary over time matter for the nowcast, and what would happen if we kept a fixed delay when the true delay was lengthening?
:::

::: {.callout-note collapse="true"}
## Solution

`rw(week)` lets the parameters of the delay distribution change from week to week, so the model can learn that recent onsets are being reported more slowly.
If we forced a fixed delay when the true delay was lengthening, the model would assume recent dates are more complete than they really are and so would *under*-nowcast the most recent counts, biasing the picture of the epidemic exactly where it matters most.
This is the same flexibility we used for the expectation module, applied to the delay instead.
:::

# Other reporting complications

We have worked through two complications end to end.
The others from the start of the session are handled by the same modular idea: each is a change to one formula, not a new model.
We give the formula sketches here rather than fitting every one, but they slot into exactly the same `epinowcast()` call.

- **Delays that vary by strata.**
Add the stratifying variable to the relevant module formula, for example a random effect of age group on the reference module (`enw_reference(~ (1 | age_group), data = pobs)`), so each stratum can have its own delay while borrowing strength across strata.
- **Weekly or batched reporting.**
This is another report-date pattern: counts arrive in lumps on particular report days, which a report-date model can represent.
- **Missing and late reporting.**
`epinowcast` has dedicated support for missing reference dates and can model the proportion of reports that are eventually observed, rather than assuming every report arrives within the maximum delay.

::: {.callout-note}
## Take 5 minutes
Suppose you also had the data split by two age groups, and you suspected reporting was slower in one of them.
Which module would you change, and what would be the advantage of fitting both age groups in one `epinowcast` model rather than two separate nowcasts?
:::

::: {.callout-note collapse="true"}
## Solution

You would add age group to the *reference* module so the delay distribution can differ by group.
Fitting both groups jointly lets a sparse group borrow strength from the other through shared (partially pooled) structure, which is more stable than fitting each group on its own, especially when counts are low.
:::

# Going further

This session only scratches the surface of what `epinowcast` can do, and `epinowcast` is only one of several tools for this kind of modelling.

`epinowcast` has extensive documentation and worked case studies that go well beyond this session; rather than try to cover everything here, the best next step is to read them:

- The [`epinowcast` package website](https://package.epinowcast.org/) [@epinowcastPackage2024], including its [model description](https://package.epinowcast.org/articles/model.html) and the [getting started vignette](https://package.epinowcast.org/articles/epinowcast.html).
- The [plotting reference](https://package.epinowcast.org/reference/plot.epinowcast.html) for more ways to visualise a fit than the default `plot()` we used here.
- Its case studies and vignettes work through stratified nowcasting (many reporting triangles at once with partially pooled effects), richer delay and reporting models, joint nowcasting with $R_t$ estimation (a renewal-based expectation model, generalising the joint model from the [joint nowcasting session](joint-nowcasting)), and evaluation of nowcasts with proper scoring rules (as we did for forecasts in the [combining nowcasting and forecasting session](forecasting-nowcasting)).

`epinowcast` is not the only option, and the right tool depends on the problem.

- The bespoke Stan models we built in this course remain a perfectly good choice when you understand the reporting process and want full control.
- The [`baselinenowcast`](https://baselinenowcast.epinowcast.org) package implements simple multiplicative nowcasting methods that make useful benchmarks.
- A range of other methods and frameworks are used in practice; the collaborative nowcasting comparison in [@wolfframCollaborativeNowcastingCOVID192023] surveys several and discusses what drives differences in their performance.

For the methodological background to generative nowcasting models of this kind, see [@lisonGenerativeBayesianModeling2024].

# Wrap up

- Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
- Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
