---
title: "Jointly fitting multiple data streams"
order: 14.7
bibliography: ../nfidd.bib
---

# Introduction

So far we have estimated the reproduction number from a *single* observed time series: symptom onsets, generated as a convolution of a shared infection process with one delay distribution.
In reality we often observe an outbreak through *several* surveillance streams at once: confirmed cases, deaths, and a wastewater signal, all generated by the same underlying epidemic but each seen through a different lens of a different delay, a different scaling, and a different observation model.

Rather than reach straight for a model that swallows all of these at once, in this session we follow a modelling *workflow*.
We start from the question we are trying to answer, sketch what an ideal model would look like, and then *build that model up in parts*: first fitting each data source on its own, then linking two together through a shared latent infection process, and finally exposing what happens when streams disagree.
This is the modelling idea behind tools such as `epidemia`, which jointly models cases and deaths from a shared renewal process [@bhattSemiMechanisticBayesianModeling2020].

## Slides

-   [Introduction to the reproduction number](slides/introduction-to-reproduction-number)

## Objectives

The aim of this session is to develop a joint model of several surveillance streams the way a modeller actually would: question first, then an ideal model, then building it up one stream at a time, checking what each step buys us and what happens when streams *agree* or *conflict*, all framed by a principled modelling workflow.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/multiple-data-streams.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to access stan models and helper functions, the `dplyr` and `tidyr` packages for data wrangling, the `ggplot2` library for plotting, the `tidybayes` package for extracting results of the inference, and the `purrr` package for [functional programming](https://adv-r.hadley.nz/fp.html).

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
library("tidybayes")
library("purrr")
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
We also set an option that makes `cmdstanr` show line numbers when printing model code.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(123)
options(cmdstanr_print_line_numbers = TRUE)
options(cmdstanr_warn_inits = FALSE)
```
:::

# The question we are trying to answer

Before reaching for a model it is worth being explicit about what we want from it.
Our **research question and target estimands** are the same as in earlier sessions: we want to recover the latent infection trajectory and the reproduction number $R_t$ that drives it, in as timely and as well-anchored a way as possible.

No single data stream gives us a perfect view of those quantities.

-   **Confirmed cases** are timely but depend on testing and reporting, and only a fraction of infections are ever ascertained.
-   **Deaths** are a more stable signal but arrive with a long delay and capture only a small, severity-dependent fraction of infections.
-   A **wastewater** signal does not depend on people seeking care at all, but it measures something more indirect (shed pathogen concentration) on its own scale.

Each of these is a different, delayed, scaled observation of the *same* underlying infections.
The promise of using them together is that the timely-but-noisy streams can sharpen our estimate of the recent trajectory, while the stable-but-delayed streams anchor the longer-term level, so that $R_t$ and infections are informed by *all* of the data at once.
The question for this session is therefore not just "can we fit several streams?" but "what does each stream add, and what happens when they pull in different directions?".

# What would the ideal model look like?

Fitting several data streams is not a one-off trick: it is one step in a wider, iterative workflow for building, checking and comparing infectious disease models [@abbottWorkflowInfectiousDisease2026].
We will let that workflow drive how we develop the model in this session, so it is worth looking at it before we write any code.

![A workflow for infectious disease modelling. We move from a research question through a process model and per-stream observation models to choices about how to integrate data sources and do inference, and back again -- model checking, identifiability and conflict between data sources all feed revisions. Figure by Sam Abbott, reproduced from the `a-workflow-for-infectious-disease-modelling` repository [@abbottWorkflowInfectiousDisease2026] under the MIT licence.](figures/workflow-diagram.png)

Reading the diagram from the bottom-left, an ideal model for our question is assembled from a small number of pieces, and the rest of this session walks through them in order:

1.  **Research question and target estimands.** We fixed these above: latent infections and $R_t$.
2.  **Process model.** A single shared epidemic process. We reuse the renewal equation with a geometric random walk on $R_t$ from the [previous session](R-estimation-and-the-renewal-equation), so there is *one* infection trajectory underneath everything.
3.  **Data source selection.** Catalogue the streams available to us (cases, deaths, wastewater) and, crucially, what each one measures and how it is biased.
4.  **Observation model per source.** For *each* stream, a delay (infection to observation) and a scaling (what fraction of infections, on what scale) and a likelihood. These are separate sub-models, one per data source.
5.  **Data integration choices.** How to combine those per-stream observation models. This is the heart of the session, and the diagram flags two things about it: the *choice* of how to integrate sources, and a feedback arrow warning that *conflict between data sources reveals model misspecification*.
6.  **Inference, checking and revision.** Fit the model, check it against each stream, and let identifiability problems or conflict between sources send us back to revise it.

The ideal model, then, is *modular*: a shared process model with a separate, swappable observation model bolted on for each data source, combined by an explicit integration choice.
That modularity is exactly what lets us **build it in parts**, which is what we do next.
We will fit each observation model on its own first, then make the integration choice that links them, and only then stress-test that choice.

::: callout-note
## Parallel (joint) vs sequential modelling

The "data integration choices" box hides an important decision, and the distinction matters.

In a **parallel** model -- the approach we use in this session -- every stream is its own convolution *of the shared latent infections*, and the streams are *conditionally independent given those infections*.
Schematically:

$$
\text{cases}_t \sim f\big(\text{convolve}(I, p_\text{cases})\big), \qquad \text{deaths}_t \sim g\big(\text{convolve}(I, p_\text{deaths})\big)
$$

Both cases and deaths look *directly* at the same infections $I$, each through their own delay distribution $p$.
Once we condition on $I$, knowing the cases tells us nothing extra about the deaths: all the shared information flows through the infections.

In a **sequential** model, by contrast, one stream is modelled as a convolution *of another stream* rather than of the infections, for example

$$
\text{deaths}_t \sim g\big(\text{convolve}(\text{cases}, p)\big),
$$

chaining deaths off cases instead of off the shared infections.

We use the **parallel / conditionally-independent-given-infections** formulation: a single renewal process generates the latent infections, and each stream is a separate convolution of those infections with its own delay and scaling.
This keeps the infection process as the one shared quantity that every stream informs.
:::

# Simulating parallel data streams from a shared infection trajectory

As in previous sessions we first simulate the data-generating process so that we know the truth we are trying to recover.
This is the "data source selection" and per-stream "observation model" parts of the workflow made concrete: a single shared infection trajectory, with three parallel streams generated from it.

We use the same gamma-distributed generation time as in the [renewal equation session](R-estimation-and-the-renewal-equation), discretised with `censored_delay_pmf()` and renormalised so that the generation time is strictly positive.

```{r gen_time}
gen_time_pmf <- censored_delay_pmf(rgamma, max = 14, shape = 4, rate = 1)
gen_time_pmf <- gen_time_pmf[-1] ## remove first element
gen_time_pmf <- gen_time_pmf / sum(gen_time_pmf) ## renormalise
```

We define a reproduction number that starts above 1 and then drops below it, and use the `renewal()` function to generate the shared infection time series.

```{r simulate_infections}
n <- 80
I0 <- 5
R <- c(rep(1.4, 30), rep(0.8, n - 30))
infections <- renewal(I0, R, gen_time_pmf)

inf_ts <- tibble(day = seq_len(n), infections = infections)
ggplot(inf_ts, aes(x = day, y = infections)) +
  geom_line() +
  labs(title = "Shared latent infections", x = "Day", y = "Infections")
```

Now we create three parallel streams.
Each is a convolution of the *same* infections with its own delay distribution and scaling, observed with its own noise model.

-   **Cases**: a short infection-to-report delay, scaled by an *ascertainment* fraction (most infections are never confirmed), observed as counts.
-   **Deaths**: a long infection-to-death delay, scaled by the *infection fatality ratio* (IFR), observed as counts.
-   **Wastewater**: a short infection-to-shedding delay, scaled by a constant, observed on the log scale with log-normal noise.

```{r delays}
case_delay_pmf  <- censored_delay_pmf(rgamma, max = 10, shape = 3,  rate = 1)
death_delay_pmf <- censored_delay_pmf(rgamma, max = 21, shape = 16, rate = 1)
ww_delay_pmf    <- censored_delay_pmf(rgamma, max = 10, shape = 2,  rate = 1)
```

```{r simulate_streams}
ascertainment <- 0.4
ifr <- 0.01
ww_scale <- 2

exp_cases  <- ascertainment * convolve_with_delay(infections, case_delay_pmf)
exp_deaths <- ifr * convolve_with_delay(infections, death_delay_pmf)
exp_ww     <- ww_scale * convolve_with_delay(infections, ww_delay_pmf)

obs <- inf_ts |>
  mutate(
    cases  = rpois(n, exp_cases),
    deaths = rpois(n, exp_deaths),
    ww     = rnorm(n, log(exp_ww), 0.2)
  )
```

We can plot the three streams side by side.
Note how each one is a delayed, rescaled echo of the same infection curve: deaths peak later than cases, and the wastewater signal lives on a completely different scale.

```{r plot_streams}
obs |>
  select(day, cases, deaths, ww) |>
  pivot_longer(-day, names_to = "stream", values_to = "value") |>
  ggplot(aes(x = day, y = value)) +
  geom_line() +
  facet_wrap(~stream, scales = "free_y") +
  labs(title = "Three parallel observation streams", x = "Day", y = NULL)
```

::: callout-note
## Take 2 minutes

The wastewater stream is stored and plotted on the *log* scale (`ww`), whereas cases and deaths are counts.
Why might a log-normal observation model be a more natural choice for a continuous concentration measurement than a count distribution such as the Poisson?
:::

::: {.callout-note collapse="true"}
## Solution

A wastewater concentration is a positive, continuous measurement rather than a count of discrete events, so a count distribution does not apply directly.
Measurement error in such assays tends to be multiplicative (proportional to the signal) rather than additive, which a log-normal observation model captures well.
This is the observation model used by dedicated wastewater tools such as `EpiSewer` [@lisonRobustRealtimeEstimation2025].
:::

# Building the model in parts

We now build the ideal model up piece by piece, exactly as the workflow suggests: an observation model per source first, then the integration choice that links them, then a check of that choice.

To do this without writing (and compiling) a different model for every combination of streams, we use a *single* Stan model with a switch for each stream.
The model is a direct extension of the random-walk renewal model from the previous session: one shared geometric random walk for $R_t$ drives one shared infection process, each stream adds its own convolution, scaling, and likelihood term, and three integer flags (`use_cases`, `use_deaths`, `use_ww`) decide which streams contribute to the likelihood.

```{r stan_model}
mod <- nfidd_cmdstan_model("estimate-inf-and-r-multi-stream")
mod
```

::: callout-tip
## Take 5 minutes

Familiarise yourself with the model above.
Compare it to the random-walk model `estimate-inf-and-r-rw` from the [previous session](R-estimation-and-the-renewal-equation#estimating-r_t-with-a-geometric-random-walk-prior).
Which parts are shared across all three streams, and which parts are specific to each stream?
Where in the `model` block does each stream contribute to the likelihood, and what do the `use_*` flags do there?
:::

::: {.callout-note collapse="true"}
## Solution

The `R`, `infections`, generation time, random walk noise and random walk standard deviation are *shared*: there is a single latent infection process.
Each stream has its own delay PMF, its own scaling parameter (`ascertainment`, `ifr`, `ww_scale`) and its own observation term in the `model` block.
The three likelihood lines (`cases ~ poisson(...)`, `deaths ~ poisson(...)`, `ww ~ normal(log(...), ww_sigma)`) are each guarded by a `use_*` flag, so we can switch any stream on or off.
Because the streams are conditionally independent given the infections, whichever terms are switched on are simply *added* together to form the joint log-likelihood.
:::

We assemble a base list of data shared by every fit.
We always pass all three streams; the `use_*` flags decide which ones actually enter the likelihood.

```{r base_data}
base_data <- list(
  n = n,
  I0 = I0,
  gen_time_max = length(gen_time_pmf),
  gen_time_pmf = gen_time_pmf,
  cases = obs$cases,
  case_delay_max = length(case_delay_pmf) - 1,
  case_delay_pmf = case_delay_pmf,
  deaths = obs$deaths,
  death_delay_max = length(death_delay_pmf) - 1,
  death_delay_pmf = death_delay_pmf,
  ww = obs$ww,
  ww_delay_max = length(ww_delay_pmf) - 1,
  ww_delay_pmf = ww_delay_pmf
)
```

A small helper keeps the rest of the session tidy: it switches the requested streams on, fits the model, and uses the same sampler settings throughout.
As in the previous session we increase `max_treedepth` and provide initial values to help the sampler, because this is a more complex model.

```{r fit_helper}
fit_streams <- function(use_cases, use_deaths, use_ww, data = base_data) {
  data <- c(data, list(
    use_cases = use_cases, use_deaths = use_deaths, use_ww = use_ww
  ))
  nfidd_sample(
    mod, data = data, max_treedepth = 12,
    init = \() list(init_R = 1, rw_sd = 0.01)
  )
}
```

## Part 1: fit each data source on its own

The first thing the workflow asks of us is an observation model *per source*.
Before combining anything, we check that each stream, on its own, can recover the shared infection process through its own delay and scaling.

We fit the cases stream alone, then the deaths stream alone.

```{r fit_single, results = 'hide', message = FALSE}
fit_cases  <- fit_streams(use_cases = 1, use_deaths = 0, use_ww = 0)
fit_deaths <- fit_streams(use_cases = 0, use_deaths = 1, use_ww = 0)
```

We extract the estimated infections from each single-stream fit and compare them to the truth.

```{r plot_single}
single_inf <- bind_rows(
  fit_cases  |> gather_draws(infections[day]) |> mutate(model = "cases only"),
  fit_deaths |> gather_draws(infections[day]) |> mutate(model = "deaths only")
) |>
  ungroup() |>
  group_by(model) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ungroup()

ggplot(single_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_ts, aes(y = infections), colour = "red") +
  facet_wrap(~model) +
  labs(title = "Infections from single-stream fits (grey) vs truth (red)",
       x = "Day", y = "Infections")
```

::: callout-tip
## Take 2 minutes

Compare the two single-stream fits.
Which stream pins down the *recent* part of the infection curve (the right-hand end) more tightly, and why?
What does the long infection-to-death delay do to the deaths-only fit near the present?
:::

::: {.callout-note collapse="true"}
## Solution

The cases-only fit constrains the recent infections relatively well, because the infection-to-report delay is short, so recent infections still feed into observed cases.
The deaths-only fit is much more uncertain near the present: with a long infection-to-death delay, recent infections have barely begun to show up in deaths, so the most recent infections are informed almost entirely by the prior.
Each stream is individually a valid -- but partial -- view of the same infections, which is exactly why combining them is attractive.
:::

## Part 2: link two streams through the shared infections

Each stream works on its own, so we can now make the **data integration choice** from the workflow: link two of them.
This is where the parallel formulation does its work.
We do *not* chain deaths off cases (a sequential model); instead both cases and deaths look directly at the *same* latent infections, each through its own delay and scaling, and their likelihood terms are added.

```{r fit_two, results = 'hide', message = FALSE}
fit_cd <- fit_streams(use_cases = 1, use_deaths = 1, use_ww = 0)
```

By linking the timely cases with the stable (if delayed) deaths, the two streams together should constrain the infection trajectory better than either alone.

```{r plot_two}
two_inf <- fit_cd |>
  gather_draws(infections[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100))

ggplot(two_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_ts, aes(y = infections), colour = "red") +
  labs(title = "Infections, cases + deaths jointly (grey) vs truth (red)",
       subtitle = "Two streams sharing one latent infection process")
```

## Part 3: add the third stream

Adding a stream is now just another integration step: switch wastewater on as well.

```{r fit_all, results = 'hide', message = FALSE}
fit <- fit_streams(use_cases = 1, use_deaths = 1, use_ww = 1)
```

```{r fit_summary}
fit
```

We extract the posterior for the shared infections and reproduction number from the full three-stream fit.

```{r posteriors}
posteriors <- fit |>
  gather_draws(infections[day], R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100))
```

The infections are now informed by all three streams at once:

```{r plot_inf}
inf_posterior <- posteriors |>
  filter(.variable == "infections")
ggplot(inf_posterior, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_ts, aes(y = infections), colour = "red") +
  labs(title = "Infections, estimated (grey) and true (red)",
       subtitle = "Joint model fit to cases, deaths and wastewater")
```

and so is the reproduction number:

```{r plot_r}
r_posterior <- posteriors |>
  filter(.variable == "R")
ggplot(r_posterior, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(title = "Estimated Rt",
       subtitle = "Joint model fit to all three streams")
```

::: callout-tip
## Take 2 minutes

Look at the scaling parameters (`ascertainment`, `ifr`, `ww_scale`) in the model summary above.
Are they recovered as precisely as $R_t$?
Why might the *absolute level* of infections be harder to pin down than the *trajectory* of $R_t$?
:::

::: {.callout-note collapse="true"}
## Solution

The scaling parameters are recovered much less precisely than $R_t$.
Cases and deaths only constrain the product of infections and their scaling, so multiplying infections by a constant and dividing the scalings by the same constant gives an almost identical fit.
The *shape* of the trajectory -- and hence $R_t$ -- is well identified by all three streams, but the *absolute level* is only weakly identified.
In the workflow this is the "identifiability" feedback arrow: a problem we can either live with or address by fixing or strongly informing one of these scalings (for example, a literature value for the IFR) to anchor the level.
:::

# Part 4: what happens when streams conflict?

Combining streams is powerful when they tell a consistent story.
But real surveillance streams can *disagree*, and it is worth being precise about what that means in this model.
This is the feedback arrow in the workflow that we singled out earlier: *conflict between data sources reveals model misspecification*.

## What does it mean for streams to conflict?

In our parallel model every stream is a delayed, scaled view of the *same* latent infections.
So a stream implicitly carries its own answer to the question "what must the infection trajectory have looked like?".
Two streams **conflict** when those implied trajectories are incompatible: the cases imply infections that are falling while the deaths imply infections that are still rising, say, and no single $R_t$ path can produce both at once.

Why does this happen in practice?
The model assumes each stream is the shared infections convolved with a *fixed* delay and multiplied by a *constant* scaling.
Conflict is what we see when one of those assumptions is wrong for at least one stream, for example:

- **Changing ascertainment.** A change in testing policy or test availability bends the case curve without any change in transmission, so the constant-ascertainment assumption breaks.
- **Mis-specified delays.** If the true infection-to-death delay is longer or more dispersed than the PMF we supplied, deaths will appear shifted relative to the infections the other streams imply.
- **Mis-specified scalings.** A drifting infection fatality ratio (e.g. with changing age structure or treatment) or a reporting artefact pushes a stream off the level the others support.
- **Genuinely different populations.** Streams that sample different groups (e.g. hospital deaths vs community cases) need not share one infection process at all.

To see conflict in action, we re-simulate the deaths stream from a *different* infection trajectory that keeps rising in the second half of the outbreak, while cases and wastewater still come from the original (declining) trajectory.
This is a deliberately stark version of the mis-specified-IFR/changing-severity case above: the deaths now demand a rising epidemic that the other two streams flatly contradict.

```{r simulate_conflict}
conflict_inf <- infections
conflict_inf[40:n] <- conflict_inf[40:n] * seq(1, 4, length.out = n - 39)
exp_deaths_conflict <- ifr * convolve_with_delay(conflict_inf, death_delay_pmf)

obs_conflict <- obs |>
  mutate(deaths = rpois(n, exp_deaths_conflict))
```

## Fitting to conflicting streams

We refit the same joint, three-stream model to this conflicting data, reusing the helper and just swapping in the conflicting deaths.

```{r fit_conflict, results = 'hide', message = FALSE}
data_conflict <- base_data
data_conflict$deaths <- obs_conflict$deaths
fit_conflict <- fit_streams(
  use_cases = 1, use_deaths = 1, use_ww = 1, data = data_conflict
)
```

We compare the $R_t$ estimates from the consistent and conflicting fits.

```{r plot_conflict}
r_consistent <- fit |>
  gather_draws(R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100)) |>
  mutate(data = "consistent")

r_conflict <- fit_conflict |>
  gather_draws(R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100)) |>
  mutate(data = "conflicting deaths")

ggplot(
  rbind(r_consistent, r_conflict),
  aes(x = day, y = .value, group = .draw, colour = data)
) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  labs(title = "Rt under consistent vs conflicting streams") +
  theme(legend.position = "bottom")
```

::: callout-tip
## Take 2 minutes

How does the $R_t$ estimate change when the deaths stream conflicts with the others?
Is the result closer to the "true" declining $R_t$, the rising signal implied by deaths, or somewhere in between?
:::

::: {.callout-note collapse="true"}
## Solution

The conflicting-deaths fit pulls $R_t$ in the second half of the outbreak *upwards*, away from the true declining trajectory, towards a compromise between the declining cases/wastewater and the rising deaths.
A single shared infection process cannot simultaneously be declining (to match cases and wastewater) and rising (to match deaths), so the model splits the difference.
The joint estimate is no longer faithful to any single stream.
:::

## How the joint model surfaces conflict

The $R_t$ comparison above already hints at trouble, but in a real analysis we would not know the truth to compare against.
So how would conflict actually announce itself?
It shows up in three linked ways, and a principled workflow looks for all of them.

**1. Poor fit to one stream.**
The clearest symptom is a *posterior predictive check*: simulate each stream from the fitted model and overlay it on the data.
Because the shared infections are dragged towards a compromise, at least one stream ends up fitted badly.
We can extract the expected counts the model implies for each stream (`exp_cases`, `exp_deaths`) and compare them to the conflicting data.

```{r ppc_conflict}
exp_draws <- fit_conflict |>
  gather_draws(exp_cases[day], exp_deaths[day]) |>
  group_by(.variable, day) |>
  median_qi(.value, .width = 0.9)

obs_long <- obs_conflict |>
  select(day, cases, deaths) |>
  pivot_longer(-day, names_to = "stream", values_to = "value") |>
  mutate(.variable = paste0("exp_", stream))

ggplot(exp_draws, aes(x = day)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.3) +
  geom_line(aes(y = .value)) +
  geom_point(data = obs_long, aes(y = value), size = 0.6) +
  facet_wrap(~.variable, scales = "free_y") +
  labs(title = "Posterior predictive fit under conflicting streams",
       subtitle = "Fitted expectation (line + 90% band) vs observed (points)",
       x = "Day", y = NULL)
```

The compromise infection trajectory cannot match both streams: typically the fitted expectation systematically *under*-shoots the late, rising deaths while *over*-shooting the declining cases (or vice versa), with the data points drifting outside the predictive band on one or both panels.
A consistent run of points on one side of the fitted line for a single stream is the tell-tale sign that that stream is fighting the others.

**2. Tension in the shared latent infections.**
Conflict does not stay neatly inside one stream.
Because all streams pull on the *same* infections, an irreconcilable stream distorts the shared trajectory itself -- and hence $R_t$, as we saw -- and inflates its uncertainty as the sampler is pulled between two incompatible explanations.
A latent quantity that suddenly becomes much more uncertain, or that no longer matches *any* stream when you propagate it back through the observation models, is a second symptom of conflict.

**3. Degraded sampling.**
When the geometry of the posterior is contorted by trying to satisfy contradictory data, the sampler struggles.
In practice this surfaces as divergent transitions, low effective sample sizes, or high `rhat` in the fit summary.
It is always worth scanning the diagnostics of the conflicting fit (`fit_conflict`) and comparing them to the clean fit: sampler complaints are not just a computational nuisance, they can be a signature of a misspecified model.

::: callout-tip
## Take 5 minutes

Look at the posterior predictive plot and at the diagnostics of `fit_conflict` (the `rhat` and `ess_bulk` columns in `fit_conflict`, and any divergence warnings).
Which stream is fitted worst, and in which direction does the model miss?
Do the sampler diagnostics look worse than for the consistent fit?
:::

::: {.callout-note collapse="true"}
## Solution

The deaths panel is fitted worst: the rising deaths in the second half of the outbreak sit *above* the fitted expectation, because the model has been pulled down towards the declining cases and wastewater.
Symmetrically, the late cases tend to sit slightly *below* their fitted line, because the deaths have tugged the shared infections up.
Neither stream is fitted as cleanly as in the consistent case, which is the data-space signature of the $R_t$ compromise.
The diagnostics for the conflicting fit are usually a little worse (larger `rhat`, smaller `ess_bulk`, occasional divergences) because the posterior is harder to explore when the data cannot be reconciled.
:::

## What should a modeller do about conflict?

When streams genuinely conflict, the answer is rarely to average them blindly: the compromise is faithful to no stream and, as we have seen, can be actively misleading about $R_t$.
Conflict is *information*, and in the workflow above it is exactly the feedback arrow *conflict between data sources reveals model misspecification* sending us back to revise the model rather than to trust the fit.
A reasonable sequence is:

- **Diagnose, do not average.** Treat a bad posterior predictive fit, distorted latent infections, or poor sampler diagnostics as evidence that an assumption is wrong, and identify *which* stream and *which* assumption (delay, scaling, or shared-population) is implicated.
- **Relax the offending assumption.** Encode the suspected mechanism in the model: a time-varying ascertainment or IFR (e.g. its own random walk) can absorb a drifting scaling; a more flexible or re-estimated delay can absorb a shifted timing. This is the "add more dynamic components" move in the workflow diagram.
- **Compare models.** Refit with and without the extra flexibility and compare them (e.g. predictive performance, or how much the conflict in the predictive checks is resolved). Conflict that disappears once a mechanism is added tells you something real about the data-generating process.
- **Down-weight or drop as a last resort.** If a stream is simply not measuring the same infections (a different population, or an artefact you cannot model), it may be right to down-weight it or exclude it -- but only after you understand *why*, so the choice is principled rather than convenient.

This loop of building in parts, checking each addition, and letting conflict and identifiability send us back to revise -- exactly the arrows in the workflow diagram -- is the core of a principled modelling workflow [@abbottWorkflowInfectiousDisease2026].

# Going further

## Scope

This session extended the **renewal-with-delays** model to multiple parallel streams.
We deliberately kept it to renewal plus convolutions and did *not* bring in nowcasting or the reporting triangle: each stream here is assumed fully observed up to the present.
Combining this joint, multi-stream approach with the nowcasting models from the [joint nowcasting session](joint-nowcasting) is a natural next step, but it is beyond the scope of this session.

## Challenge

-   We used a Poisson observation model for both cases and deaths. Reporting is often *overdispersed*. How would you change the model to use a negative binomial distribution instead, and what extra parameter would that introduce?
-   We treated `ascertainment`, `ifr` and `ww_scale` as constant. What would happen to the conflict example if you allowed `ascertainment` to vary smoothly over time (for example as its own random walk)? Which stream conflicts would that be able to absorb, and which would it not?
-   In Part 1 we fit each stream on its own and in Parts 2--3 we added them one at a time. Use the `use_*` flags to drop one stream at a time from the full fit. How much does each stream contribute to the precision of the $R_t$ estimate near the end of the outbreak?

## Methods in practice

-   `epidemia` implements joint renewal-equation models of cases and deaths of the kind introduced here [@bhattSemiMechanisticBayesianModeling2020].
-   `EpiSewer` is a Bayesian generative model for estimating transmission dynamics from a wastewater signal, with a carefully specified shedding and measurement model [@lisonRobustRealtimeEstimation2025; @lisonEpiSewer2025].
-   The CDC's wastewater-informed modelling work jointly fits wastewater concentration data and epidemiological count data to produce nowcasts and forecasts [@johnsonWwinference2026]. Unlike our deliberately single-site example, it combines *multiple* wastewater sites with hierarchical structure -- a multi-site, multi-resolution extension of the ideas in this session.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
-   Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
