---
title: "Jointly fitting multiple data streams"
order: 14.7
bibliography: ../nfidd.bib
---

# Introduction

So far we have estimated the reproduction number from a *single* observed time series.
In reality we often observe an outbreak through *several* surveillance streams at once: confirmed cases, deaths, and a wastewater signal, each a delayed, scaled view of the same underlying epidemic.

Rather than reach straight for a model that swallows all of these at once, in this session we follow one *proposed* modelling workflow.
We start from the question we are trying to answer, sketch what a model for it might look like, and then *build that model up in parts*: fitting each data source on its own, linking two together through a shared latent infection process, and finally exposing what happens when streams disagree.
This is the modelling idea behind tools such as `epidemia`, which jointly models cases and deaths from a shared renewal process [@scottEpidemiaRPackage2021].

## Slides

-   [Jointly fitting multiple data streams](slides/multiple-data-streams)

## Objectives

The aim of this session is to develop a joint model of several surveillance streams the way a modeller might: question first, then a sketch of a model for it, then building it up one stream at a time, checking what each step buys us and what happens when streams conflict.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/multiple-data-streams.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to access stan models and helper functions, the `dplyr` and `tidyr` packages for data wrangling, the `ggplot2` library for plotting, the `tidybayes` package for extracting results of the inference, and the `purrr` package for [functional programming](https://adv-r.hadley.nz/fp.html).

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
library("tidybayes")
library("purrr")
```

::: {.callout-tip}
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
We also set an option that makes `cmdstanr` show line numbers when printing model code.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(123)
options(cmdstanr_print_line_numbers = TRUE)
options(cmdstanr_warn_inits = FALSE)
```
:::

# The question we are trying to answer

Before reaching for a model it is worth being explicit about what we want from it.
Our **research question and target estimands** are the same as in earlier sessions: we want to recover the latent infection trajectory and the reproduction number $R_t$ that drives it, in as timely and as well-anchored a way as possible.

No single data stream gives us a perfect view of those quantities: each is a delayed, scaled and noisy window on the same infections, and we look at what each one measures in the next section.
The promise of using them together is that the timely-but-noisy streams sharpen our estimate of the recent trajectory while the stable-but-delayed streams anchor the longer-term level.
The question for this session is therefore not just "can we fit several streams?" but "what does each stream add, and what happens when they pull in different directions?".

# What might an ideal model look like?

Throughout this course we have been building models **iteratively** — each session added a piece and we checked it. That is, informally, the *Bayesian workflow*: the iterative loop of building, checking, and revising a model [@gelman2020bayesian].

Here we make that explicit. Fitting several data streams is one step in a wider, iterative workflow for building, checking and comparing infectious disease models [@abbottWorkflowInfectiousDisease2026] — a recently proposed, infectious-disease-specific instance of that loop. It is one *proposed* workflow rather than the only way to build a model, but it gives us a concrete order to work in, so it is worth looking at before we write any code.

![A workflow for infectious disease modelling. We move from a research question through a process model and per-stream observation models to choices about how to integrate data sources and do inference, and back again -- model checking, identifiability and conflict between data sources all feed revisions. Figure by Abbott et al., reproduced from the `a-workflow-for-infectious-disease-modelling` repository [@abbottWorkflowInfectiousDisease2026] under the MIT licence.](figures/workflow-diagram.png)

There is a lot going on in this diagram, and we will not touch every box in one session.
The point is not to memorise it but to take from it an order of work: according to this approach, a model for our question is assembled from a handful of pieces, roughly as follows.

1.  **Research question and target estimands.** We fixed these above: latent infections and $R_t$.
2.  **Process model.** A single shared epidemic process. We reuse the renewal equation with a geometric random walk on $R_t$ from the [previous session](R-estimation-and-the-renewal-equation), so there is *one* infection trajectory underneath everything.
3.  **Data source selection.** Catalogue the streams available to us (cases, deaths, wastewater) and what each one measures and how it is biased.
4.  **Observation model per source.** For *each* stream, a delay (infection to observation) and a scaling (what fraction of infections, on what scale) and a likelihood. These are separate sub-models, one per data source.
5.  **Data integration choices.** How to combine those per-stream observation models. The diagram flags two things about this step: the *choice* of how to integrate sources, and a feedback arrow warning that *conflict between data sources reveals model misspecification*.
6.  **Inference, checking and revision.** Fit the model, check it against each stream, and let identifiability problems or conflict between sources send us back to revise it.

# Reviewing our data sources

The workflow's *data source selection* step asks us to catalogue the streams we have and, for each, what it measures and how it is biased, before we commit to an observation model for it.
Each stream will be one of the branches in the DAG we design next, and reviewing them now tells us what delay, scaling and likelihood each branch will need.

-   **Confirmed cases** are timely, with a short infection-to-report delay, but only an ascertained *fraction* of infections is ever confirmed, and that fraction depends on testing capacity and policy.
    They pin down the *timing* of recent infections but carry reporting noise.
-   **Deaths** are a more stable, less noisy signal, but arrive with a long infection-to-death delay and capture only a small, severity-dependent fraction of infections (the infection fatality ratio).
    They say little about the present but anchor the longer-term level.
-   **Wastewater** measures shed pathogen concentration, so it does not depend on testing or care-seeking at all and picks up infections regardless of ascertainment.
    It is a continuous signal on its own scale, is relatively indirect and noisy, and needs a scaling before it can be compared to infections.

Each is a delayed, scaled and noisy window on the *same* infections, which is what makes combining them worthwhile: the timely streams sharpen the recent trajectory while the stable ones anchor the level.

# Designing a DAG of the observation process

With the data sources catalogued, the next step is to draw those relationships as a directed acyclic graph (DAG): a picture of how each quantity generates the next, from the shared process at the top down to the observations at the bottom.
For our question the DAG is small.
A geometric random walk on $R_t$ drives a single shared latent infection series $I$ through the renewal equation.
Each stream then takes that *same* $I$, passes it through its own delay and scaling to give an expected signal, and observes that signal through its own likelihood: cases as overdispersed (negative binomial) counts, deaths as Poisson counts, wastewater as a log-normal concentration.

```{=html}
<figure style="margin:1.2em 0;">
<svg viewBox="0 0 880 470" width="100%" style="max-width:760px;display:block;margin:0 auto;font-family:inherit;" role="img" aria-labelledby="dag-title dag-desc">
  <title id="dag-title">Directed acyclic graph for the multi-stream model</title>
  <desc id="dag-desc">A geometric random walk on R(t) drives a single shared latent infection series through the renewal equation. Each stream passes the same infections through its own delay and scaling to an expected signal, which is then observed: cases as overdispersed negative-binomial counts, deaths as Poisson counts, wastewater as a log-normal concentration.</desc>
  <defs>
    <marker id="dag-arrow" viewBox="0 0 10 10" refX="8.5" refY="5" markerWidth="7" markerHeight="7" orient="auto-start-reverse">
      <path d="M0,0 L10,5 L0,10 z" fill="#5b6b7b"/>
    </marker>
  </defs>
  <g stroke="#5b6b7b" stroke-width="2" fill="none" marker-end="url(#dag-arrow)">
    <path d="M440,64 L440,104"/>
    <path d="M330,158 C230,200 160,206 135,252"/>
    <path d="M440,158 L440,252"/>
    <path d="M550,158 C650,200 720,206 745,252"/>
    <path d="M135,306 L135,372"/>
    <path d="M440,306 L440,372"/>
    <path d="M745,306 L745,372"/>
  </g>
  <g font-size="13.5" fill="#374553" text-anchor="middle" paint-order="stroke" stroke="#ffffff" stroke-width="4" stroke-linejoin="round">
    <text x="205" y="214">delay + ascertainment</text>
    <text x="440" y="214">delay + IFR</text>
    <text x="665" y="214">delay + scaling</text>
  </g>
  <g text-anchor="middle">
    <rect x="310" y="14" width="260" height="50" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="440" y="45" font-size="16" fill="#1d2b3a">R<tspan baseline-shift="sub" font-size="11">t</tspan> (geometric random walk)</text>
    <rect x="300" y="104" width="280" height="54" rx="9" fill="#2c7fb8" stroke="#1d5a86" stroke-width="1.5"/>
    <text x="440" y="137" font-size="16.5" font-weight="600" fill="#ffffff">Shared latent infections I</text>
    <rect x="30" y="252" width="210" height="54" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="135" y="284" font-size="15" fill="#1d2b3a">Expected cases</text>
    <rect x="335" y="252" width="210" height="54" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="440" y="284" font-size="15" fill="#1d2b3a">Expected deaths</text>
    <rect x="640" y="252" width="210" height="54" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="745" y="284" font-size="15" fill="#1d2b3a">Expected wastewater</text>
    <rect x="30" y="372" width="210" height="54" rx="9" fill="#ffffff" stroke="#33506e" stroke-width="1.5"/>
    <text x="135" y="404" font-size="15" fill="#1d2b3a">cases ~ NegBinomial</text>
    <rect x="335" y="372" width="210" height="54" rx="9" fill="#ffffff" stroke="#33506e" stroke-width="1.5"/>
    <text x="440" y="404" font-size="15" fill="#1d2b3a">deaths ~ Poisson</text>
    <rect x="640" y="372" width="210" height="54" rx="9" fill="#ffffff" stroke="#33506e" stroke-width="1.5"/>
    <text x="745" y="404" font-size="15" fill="#1d2b3a">ww ~ LogNormal</text>
  </g>
</svg>
<figcaption style="text-align:center;font-size:0.9em;color:#5b6b7b;">A DAG for our model: one shared infection process, observed through a delay, a scaling and a likelihood for each stream.</figcaption>
</figure>
```

## Breaking the DAG into components

The workflow's next step is to break that DAG into modular components, or sub-models.
Ours splits cleanly: the shared process (the $R_t$ random walk and the renewal equation) is one module, and each stream's delay-scaling-likelihood branch is another, self-contained module hanging off the shared infections.
That decomposition is exactly what lets us **build the model in parts** rather than in one leap: we can bring in one branch at a time and check it before adding the next.

::: {.callout-note}
## Top-down vs bottom-up

We are working *top-down* here: sketch the full model we would ideally like, then build it up in parts.
That is not the only order.
A perfectly valid *bottom-up* alternative is to start from the smallest subset of data you actually need to answer the question, design a DAG for just that, and add components as more data (or more need) arrives.
The two meet in the middle; which you prefer often comes down to whether you already know the full picture or are discovering it as you go.
:::

::: {.callout-note .optional collapse="true"}
## Parallel (joint) vs sequential modelling

The "data integration choices" box hides a decision about how the streams relate to each other.

In a **parallel** model -- the approach we use in this session -- every stream is its own convolution *of the shared latent infections*, and the streams are *conditionally independent given those infections*.
Schematically:

$$
\text{cases}_t \sim f\big(\text{convolve}(I, p_\text{cases})\big), \qquad \text{deaths}_t \sim g\big(\text{convolve}(I, p_\text{deaths})\big)
$$

Both cases and deaths look *directly* at the same infections $I$, each through their own delay distribution $p$.
Once we condition on $I$, knowing the cases tells us nothing extra about the deaths: all the shared information flows through the infections.

In a **sequential** model, by contrast, one stream is modelled as a convolution *of another stream* rather than of the infections, for example

$$
\text{deaths}_t \sim g\big(\text{convolve}(\text{cases}, p)\big),
$$

chaining deaths off cases instead of off the shared infections.

**Which should you choose?**
The choice is really about one assumption: is it valid to treat one stream as generated *from* another?
You can diagnose conflict and propagate uncertainty under either formulation, so the decision does not hinge on those alone.

-   A **parallel** model assumes only that each stream is a delayed, scaled view of the same infections, which is usually a safe assumption for cases, deaths and wastewater.
-   A **sequential** model additionally assumes one stream really is drawn from another.
    This is often *not* something you can claim: deaths are not drawn from the *reported* cases (which are themselves an incomplete, delayed observation), nor even from cases plus a delay, because that would ignore the case observation process entirely.

The trade-off is therefore about the validity of that generative assumption rather than about one formulation having uniquely good diagnostics or uncertainty.
A sequential model can be reasonable when the chain truly holds -- for example when a downstream count really is a subset of an upstream one -- but it commits you to trusting the upstream stream as if it were the truth.
We use the **parallel** formulation throughout, which keeps the infection process as the one shared quantity that every stream informs and avoids privileging any single stream.
:::

# Simulating a shared infection trajectory

As in previous sessions we start by simulating the data-generating process, so that we know the truth we are trying to recover.
Rather than invent a new epidemic we reuse the *same* simulated infections the course has used in the [renewal equation](R-estimation-and-the-renewal-equation) and [nowcasting](nowcasting) sessions, and treat them as the shared latent process that every stream observes.

```{r shared_infections}
gen_time_pmf <- make_gen_time_pmf()
inf_ts <- make_daily_infections(infection_times)
```

We take a short window around the epidemic peak so that this session, which fits several models, renders quickly while still spanning a clear rise, peak and decline.

```{r shared_infections_window}
infections <- inf_ts$infections[64:104]
n <- length(infections)
I0 <- infections[1]
inf_df <- tibble(day = seq_len(n), infections = infections)

ggplot(inf_df, aes(x = day, y = infections)) +
  geom_line() +
  labs(title = "Shared latent infections",
       subtitle = "The trajectory every stream observes",
       x = "Day", y = "Infections")
```

We will layer the observation models on top of these shared infections one stream at a time, simulating each stream only at the point we introduce it.

# Building the model in parts

We now build the model up piece by piece, following the build order the workflow suggests: fit each stream on its own, then integrate.

To do this without writing (and compiling) a different model for every combination of streams, we use a *single* Stan model with a switch for each stream.
The model extends the random-walk renewal model from the previous session, and its blocks map directly onto the DAG above: one shared geometric random walk for $R_t$ drives one shared infection process, each stream adds its own convolution, scaling, and likelihood term, and integer flags (`use_cases`, `use_deaths`, `use_ww`) decide which streams contribute to the likelihood.

```{r stan_model}
mod <- nfidd_cmdstan_model("estimate-inf-and-r-multi-stream")
```

Rather than read the whole model at once, we look at it a piece at a time, matching each piece to a component of the DAG.
The full model is on [GitHub](https://github.com/nfidd/sismid-nowcasting/blob/main/inst/stan/estimate-inf-and-r-multi-stream.stan) if you want to see it end to end.

::: {.callout-note}
## Stan and modular models

We use a single Stan file here because it is what the rest of the course uses, but Stan is not an ideal language for building models *modularly* like this.
Stan has no built-in way to package the shared process and each observation branch as separate, reusable components and then compose them: in practice you end up with one monolithic file, as we have.
You *can* get some way there by vendoring Stan functions between projects with tooling such as [`primarycensored`](https://primarycensored.epinowcast.org/), which packages reusable Stan functions (for censored and truncated delays) and copies them into your model at build time.
Other probabilistic programming languages support this modular, composable style more naturally; see the Julia prototype [`EpiAwarePrototype.jl`](https://github.com/EpiAware/EpiAwarePrototype.jl) for one take on building this kind of model from swappable components.
:::

The top of the DAG is the shared process: the $R_t$ random walk and the renewal equation that turns it into one latent infection series.
In the `transformed parameters` block this is:

```stan
// one shared infection / Rt process feeds every stream
array[n] real R = geometric_random_walk(init_R, rw_noise, rw_sd); // <1>
array[gen_time_max] real seed_infections;                         // <2>
for (t in 1:gen_time_max)
  seed_infections[t] = seed_base * exp(initial_growth * (t - gen_time_max));
array[n] real infections =
  renewal_seeded(seed_infections, R, gen_time_pmf);              // <3>
```
1. The geometric random walk on $R_t$ (the top node of the DAG), exactly as in the previous session.
2. A seeded initial history. We start our window mid-outbreak, so there are already infections *before* day one. We build a full `gen_time_max`-day history by exponential growth from an estimated level `seed_base` (near the observed starting level) and an estimated initial growth rate.
3. The renewal equation turns $R_t$ into the shared latent infections $I$. Unlike the plain `renewal(I0, ...)` used elsewhere, which starts from a single value and would collapse to near zero here, `renewal_seeded` starts from that full history, so the estimated infections begin at the right level.

::: {.callout-note collapse="true"}
## Mathematical definition

The shared process is a geometric random walk on $R_t$ feeding the renewal equation:
$$
\log R_t = \log R_{t-1} + \sigma_{rw}\,\epsilon_t, \quad \epsilon_t \sim \mathrm{Normal}(0, 1),
$$
$$
I_t = R_t \sum_{s \ge 1} I_{t-s}\, g(s),
$$
where $g$ is the generation-time PMF and $\sigma_{rw}$ the random-walk standard deviation.
Because we start mid-outbreak, the sum needs infections from *before* day one, so we seed the first `gen_time_max` days with an exponential-growth history $I_t = I_0^{\text{seed}}\,e^{\rho(t - t_{\text{seed}})}$ (estimated level $I_0^{\text{seed}}$ and initial growth rate $\rho$) rather than a single starting value.
:::

Each stream is then a branch hanging off those shared infections: the same $I$, convolved with the stream's own delay and multiplied by its own scaling to give an expected signal.

```stan
array[n] real case_conv =
  convolve_with_delay(infections, case_delay_pmf); // <1>
for (i in 1:n) {
  exp_cases[i] = ascertainment * case_conv[i];     // <2>
  exp_deaths[i] = death_scale[i] * death_conv[i];
  exp_ww[i] = ww_scale * ww_conv[i];
}
```
1. The delay edge in the DAG: convolve the shared infections with this stream's delay PMF (deaths and wastewater have their own `..._conv` lines just above).
2. The scaling: multiply by `ascertainment`, `ifr`/`death_scale` or `ww_scale` to get each stream's expected signal.

::: {.callout-note collapse="true"}
## Mathematical definition

For stream $k$ the expected signal is the shared infections convolved with the stream's delay PMF $p_k$ and multiplied by its scaling $\phi_k$ (ascertainment, IFR/`death_scale`, or wastewater scale):
$$
\mu^{(k)}_t = \phi_k \sum_{d \ge 0} I_{t-d}\, p_k(d).
$$
:::

Finally each expected signal is observed through its own likelihood, and a `use_*` flag decides whether that stream enters the fit.

```stan
if (use_cases) {
  cases ~ neg_binomial_2(                         // <1>
    exp_cases, inv_square(cases_overdispersion)
  );
}
if (use_deaths) {
  deaths ~ poisson(exp_deaths);       // <2>
}
```
1. Cases: overdispersed (negative binomial) counts of the expected cases (the cases observation node). Reporting noise usually exceeds the Poisson, so cases get an extra overdispersion parameter; the reciprocal-overdispersion form near zero recovers the Poisson.
2. Deaths: Poisson counts of the expected deaths. We will meet the wastewater likelihood, which is log-normal, when we add that stream.

::: {.callout-note collapse="true"}
## Mathematical definition

Each stream is observed through its own likelihood on the expected signal $\mu^{(k)}_t$:
$$
C_t \sim \mathrm{NegBinomial}\big(\mu^{(\text{cases})}_t,\ \phi\big), \qquad D_t \sim \mathrm{Poisson}\big(\mu^{(\text{deaths})}_t\big),
$$
$$
\log W_t \sim \mathrm{Normal}\big(\log \mu^{(\text{ww})}_t,\ \sigma_{ww}\big),
$$
for cases $C_t$, deaths $D_t$ and wastewater concentration $W_t$.
:::

Because the streams are conditionally independent given the infections, whichever terms are switched on are simply *added* together to form the joint log-likelihood.

::: {.callout-tip}
## Take 5 minutes

Looking at the pieces above (and comparing with the random-walk model `estimate-inf-and-r-rw` from the [previous session](R-estimation-and-the-renewal-equation#estimating-r_t-with-a-geometric-random-walk-prior)): which parts are shared across all three streams, and which are specific to each stream?
Map each `use_*` flag to a branch of the DAG.
:::

The Stan model always expects data for all three streams; the `use_*` flags decide which ones actually enter the likelihood.
So that we can build up the data the same way we build up the model, we start with a small helper that assembles the data list for whichever streams we have simulated so far, filling any stream we have not yet introduced with harmless placeholders (they never enter the likelihood while its `use_*` flag is `0`).
The helper also carries each stream's scaling prior as a `(mean, sd)` pair.
These default to *relaxed* priors we use for the joint fits; we pass *tight* priors when we fit a single stream on its own, for a reason we come to below.

```{r data_helper}
make_data <- function(cases, deaths, ww,
                      case_delay_pmf, death_delay_pmf, ww_delay_pmf,
                      ascertainment_p = c(0.4, 0.06),
                      ifr_p = c(0.1, 0.02),
                      ww_scale_p = c(2, 0.3),
                      use_nb_cases = 1) {
  list(
    n = length(cases),
    I0 = I0,
    gen_time_max = length(gen_time_pmf),
    gen_time_pmf = gen_time_pmf,
    cases = cases,
    case_delay_max = length(case_delay_pmf) - 1,
    case_delay_pmf = case_delay_pmf,
    deaths = deaths,
    death_delay_max = length(death_delay_pmf) - 1,
    death_delay_pmf = death_delay_pmf,
    ww = ww,
    ww_delay_max = length(ww_delay_pmf) - 1,
    ww_delay_pmf = ww_delay_pmf,
    ascertainment_p = ascertainment_p,
    ifr_p = ifr_p,
    ww_scale_p = ww_scale_p,
    use_nb_cases = use_nb_cases,
    tv_death_scale = 0
  )
}
```

A second helper switches the requested streams on and fits the model.
We provide initial values to help the sampler, and expose `adapt_delta`: the multi-stream *joint* fits have a more difficult posterior, so for those we raise `adapt_delta` to `0.95`, the same lever used in the [joint nowcasting session](joint-nowcasting).
The single-stream diagnostic fits are easier, so they use the course-standard default.
To keep this session quick to run we fit **2 chains** and a shorter warmup (`iter_warmup = 250`, keeping `iter_sampling = 500`); the models still converge well at these settings.

```{r fit_helper}
fit_streams <- function(data, use_cases, use_deaths, use_ww,
                        adapt_delta = 0.8, iter_warmup = 250) {
  data <- c(data, list(
    use_cases = use_cases, use_deaths = use_deaths, use_ww = use_ww
  ))
  nfidd_sample(
    mod, data = data,
    chains = 2, parallel_chains = 2,
    iter_warmup = iter_warmup, iter_sampling = 500,
    adapt_delta = adapt_delta,
    max_treedepth = 12,
    init = \() list(init_R = 1, rw_sd = 0.01)
  )
}
```

## Part 1: simulate and fit cases and deaths on their own

*Workflow stage: observation model per source.*
The first thing the workflow asks of us, once the shared process is fixed, is an observation model *per source*.
We start with the two count streams.
Each is a convolution of the shared infections with its own delay distribution, scaled by a fraction of infections, observed as counts:

-   **Cases**: a short infection-to-report delay, scaled by an *ascertainment* fraction (most infections are never confirmed), and reported with noise beyond the Poisson, so we simulate them as *negative binomial*.
-   **Deaths**: a long infection-to-death delay, scaled by the *infection fatality ratio* (IFR), simulated as Poisson.

```{r simulate_cd}
case_delay_pmf  <- censored_delay_pmf(rgamma, max = 10, shape = 3, rate = 1)
death_delay_pmf <- censored_delay_pmf(rgamma, max = 12, shape = 8, rate = 1)

ascertainment <- 0.4
ifr <- 0.1
case_dispersion <- 5   # negative-binomial size: smaller = noisier cases

exp_cases  <- ascertainment * convolve_with_delay(infections, case_delay_pmf)
exp_deaths <- ifr * convolve_with_delay(infections, death_delay_pmf)

obs <- inf_df |>
  mutate(
    cases  = rnbinom(n, mu = exp_cases, size = case_dispersion),
    deaths = rpois(n, exp_deaths)
  )
```

We can plot the two simulated count streams, the same way we plotted simulated onsets in earlier sessions.
Each is a delayed, rescaled echo of the same infection curve: deaths peak later than cases.
The cases are the timely stream but also the noisiest: drawing them from a negative binomial makes them visibly rougher than the smooth Poisson deaths, which is what we would expect of real case reporting.

```{r plot_cd}
obs |>
  select(day, cases, deaths) |>
  pivot_longer(c(cases, deaths), names_to = "stream", values_to = "value") |>
  ggplot(aes(x = day, y = value)) +
  geom_col() +
  facet_grid(stream ~ ., scales = "free_y") +
  labs(title = "Simulated cases and deaths", x = "Day", y = "Count")
```

### Checking each stream on its own

*Workflow stage: validate each sub-model independently.*
Before combining anything, the workflow tells us to check each sub-model on its own: can each stream, through its own delay and scaling, recover the shared infection process?
This is the modular pay-off from breaking the DAG into components: we validate one branch at a time before we link them.

There is a subtlety here worth being explicit about, because it drives how we fit these single streams.
A single count stream only ever tells us about infections *multiplied by* its scaling: doubling the infections and halving the ascertainment gives exactly the same expected cases.
So a stream identifies the *trend* of infections (and hence $R_t$) but **not their absolute level**.
This is a property of the model rather than of having only one stream: as we will see, adding more streams does not break the invariance, because each new stream arrives with its own free scaling.
To recover the level from one stream we have to inform its scaling.
We therefore give each single-stream diagnostic fit a *tight* prior on that stream's scaling, centred near the truth, as a deliberate device: it lets us check that the stream recovers the infection level, not just its shape.

```{r fit_single, results = 'hide', message = FALSE}
## placeholder wastewater: never enters the likelihood while use_ww = 0
ww_placeholder_pmf <- censored_delay_pmf(rgamma, max = 10, shape = 2, rate = 1)
cd_data <- make_data(
  cases = obs$cases, deaths = obs$deaths, ww = rep(0, n),
  case_delay_pmf = case_delay_pmf,
  death_delay_pmf = death_delay_pmf,
  ww_delay_pmf = ww_placeholder_pmf
)

## tight, informed priors on each stream's own scaling anchor the level
cases_data <- cd_data
cases_data$ascertainment_p <- c(0.4, 0.03)
deaths_data <- cd_data
deaths_data$ifr_p <- c(0.1, 0.01)

fit_cases <- fit_streams(
  cases_data, use_cases = 1, use_deaths = 0, use_ww = 0
)
fit_deaths <- fit_streams(
  deaths_data, use_cases = 0, use_deaths = 1, use_ww = 0
)
```

We extract the estimated infections from each single-stream fit and compare them to the truth.
With the scaling informed, each stream now recovers the infection *level*, not just the trend.

```{r plot_single}
single_inf <- bind_rows(
  fit_cases  |> gather_draws(infections[day]) |> mutate(model = "cases only"),
  fit_deaths |> gather_draws(infections[day]) |> mutate(model = "deaths only")
) |>
  ungroup() |>
  group_by(model) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ungroup()

ggplot(single_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  facet_wrap(~model) +
  labs(title = "Infections from single-stream fits (grey) vs truth (red)",
       x = "Day", y = "Infections")
```

The reproduction number is recovered from each stream too.
Unlike the level, $R_t$ is a property of the *trend*, so each stream identifies it without any help from the scaling prior.
We plot each single-stream $R_t$ estimate; we do not draw a "true" $R_t$ line, because the quantity we are recovering is the latent $R_t$ and we compare estimates to each other rather than to a known truth.

```{r plot_single_r}
single_r <- bind_rows(
  fit_cases  |> gather_draws(R[day]) |> mutate(model = "cases only"),
  fit_deaths |> gather_draws(R[day]) |> mutate(model = "deaths only")
) |>
  ungroup() |>
  group_by(model) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ungroup()

ggplot(single_r, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~model) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt from single-stream fits (sampled trajectories)",
       x = "Day", y = "Rt")
```

::: {.callout-tip}
## Take 2 minutes

Compare the two single-stream fits.
Which stream pins down the *recent* part of the infection curve (the right-hand end) more tightly, and why?
What does the long infection-to-death delay do to the deaths-only fit near the present?
:::

::: {.callout-note collapse="true"}
## Solution

The cases-only fit constrains the recent infections relatively well, because the infection-to-report delay is short, so recent infections still feed into observed cases.
The deaths-only fit is much more uncertain near the present: with a long infection-to-death delay, recent infections have barely begun to show up in deaths, so the most recent infections are informed almost entirely by the prior.
Each stream is a valid but partial view of the same infections, which is why combining them is attractive.
:::

## Part 2: link cases and deaths through the shared infections

*Workflow stage: data integration choice.*
Each stream works on its own, so we can now make the **data integration choice** from the workflow: link the two of them.
Both cases and deaths look directly at the *same* latent infections, each through its own delay and scaling, and their likelihood terms are added.
With two streams tied to one shared infection level, two scaling priors now bear on that level instead of one, so we can drop the tight single-stream priors and fall back to the *relaxed* scaling priors baked into `make_data` rather than pinning any single scaling tightly.

```{r fit_two, results = 'hide', message = FALSE}
fit_cd <- fit_streams(
  cd_data, use_cases = 1, use_deaths = 1, use_ww = 0, adapt_delta = 0.95
)
```

By linking the timely cases with the stable (if delayed) deaths, the two streams together should constrain the infection trajectory better than either alone.

```{r plot_two}
two_inf <- fit_cd |>
  gather_draws(infections[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100))

ggplot(two_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  labs(title = "Infections, cases + deaths jointly (grey) vs truth (red)",
       subtitle = "Two streams sharing one latent infection process")
```

The two-stream fit gives us $R_t$ as well, now informed by both streams at once.

```{r plot_two_r}
two_r <- fit_cd |>
  gather_draws(R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100))

ggplot(two_r, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt, cases + deaths jointly (sampled trajectories)",
       x = "Day", y = "Rt")
```

## Part 3: introduce wastewater and fit it on its own

So far both streams have been counts of people.
We now introduce a third, very different stream: **wastewater**.

Wastewater surveillance measures the concentration of pathogen genetic material (e.g. viral RNA) shed into the sewage system.
Because it samples whatever is in the sewer, it is a *population-level* signal that does not depend on people seeking testing or care, and it picks up infections regardless of whether they are ever ascertained as cases.
That makes it an attractive complement to the count streams: it can lead case reports, captures asymptomatic infection, and is relatively cheap to collect at the scale of a whole catchment.

We model it in a deliberately simple way that mirrors the other streams: the signal is taken to be **proportional to recent infections**, via a short infection-to-shedding delay and a single scaling factor, and observed on the **log scale with log-normal noise** (a single site, with no normalisation or site effects).

This is the third stream's observation node, the one likelihood we deferred earlier.
It is the same delay-scaling-then-observe pattern as the count streams, but with a log-normal likelihood rather than Poisson:

```stan
if (use_ww) {
  ww ~ normal(log(exp_ww), ww_sigma); // <1>
}
```
1. Wastewater: the log concentration is normal around the log of the expected signal, i.e. the signal itself is log-normal.

```{r simulate_ww}
ww_delay_pmf <- censored_delay_pmf(rgamma, max = 10, shape = 2, rate = 1)
ww_scale <- 2

exp_ww <- ww_scale * convolve_with_delay(infections, ww_delay_pmf)
obs <- obs |>
  mutate(ww = rnorm(n, log(exp_ww), 0.2))
```

::: {.callout-note}
## Take 2 minutes

The wastewater stream is modelled on the *log* scale with log-normal noise, whereas cases and deaths are counts with a Poisson model.
Why might a log-normal observation model be a more natural choice for a continuous concentration measurement than a count distribution?
:::

::: {.callout-note collapse="true"}
## Solution

A wastewater concentration is a positive, continuous measurement rather than a count of discrete events, so a count distribution does not apply directly.
Measurement error in such assays tends to be multiplicative (proportional to the signal) rather than additive, which a log-normal observation model captures well.
Dedicated wastewater tools build a much richer shedding and measurement model than ours; see `EpiSewer` [@lisonRobustRealtimeEstimation2025; @lisonEpiSewer2025] and the CDC `ww-inference-model` [@johnsonWwinference2026] to go further.
:::

*Workflow stage: validate each sub-model independently.*
As with the other streams, we first fit wastewater on its own to check it can recover the shared infections through its delay and scaling.
We rebuild the data, now with the real wastewater stream, and give the single-stream fit the same treatment as the counts: a tight prior on its scaling so it recovers the infection level and not only the trend.

```{r fit_ww, results = 'hide', message = FALSE}
full_data <- make_data(
  cases = obs$cases, deaths = obs$deaths, ww = obs$ww,
  case_delay_pmf = case_delay_pmf,
  death_delay_pmf = death_delay_pmf,
  ww_delay_pmf = ww_delay_pmf
)
ww_data <- full_data
ww_data$ww_scale_p <- c(2, 0.15)   # tight, informed prior on the ww scaling
fit_ww <- fit_streams(ww_data, use_cases = 0, use_deaths = 0, use_ww = 1)
```

```{r plot_ww}
ww_inf <- fit_ww |>
  gather_draws(infections[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100))

ggplot(ww_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  labs(title = "Infections, wastewater only (grey) vs truth (red)",
       subtitle = "A short-delay stream on its own scale")
```

## Part 4: add wastewater to the stack

*Workflow stage: data integration choice.*
Adding a stream is now just another integration step: switch wastewater on alongside cases and deaths.
We again use the relaxed scaling priors from `make_data`: with three streams tied to one shared infection level, we no longer need to pin any *single* scaling tightly.

```{r fit_all, results = 'hide', message = FALSE}
fit <- fit_streams(
  full_data, use_cases = 1, use_deaths = 1, use_ww = 1, adapt_delta = 0.95
)
```

```{r fit_summary}
fit
```

We extract the posterior for the shared infections and reproduction number from the full three-stream fit.

```{r posteriors}
posteriors <- fit |>
  gather_draws(infections[day], R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100))
```

The infections are now informed by all three streams at once:

```{r plot_inf}
inf_posterior <- posteriors |>
  filter(.variable == "infections")
ggplot(inf_posterior, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  labs(title = "Infections, estimated (grey) and true (red)",
       subtitle = "Joint model fit to cases, deaths and wastewater")
```

and so is the reproduction number:

```{r plot_r}
r_posterior <- posteriors |>
  filter(.variable == "R")
ggplot(r_posterior, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Estimated Rt (sampled trajectories)",
       subtitle = "Joint model fit to all three streams")
```

## Checking the joint fit against every stream

*Workflow stage: model checking.*
Before we trust the joint fit we should check it reproduces each stream it was fitted to.
The model gives us posterior predictive draws for each stream (`pp_cases`, `pp_deaths`, `pp_ww`): simulated observations *with* each stream's own observation noise, so we can compare them like for like with the data.
We overlay sampled predictive trajectories for the shared infections and all three streams against the simulated data.

```{r ppc_streams}
levels_ppc <- c("infections", "cases", "deaths", "wastewater")
ppc_streams <- bind_rows(
  fit |> gather_draws(infections[day]) |> mutate(stream = "infections"),
  fit |> gather_draws(pp_cases[day])   |> mutate(stream = "cases"),
  fit |> gather_draws(pp_deaths[day])  |> mutate(stream = "deaths"),
  fit |> gather_draws(pp_ww[day])      |> mutate(stream = "wastewater")
) |>
  ungroup() |>
  group_by(stream) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ungroup() |>
  mutate(stream = factor(stream, levels = levels_ppc))

observed_streams <- bind_rows(
  inf_df |> transmute(day, stream = "infections", value = infections),
  obs    |> transmute(day, stream = "cases", value = cases),
  obs    |> transmute(day, stream = "deaths", value = deaths),
  obs    |> transmute(day, stream = "wastewater", value = ww)
) |>
  mutate(stream = factor(stream, levels = levels_ppc))

ggplot(ppc_streams, aes(x = day, y = .value)) +
  geom_line(aes(group = .draw), alpha = 0.1) +
  geom_point(data = observed_streams, aes(y = value),
             colour = "red", size = 0.6) +
  facet_wrap(~stream, scales = "free_y") +
  labs(title = "Posterior predictive check: infections and each stream",
       subtitle = paste("Sampled trajectories with observation error (grey)",
                        "vs simulated data (red)"),
       x = "Day", y = NULL)
```

Each stream's predictive draws should bracket its data (the infections panel shows the latent trajectory against the truth we simulated), which tells us the shared infection process is compatible with all three streams at once.

::: {.callout-tip}
## Take 2 minutes

Look at the scaling parameters (`ascertainment`, `ifr`, `ww_scale`) in the model summary above.
Are they recovered as precisely as $R_t$?
Why might the *absolute level* of infections be harder to pin down than the *trajectory* of $R_t$?
:::

::: {.callout-note collapse="true"}
## Solution

The scaling parameters are recovered less precisely than $R_t$.
Each stream only constrains the product of infections and its scaling, so multiplying the infections by a constant $c$ and dividing *every* scaling by the same $c$ leaves the expected signal of every stream exactly unchanged.
The *shape* of the trajectory -- and hence $R_t$ -- is well identified, but the *absolute level* is not identified by the data at all: it is pinned only by our prior information.
Three things supply that information here: the scaling priors, the prior on the seeded starting level `seed_base`, and the `<lower = 0, upper = 1>` bounds on `ascertainment` and `ifr`, which stop the infections being scaled *down* so far that those scalings would exceed one.
This is why the single-stream fits earlier needed a *tight* prior on their one scaling to recover the level: with relaxed priors, one stream on its own cannot.
Crucially, adding streams does **not** break that invariance, because each new stream arrives with its own free scaling to absorb $c$.
What the joint fit buys is therefore not identification but *more prior information*: three scaling priors now bear on the same infection level instead of one, so it is pinned more tightly than any single stream with the same loose prior could manage.
In the workflow this is the "identifiability" feedback arrow, and it never fully goes away: here every scaling prior is centred on the value we simulated from, and `seed_base ~ normal(I0, I0)` is centred on the true starting infections, so recovering the level is slightly cheating.
Note that the reparameterisation $I \to cI$ is exactly $I_0^{\text{seed}} \to c\,I_0^{\text{seed}}$ (the `seed_base` parameter above), so that seeding prior acts directly on the level and would constrain it even with flat scaling priors.
On real data the level will only ever be as good as the priors you can justify.
:::

# Comparing $R_t$ across streams

*Workflow stage: model checking.*
We have now fitted $R_t$ from cases alone, from deaths alone, and from the joint model.
Putting the three on one figure makes the pay-off of combining streams concrete.
Each single stream identifies $R_t$ (a trend), but with different strengths: cases are timely and sharpen the recent end, deaths are stable but blur it, and the joint fit borrows from both.

```{r combined_r}
r_by_source <- bind_rows(
  fit_cases  |> gather_draws(R[day]) |> mutate(model = "cases only"),
  fit_deaths |> gather_draws(R[day]) |> mutate(model = "deaths only"),
  fit        |> gather_draws(R[day]) |> mutate(model = "joint (all three)")
) |>
  ungroup() |>
  group_by(model) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ungroup() |>
  mutate(model = factor(
    model, levels = c("cases only", "deaths only", "joint (all three)")
  ))

ggplot(r_by_source, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~model) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt by data source: single streams vs the joint fit",
       subtitle = "Sampled trajectories from each fit",
       x = "Day", y = "Rt")
```

# Real-time snapshots at day 20 and day 40

*Workflow stage: model checking (fitness for real-time use).*
Everything above used the *whole* outbreak.
In practice $R_t$ estimation happens *during* an outbreak, when the data only run up to today.
As in the [renewal equation session](R-estimation-and-the-renewal-equation#estimating-r_t-in-real-time) and the [nowcasting session](nowcasting), we can mimic this by truncating the streams at a `cutoff` and refitting, so each fit sees only what would have been available at that point.

We take two snapshots: **day 20**, during the growth phase, and **day 40**, around the peak.
We treat the **day-40** estimate as our *retrospective goal*: with more of the outbreak observed it is the better-anchored estimate, so the question is how close each stream's **real-time day-20** estimate gets to that later, more complete view.
At each cutoff we fit cases only, deaths only, and the joint model, each running only up to its own cutoff.
Unlike the single-stream *level* checks above, these snapshots use the standard (relaxed) scaling priors: $R_t$ is a trend and is identified regardless of the scaling, so we do not need to inform the scale here.
A small helper truncates the data and fits, reusing the same compiled model on the shorter series.

```{r realtime_helper}
fit_at_cutoff <- function(cutoff, use_cases, use_deaths, use_ww, ...) {
  keep <- seq_len(cutoff)
  d <- make_data(
    cases = obs$cases[keep], deaths = obs$deaths[keep], ww = obs$ww[keep],
    case_delay_pmf = case_delay_pmf,
    death_delay_pmf = death_delay_pmf,
    ww_delay_pmf = ww_delay_pmf
  )
  fit_streams(d, use_cases, use_deaths, use_ww, ...)
}
```

```{r realtime_fits, results = 'hide', message = FALSE}
cutoffs <- c(20, 40)
rt_snapshots <- map_dfr(cutoffs, function(ct) {
  fc <- fit_at_cutoff(ct, 1, 0, 0)
  fd <- fit_at_cutoff(ct, 0, 1, 0)
  fj <- fit_at_cutoff(ct, 1, 1, 1, adapt_delta = 0.95)
  bind_rows(
    fc |> gather_draws(R[day]) |> mutate(model = "cases only"),
    fd |> gather_draws(R[day]) |> mutate(model = "deaths only"),
    fj |> gather_draws(R[day]) |> mutate(model = "joint (all three)")
  ) |>
    ungroup() |>
    group_by(model) |>
    filter(.draw %in% sample(.draw, 100)) |>
    ungroup() |>
    mutate(cutoff = ct)
}) |>
  mutate(
    model = factor(
      model, levels = c("cases only", "deaths only", "joint (all three)")
    ),
    cutoff_label = factor(
      if_else(cutoff == 20, "real-time (day 20)", "retrospective (day 40)"),
      levels = c("real-time (day 20)", "retrospective (day 40)")
    )
  )
```

We put both snapshots on one figure, coloured by cutoff and faceted by stream, so that within each panel we can read the real-time (day 20) estimate against the retrospective (day 40) goal.

```{r realtime_plot}
ggplot(rt_snapshots, aes(x = day, y = .value, group = .draw,
                         colour = cutoff_label)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~model) +
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Real-time (day 20) vs retrospective (day 40) Rt",
       subtitle = paste("Each fit uses only data up to its cutoff;",
                        "the day-40 estimate is the retrospective goal"),
       x = "Day", y = "Rt", colour = "estimate") +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
## Take 2 minutes

In each panel, compare the real-time (day 20) trajectories with the retrospective (day 40) ones over their overlapping days.
For which stream does the real-time estimate sit closest to the retrospective goal?
:::

::: {.callout-note collapse="true"}
## Solution

The joint model's real-time (day 20) estimate firms up closest to its retrospective (day 40) goal.
At day 20 the deaths carry almost no signal yet (the long delay means recent infections have not caused deaths), so the deaths-only estimate is barely more than its prior; the cases lead but are noisy, and the joint fit tracks the later retrospective estimate most tightly.
Each single stream is a partial, real-time view; combining them gets us closest, in real time, to the estimate we would only otherwise reach retrospectively.
:::

# When streams conflict

Combining streams works well when they tell a consistent story, but real surveillance streams can *disagree*.
This is the feedback arrow in the workflow that we singled out earlier: *conflict between data sources reveals model misspecification*.

## What does it mean for streams to conflict?

Because every stream is a delayed, scaled view of the *same* latent infections, each implicitly carries its own answer to the question "what must the infection trajectory have looked like?".
Two streams **conflict** when those implied trajectories are incompatible: the cases imply infections that are falling while the deaths imply infections that are still rising, say, and no single $R_t$ path can produce both at once.

Why does this happen in practice?
The model assumes each stream is the shared infections convolved with a *fixed* delay and multiplied by a *constant* scaling.
Conflict is what we see when one of those assumptions is wrong for at least one stream, for example:

- **Changing ascertainment.** A change in testing policy or test availability bends the case curve without any change in transmission, so the constant-ascertainment assumption breaks.
- **Mis-specified delays.** If the true infection-to-death delay is longer or more dispersed than the PMF we supplied, deaths will appear shifted relative to the infections the other streams imply.
- **Mis-specified scalings.** A drifting infection fatality ratio (e.g. with changing age structure or treatment) or a reporting artefact pushes a stream off the level the others support.
- **Truly different populations.** Streams that sample different groups (e.g. hospital deaths vs community cases) need not share one infection process at all.

To see conflict in action we give it a concrete **mechanism**: a *drifting infection fatality ratio*.
We re-simulate the deaths from the *same* shared infections as before, but let the death scaling rise over the second half of the outbreak (as it might if severity increased with age structure or a new variant).
Cases and wastewater are unchanged.
Under our constant-scaling model the rising deaths therefore *look like* a rising epidemic, which the flat-scaling cases and wastewater flatly contradict.

```{r simulate_conflict}
## death scaling drifts up over the second half of the outbreak
ifr_drift <- ifr * c(rep(1, 20), seq(1, 4, length.out = n - 20))
exp_deaths_conflict <-
  ifr_drift * convolve_with_delay(infections, death_delay_pmf)

obs_conflict <- obs |>
  mutate(deaths = rpois(n, exp_deaths_conflict))
```

## Fitting the constant-scaling model to conflicting streams

We refit the same joint, three-stream model -- still assuming a *constant* death scaling -- to this conflicting data, reusing the helpers and just swapping in the conflicting deaths.

```{r fit_conflict, results = 'hide', message = FALSE}
data_conflict <- full_data
data_conflict$deaths <- obs_conflict$deaths
fit_conflict <- fit_streams(
  data_conflict, use_cases = 1, use_deaths = 1, use_ww = 1, adapt_delta = 0.95
)
```

We compare the $R_t$ estimates from the consistent and conflicting fits.

```{r plot_conflict}
r_consistent <- fit |>
  gather_draws(R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100)) |>
  mutate(data = "consistent")

r_conflict <- fit_conflict |>
  gather_draws(R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100)) |>
  mutate(data = "conflicting deaths")

ggplot(
  rbind(r_consistent, r_conflict),
  aes(x = day, y = .value, group = .draw, colour = data)
) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  labs(title = "Rt under consistent vs conflicting streams") +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
## Take 2 minutes

How does the $R_t$ estimate change when the deaths stream conflicts with the others?
Is the result closer to the "true" declining $R_t$, the rising signal implied by deaths, or somewhere in between?
:::

::: {.callout-note collapse="true"}
## Solution

The conflicting-deaths fit pulls $R_t$ in the second half of the outbreak *upwards*, away from the true declining trajectory, towards a compromise between the declining cases/wastewater and the rising deaths.
A single shared infection process cannot simultaneously be declining (to match cases and wastewater) and rising (to match deaths), so the model splits the difference.
The joint estimate is no longer faithful to any single stream.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: when streams conflict

```{=html}
<iframe class="nfidd-embed" src="../interactive/data-conflict.html?embed=conflict" title="when surveillance streams conflict interactive" loading="lazy" style="width:100%;height:640px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Drag the death-scaling drift to open a conflict, then let the scaling vary to snap the estimate back onto the truth.
:::

## How the joint model surfaces conflict

The $R_t$ comparison above already hints at trouble, but in a real analysis we would not know the truth to compare against.
So how would conflict actually announce itself?
It shows up in three linked ways, and a principled workflow looks for all of them.

**1. Poor fit to one stream.**
The clearest symptom is a *posterior predictive check*: simulate each stream from the fitted model, *with* its observation noise, and overlay it on the data.
Because the shared infections are dragged towards a compromise, at least one stream ends up fitted badly.
We draw the posterior predictive counts for each stream (`pp_cases`, `pp_deaths`, from each stream's own likelihood) and compare sampled trajectories to the conflicting data.

```{r ppc_conflict}
pp_conflict <- fit_conflict |>
  gather_draws(pp_cases[day], pp_deaths[day]) |>
  ungroup() |>
  group_by(.variable) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ungroup()

obs_long <- obs_conflict |>
  select(day, cases, deaths) |>
  pivot_longer(-day, names_to = "stream", values_to = "value") |>
  mutate(.variable = paste0("pp_", stream))

ggplot(pp_conflict, aes(x = day, y = .value)) +
  geom_line(aes(group = .draw), alpha = 0.1) +
  geom_point(data = obs_long, aes(y = value), colour = "red", size = 0.6) +
  facet_wrap(~.variable, scales = "free_y") +
  labs(title = "Posterior predictive fit under conflicting streams",
       subtitle = "Sampled predictive trajectories (grey) vs observed (red)",
       x = "Day", y = NULL)
```

The compromise infection trajectory cannot match both streams: typically the predictive draws systematically *under*-shoot the late, rising deaths while *over*-shooting the declining cases (or vice versa), with the data points drifting outside the spread of trajectories on one or both panels.
A consistent run of points on one side of the predictive draws for a single stream is the tell-tale sign that that stream is fighting the others.

**2. Tension in the shared latent infections.**
Conflict does not stay neatly inside one stream.
Because all streams pull on the *same* infections, an irreconcilable stream distorts the shared trajectory itself -- and hence $R_t$, as we saw -- and inflates its uncertainty as the sampler is pulled between two incompatible explanations.
A latent quantity that suddenly becomes much more uncertain, or that no longer matches *any* stream when you propagate it back through the observation models, is a second symptom of conflict.

**3. Degraded sampling.**
When the geometry of the posterior is contorted by trying to satisfy contradictory data, the sampler struggles.
In practice this surfaces as divergent transitions, low effective sample sizes, or high `rhat` in the fit summary.
It is always worth scanning the diagnostics of the conflicting fit (`fit_conflict`) and comparing them to the clean fit: sampler complaints are not just a computational nuisance, they can be a signature of a misspecified model.

::: {.callout-tip}
## Take 5 minutes

Look at the posterior predictive plot and at the diagnostics of `fit_conflict` (the `rhat` and `ess_bulk` columns in `fit_conflict`, and any divergence warnings).
Which stream is fitted worst, and in which direction does the model miss?
Do the sampler diagnostics look worse than for the consistent fit?
:::

::: {.callout-note collapse="true"}
## Solution

The deaths panel is fitted worst: the rising deaths in the second half of the outbreak sit *above* the fitted expectation, because the model has been pulled down towards the declining cases and wastewater.
Symmetrically, the late cases tend to sit slightly *below* their fitted line, because the deaths have tugged the shared infections up.
Neither stream is fitted as cleanly as in the consistent case, which is the data-space signature of the $R_t$ compromise.
The diagnostics for the conflicting fit are usually a little worse (larger `rhat`, smaller `ess_bulk`, occasional divergences) because the posterior is harder to explore when the data cannot be reconciled.
:::

## What should a modeller do about conflict?

It is worth being clear about what the shared latent process does and does not do here.
Sharing one infection process across streams *surfaces* a conflict (the symptoms above) and yields a single compromise estimate, but it does not *resolve* the disagreement: the compromise is faithful to no stream and, as we have seen, can be actively misleading about $R_t$.
Real resolution comes from modelling the *reason* the streams disagree, not from averaging the streams together.

::: {.callout-note}
## Options for resolving conflict

When a principled model check flags conflict between data sources, there are a handful of standard responses [@presanis2013conflict; @deangelis2018analysing; @gelman2020bayesian]:

- **Model the differing mechanism.** Extend the model so the suspected source of divergence is explicit -- a time-varying ascertainment or IFR, a re-estimated or more dispersed delay, or stream-specific reporting effects. This is the principled default when the streams really do observe the same infections [@deangelis2018analysing; @nicholson2022interoperability].
- **Expand or relax the observation model.** Allow overdispersion, or weaken an over-confident likelihood, so a stream is no longer forced onto a level it cannot support.
- **Down-weight or drop a stream.** If a stream measures a truly different population, or carries an artefact you cannot model, it may be right to exclude it -- but only once you understand *why* [@presanis2013conflict].
- **Revisit the data.** Conflict is sometimes a data problem (a definition change, a reporting glitch) rather than a model problem, and is best fixed upstream.

Conflict diagnostics in Bayesian evidence synthesis make this formal: they quantify how much each source disagrees with the rest before deciding what to do [@presanis2013conflict].
:::

We will *do* the first of these: model the mechanism behind the divergence.
We suspected a drifting death scaling, so we relax the constant-IFR assumption and let the death scaling follow its own random walk over time.
The model already has a switch for this, `tv_death_scale`; setting it to `1` turns the scalar `ifr` into a geometric random walk.

```{r fit_resolved, results = 'hide', message = FALSE}
data_resolved <- data_conflict
data_resolved$tv_death_scale <- 1
fit_resolved <- fit_streams(
  data_resolved, use_cases = 1, use_deaths = 1, use_ww = 1
)
```

First, does it now fit the deaths?
We repeat the posterior predictive check, comparing the constant-scaling and time-varying-scaling fits.

```{r ppc_resolved}
ppc_for <- function(fit, label) {
  fit |>
    gather_draws(pp_cases[day], pp_deaths[day]) |>
    ungroup() |>
    group_by(.variable) |>
    filter(.draw %in% sample(.draw, 100)) |>
    ungroup() |>
    mutate(model = label)
}

resolved_draws <- bind_rows(
  ppc_for(fit_conflict, "constant scaling"),
  ppc_for(fit_resolved, "time-varying death scaling")
)

ggplot(resolved_draws, aes(x = day, y = .value)) +
  geom_line(aes(group = .draw), alpha = 0.1) +
  geom_point(data = obs_long, aes(y = value), colour = "red", size = 0.6) +
  facet_grid(.variable ~ model, scales = "free_y") +
  labs(title = "Posterior predictive fit: before and after modelling the drift",
       subtitle = "Sampled predictive trajectories (grey) vs observed (red)",
       x = "Day", y = NULL)
```

Letting the death scaling drift lets the deaths panel track the rising data without dragging the shared infections up, so both streams are fitted well at once.
We can confirm that $R_t$ no longer chases the deaths, by comparing it to the true declining trajectory the cases and wastewater came from.

```{r r_resolved}
r_resolved <- bind_rows(
  fit_conflict |> gather_draws(R[day]) |> mutate(model = "constant scaling"),
  fit_resolved |> gather_draws(R[day]) |>
    mutate(model = "time-varying death scaling")
) |>
  ungroup() |>
  group_by(model) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ungroup()

ggplot(r_resolved, aes(x = day, y = .value, group = .draw, colour = model)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  labs(title = "Rt before and after modelling the drifting death scaling") +
  theme(legend.position = "bottom")
```

```{r recovered_scale}
fit_resolved |>
  gather_draws(death_scale[day]) |>
  median_qi(.value, .width = 0.9) |>
  ggplot(aes(x = day, y = .value)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.3) +
  geom_line() +
  geom_line(
    data = tibble(day = seq_len(n), .value = ifr_drift),
    colour = "red"
  ) +
  labs(title = "Recovered death scaling (black) vs truth (red)",
       x = "Day", y = "Death scaling")
```

::: {.callout-tip}
## Take 5 minutes

Compare the two fits.
Does the time-varying-scaling model fit *both* the cases and the deaths?
Has $R_t$ returned to the true declining trajectory, and are the sampler diagnostics of `fit_resolved` better than `fit_conflict`?
Does the recovered death scaling track the drift we built in?
:::

::: {.callout-note collapse="true"}
## Solution

Modelling the mechanism reconciles the streams.
With a time-varying death scaling the deaths panel is fitted well *and* the cases panel is no longer distorted, because the rising deaths are now explained by a rising scaling rather than by rising infections.
$R_t$ returns to the true declining trajectory, the recovered death scaling tracks the drift we built in, and the sampler diagnostics improve (smaller `rhat`, larger `ess_bulk`, fewer or no divergences) because the posterior no longer has to satisfy contradictory data.
The lesson: you resolve real data conflict by modelling the reason for the divergence, not by hoping the shared latent process will absorb it.
:::

This loop of building in parts, checking each addition, and letting conflict and identifiability send us back to revise is the core of this modelling workflow [@abbottWorkflowInfectiousDisease2026].

Everything here has been on *simulated* data, where we know the truth and can check that the model recovers it.
That is the point of building and testing on simulated data first.
Once the model is tested and checked this way, the same workflow repeats on real data: the same steps of designing and breaking up the DAG, building in parts, and using model checks to send us back to revise, but now without a known truth to compare against, so the posterior predictive checks and conflict diagnostics from this session do the work of telling us whether the model is adequate.

# Going further

## Scope

This session extended the **renewal-with-delays** model to multiple parallel streams.
We kept it to renewal plus convolutions and did *not* bring in nowcasting or the reporting triangle: each stream here is assumed fully observed up to the present.
Combining this multi-stream approach with the nowcasting models from the [joint nowcasting session](joint-nowcasting) is a natural next step.

## Challenge

-   We used a negative binomial for cases (with an overdispersion parameter) but kept deaths Poisson. Following the same pattern, how would you also make the *deaths* overdispersed, and when might that matter?
-   We resolved the conflict by letting the *death* scaling drift. Following the same pattern as `tv_death_scale`, how would you add a time-varying *ascertainment* to absorb a conflict driven by a change in testing policy? Which stream conflicts would that be able to absorb, and which would it not?
-   We added the streams one at a time up to the full fit. Use the `use_*` flags to instead drop one stream at a time from the full fit. How much does each stream contribute to the precision of the $R_t$ estimate near the end of the outbreak?

## Methods in practice

-   `epidemia` implements joint renewal-equation models of cases and deaths of the kind introduced here [@scottEpidemiaRPackage2021].
-   `EpiSewer` is a Bayesian generative model for estimating transmission dynamics from a wastewater signal, with a carefully specified shedding and measurement model [@lisonRobustRealtimeEstimation2025; @lisonEpiSewer2025].
-   The CDC's wastewater-informed modelling work jointly fits wastewater concentration data and epidemiological count data to produce nowcasts and forecasts [@johnsonWwinference2026]. Unlike our deliberately single-site example, it combines *multiple* wastewater sites with hierarchical structure -- a multi-site, multi-resolution extension of the ideas in this session.
-   [`EpiAwarePrototype.jl`](https://github.com/EpiAware/EpiAwarePrototype.jl) is a Julia prototype that builds this kind of multi-stream model from swappable, composable components, an example of tooling designed to make the modular workflow in this session easier than it is in a single Stan file.
-   The workflow paper itself works through schematic case studies (no code) that apply these steps to real modelling problems, a useful set of worked examples of the workflow we followed here [@abbottWorkflowInfectiousDisease2026].

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
-   Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
