---
title: "Jointly fitting multiple data streams"
order: 14.5
bibliography: ../nfidd.bib
---

# Introduction

So far we have estimated the reproduction number from a *single* observed time series: symptom onsets, generated as a convolution of a shared infection process with one delay distribution.
In reality we often observe an outbreak through *several* surveillance streams at once.
We might have confirmed cases, deaths, and a wastewater signal, all generated by the same underlying epidemic but each seen through a different lens: a different delay, a different scaling, and a different observation model.

In this session we take the renewal-with-delays model from the [previous session](R-estimation-and-the-renewal-equation) and extend it so that one shared latent infection process feeds *several parallel observation streams* that we fit *jointly*.
This is the modelling idea behind tools such as `epidemia`, which jointly models cases and deaths from a shared renewal process [@bhattSemiMechanisticBayesianModeling2020].

## Slides

-   [Introduction to the reproduction number](slides/introduction-to-reproduction-number)

## Objectives

The aim of this session is to show how to combine multiple parallel surveillance streams into a single joint model built on the renewal equation, and to understand what happens when those streams *agree* and when they *conflict*.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/multiple-data-streams.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to access stan models and helper functions, the `dplyr` and `tidyr` packages for data wrangling, the `ggplot2` library for plotting, the `tidybayes` package for extracting results of the inference, and the `purrr` package for [functional programming](https://adv-r.hadley.nz/fp.html).

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
library("tidybayes")
library("purrr")
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
We also set an option that makes `cmdstanr` show line numbers when printing model code.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(123)
options(cmdstanr_print_line_numbers = TRUE)
```
:::

# Why combine multiple data streams?

No single data stream gives us a perfect view of an outbreak.
Confirmed cases are timely but depend on testing and reporting, and only a fraction of infections are ever ascertained.
Deaths are a more stable signal but arrive with a long delay and capture only a small, severity-dependent fraction of infections.
A wastewater signal does not depend on people seeking care at all, but it measures something more indirect (shed pathogen concentration) on its own scale.

Each of these is a different, delayed, scaled observation of the *same* underlying infections.
If we model them together, the timely-but-noisy streams can sharpen our estimate of the recent trajectory, while the stable-but-delayed streams anchor the longer-term level.
The reproduction number and infections are then informed by *all* of the data at once.

::: {.callout-note}
## Parallel (joint) vs sequential modelling

There is more than one way to connect several streams to the infection process, and the distinction matters.

In a **parallel** model -- the approach we use in this session -- every stream is its own convolution *of the shared latent infections*, and the streams are *conditionally independent given those infections*.
Schematically:

$$
\text{cases}_t \sim f\big(\text{convolve}(I, p_\text{cases})\big), \qquad \text{deaths}_t \sim g\big(\text{convolve}(I, p_\text{deaths})\big)
$$

Both cases and deaths look *directly* at the same infections $I$, each through their own delay distribution $p$.
Once we condition on $I$, knowing the cases tells us nothing extra about the deaths: all the shared information flows through the infections.

In a **sequential** model, by contrast, one stream is modelled as a convolution *of another stream* rather than of the infections, for example

$$
\text{deaths}_t \sim g\big(\text{convolve}(\text{cases}, p)\big),
$$

chaining deaths off cases instead of off the shared infections.

We use the **parallel / conditionally-independent-given-infections** formulation: a single renewal process generates the latent infections, and each stream is a separate convolution of those infections with its own delay and scaling.
This keeps the infection process as the one shared quantity that every stream informs.
:::

# Simulating parallel data streams from a shared infection trajectory

As in previous sessions we first simulate the data-generating process so that we know the truth we are trying to recover.
We start from a single shared infection trajectory generated by the renewal equation, and then create three parallel streams from it.

We use the same gamma-distributed generation time as in the [renewal equation session](R-estimation-and-the-renewal-equation), discretised with `censored_delay_pmf()` and renormalised so that the generation time is strictly positive.

```{r gen_time}
gen_time_pmf <- censored_delay_pmf(rgamma, max = 14, shape = 4, rate = 1)
gen_time_pmf <- gen_time_pmf[-1] ## remove first element
gen_time_pmf <- gen_time_pmf / sum(gen_time_pmf) ## renormalise
```

We define a reproduction number that starts above 1 and then drops below it, and use the `renewal()` function to generate the shared infection time series.

```{r simulate_infections}
n <- 80
I0 <- 5
R <- c(rep(1.4, 30), rep(0.8, n - 30))
infections <- renewal(I0, R, gen_time_pmf)

inf_ts <- tibble(day = seq_len(n), infections = infections)
ggplot(inf_ts, aes(x = day, y = infections)) +
  geom_line() +
  labs(title = "Shared latent infections", x = "Day", y = "Infections")
```

Now we create three parallel streams.
Each is a convolution of the *same* infections with its own delay distribution and scaling, observed with its own noise model.

-   **Cases**: a short infection-to-report delay, scaled by an *ascertainment* fraction (most infections are never confirmed), observed as counts.
-   **Deaths**: a long infection-to-death delay, scaled by the *infection fatality ratio* (IFR), observed as counts.
-   **Wastewater**: a short infection-to-shedding delay, scaled by a constant, observed on the log scale with log-normal noise.

```{r delays}
case_delay_pmf  <- censored_delay_pmf(rgamma, max = 10, shape = 3,  rate = 1)
death_delay_pmf <- censored_delay_pmf(rgamma, max = 21, shape = 16, rate = 1)
ww_delay_pmf    <- censored_delay_pmf(rgamma, max = 10, shape = 2,  rate = 1)
```

```{r simulate_streams}
ascertainment <- 0.4
ifr <- 0.01
ww_scale <- 2

exp_cases  <- ascertainment * convolve_with_delay(infections, case_delay_pmf)
exp_deaths <- ifr * convolve_with_delay(infections, death_delay_pmf)
exp_ww     <- ww_scale * convolve_with_delay(infections, ww_delay_pmf)

obs <- inf_ts |>
  mutate(
    cases  = rpois(n, exp_cases),
    deaths = rpois(n, exp_deaths),
    ww     = rnorm(n, log(exp_ww), 0.2)
  )
```

We can plot the three streams side by side.
Note how each one is a delayed, rescaled echo of the same infection curve: deaths peak later than cases, and the wastewater signal lives on a completely different scale.

```{r plot_streams}
obs |>
  select(day, cases, deaths, ww) |>
  pivot_longer(-day, names_to = "stream", values_to = "value") |>
  ggplot(aes(x = day, y = value)) +
  geom_line() +
  facet_wrap(~stream, scales = "free_y") +
  labs(title = "Three parallel observation streams", x = "Day", y = NULL)
```

::: callout-note
## Take 2 minutes

The wastewater stream is stored and plotted on the *log* scale (`ww`), whereas cases and deaths are counts.
Why might a log-normal observation model be a more natural choice for a continuous concentration measurement than a count distribution such as the Poisson?
:::

::: {.callout-note collapse="true"}
## Solution

A wastewater concentration is a positive, continuous measurement rather than a count of discrete events, so a count distribution does not apply directly.
Measurement error in such assays tends to be multiplicative (proportional to the signal) rather than additive, which a log-normal observation model captures well.
This is the observation model used by dedicated wastewater tools such as `EpiSewer` [@lisonRobustRealtimeEstimation2025].
:::

# A joint model for multiple data streams

We now fit all three streams jointly with a single Stan model.
The model is a direct extension of the random-walk renewal model from the previous session: one shared geometric random walk for $R_t$ drives one shared infection process, and each stream adds its own convolution, scaling, and likelihood term.

```{r stan_model}
mod <- nfidd_cmdstan_model("estimate-inf-and-r-multi-stream")
mod
```

::: callout-tip
## Take 5 minutes

Familiarise yourself with the model above.
Compare it to the random-walk model `estimate-inf-and-r-rw` from the [previous session](R-estimation-and-the-renewal-equation#estimating-r_t-with-a-geometric-random-walk-prior).
Which parts are shared across all three streams, and which parts are specific to each stream?
Where in the `model` block does each stream contribute to the likelihood?
:::

::: {.callout-note collapse="true"}
## Solution

The `R`, `infections`, generation time, random walk noise and random walk standard deviation are *shared*: there is a single latent infection process.
Each stream has its own delay PMF, its own scaling parameter (`ascertainment`, `ifr`, `ww_scale`) and its own observation term in the `model` block.
The three likelihood lines (`cases ~ poisson(...)`, `deaths ~ poisson(...)`, `ww ~ normal(log(...), ww_sigma)`) are added together: because the streams are conditionally independent given the infections, the joint log-likelihood is simply the sum of the per-stream terms.
:::

We assemble the data for all three streams and fit the model.
As in the previous session we increase `max_treedepth` and provide initial values to help the sampler, because this is a more complex model.

```{r fit, results = 'hide', message = FALSE}
data <- list(
  n = n,
  I0 = I0,
  gen_time_max = length(gen_time_pmf),
  gen_time_pmf = gen_time_pmf,
  cases = obs$cases,
  case_delay_max = length(case_delay_pmf) - 1,
  case_delay_pmf = case_delay_pmf,
  deaths = obs$deaths,
  death_delay_max = length(death_delay_pmf) - 1,
  death_delay_pmf = death_delay_pmf,
  ww = obs$ww,
  ww_delay_max = length(ww_delay_pmf) - 1,
  ww_delay_pmf = ww_delay_pmf
)
fit <- nfidd_sample(
  mod, data = data, max_treedepth = 12,
  init = \() list(init_R = 1, rw_sd = 0.01)
)
```

```{r fit_summary}
fit
```

We extract the posterior for the shared infections and reproduction number.

```{r posteriors}
posteriors <- fit |>
  gather_draws(infections[day], R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100))
```

The infections are informed by all three streams at once:

```{r plot_inf}
inf_posterior <- posteriors |>
  filter(.variable == "infections")
ggplot(inf_posterior, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_ts, aes(y = infections), colour = "red") +
  labs(title = "Infections, estimated (grey) and true (red)",
       subtitle = "Joint model fit to cases, deaths and wastewater")
```

and so is the reproduction number:

```{r plot_r}
r_posterior <- posteriors |>
  filter(.variable == "R")
ggplot(r_posterior, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(title = "Estimated Rt",
       subtitle = "Joint model fit to all three streams")
```

::: callout-tip
## Take 2 minutes

Look at the scaling parameters (`ascertainment`, `ifr`, `ww_scale`) in the model summary above.
Are they recovered as precisely as $R_t$?
Why might the *absolute level* of infections be harder to pin down than the *trajectory* of $R_t$?
:::

::: {.callout-note collapse="true"}
## Solution

The scaling parameters are recovered much less precisely than $R_t$.
Cases and deaths only constrain the product of infections and their scaling, so multiplying infections by a constant and dividing the scalings by the same constant gives an almost identical fit.
The *shape* of the trajectory -- and hence $R_t$ -- is well identified by all three streams, but the *absolute level* is only weakly identified.
In practice we often fix or strongly inform one of these scalings (for example, a literature value for the IFR) to anchor the level.
:::

# What happens when streams conflict?

Combining streams is powerful when they tell a consistent story.
But real surveillance streams can *disagree*: a change in testing policy can bend the case curve without any change in transmission, or a reporting artefact can distort deaths.
When we force a single shared infection process to explain conflicting signals, the model reaches a *compromise* -- and that compromise may not match any individual stream.

To see this, we re-simulate the deaths stream from a *different* infection trajectory that keeps rising in the second half of the outbreak, while cases and wastewater still come from the original (declining) trajectory.

```{r simulate_conflict}
conflict_inf <- infections
conflict_inf[40:n] <- conflict_inf[40:n] * seq(1, 4, length.out = n - 39)
exp_deaths_conflict <- ifr * convolve_with_delay(conflict_inf, death_delay_pmf)

obs_conflict <- obs |>
  mutate(deaths = rpois(n, exp_deaths_conflict))
```

We refit the same joint model to this conflicting data.

```{r fit_conflict, results = 'hide', message = FALSE}
data_conflict <- data
data_conflict$deaths <- obs_conflict$deaths
fit_conflict <- nfidd_sample(
  mod, data = data_conflict, max_treedepth = 12,
  init = \() list(init_R = 1, rw_sd = 0.01)
)
```

We compare the $R_t$ estimates from the consistent and conflicting fits.

```{r plot_conflict}
r_consistent <- fit |>
  gather_draws(R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100)) |>
  mutate(data = "consistent")

r_conflict <- fit_conflict |>
  gather_draws(R[day]) |>
  ungroup() |>
  filter(.draw %in% sample(.draw, 100)) |>
  mutate(data = "conflicting deaths")

ggplot(
  rbind(r_consistent, r_conflict),
  aes(x = day, y = .value, group = .draw, colour = data)
) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  labs(title = "Rt under consistent vs conflicting streams") +
  theme(legend.position = "bottom")
```

::: callout-tip
## Take 2 minutes

How does the $R_t$ estimate change when the deaths stream conflicts with the others?
Is the result closer to the "true" declining $R_t$, the rising signal implied by deaths, or somewhere in between?
:::

::: {.callout-note collapse="true"}
## Solution

The conflicting-deaths fit pulls $R_t$ in the second half of the outbreak *upwards*, away from the true declining trajectory, towards a compromise between the declining cases/wastewater and the rising deaths.
A single shared infection process cannot simultaneously be declining (to match cases and wastewater) and rising (to match deaths), so the model splits the difference.
The joint estimate is no longer faithful to any single stream.
:::

When streams genuinely conflict, the answer is rarely to average them blindly.
Instead we need to reason about *why* they disagree -- a change in ascertainment, a reporting artefact, a stream-specific delay that has shifted -- and encode that understanding in the model (for example by allowing a time-varying ascertainment, or by down-weighting a stream we trust less).
Thinking carefully about data sources, their biases, and how to reconcile them is a core part of a principled modelling workflow [@abbottWorkflowInfectiousDisease2026].

# Going further

## Scope

This session extended the **renewal-with-delays** model to multiple parallel streams.
We deliberately kept it to renewal plus convolutions and did *not* bring in nowcasting or the reporting triangle: each stream here is assumed fully observed up to the present.
Combining this joint, multi-stream approach with the nowcasting models from the [joint nowcasting session](joint-nowcasting) is a natural next step, but it is beyond the scope of this session.

## Challenge

-   We used a Poisson observation model for both cases and deaths. Reporting is often *overdispersed*. How would you change the model to use a negative binomial distribution instead, and what extra parameter would that introduce?
-   We treated `ascertainment`, `ifr` and `ww_scale` as constant. What would happen to the conflict example if you allowed `ascertainment` to vary smoothly over time (for example as its own random walk)? Which stream conflicts would that be able to absorb, and which would it not?
-   Try dropping one stream at a time from the data and refitting. How much does each stream contribute to the precision of the $R_t$ estimate near the end of the outbreak?

## Methods in practice

-   `epidemia` implements joint renewal-equation models of cases and deaths of the kind introduced here [@bhattSemiMechanisticBayesianModeling2020].
-   `EpiSewer` is a Bayesian generative model for estimating transmission dynamics from a wastewater signal, with a carefully specified shedding and measurement model [@lisonRobustRealtimeEstimation2025; @lisonEpiSewer2025].
-   The CDC's wastewater-informed modelling work jointly fits wastewater concentration data and epidemiological count data to produce nowcasts and forecasts [@johnsonWwinference2026]. Unlike our deliberately single-site example, it combines *multiple* wastewater sites with hierarchical structure -- a multi-site, multi-resolution extension of the ideas in this session.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
-   Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
