---
title: "Jointly fitting multiple data sources"
order: 8
bibliography: ../nfidd.bib
---

# Introduction

So far we have estimated the reproduction number from a *single* observed time series.
In reality we often observe an outbreak through *several* surveillance data sources at once: confirmed cases, deaths, and a wastewater signal, each a delayed, scaled view of the same underlying epidemic.

Rather than reach straight for a model that swallows all of these at once, in this session we follow one *proposed* modelling workflow.
We start from the question we are trying to answer, sketch what a model for it might look like, and then *build that model up in parts*: fitting each data source on its own, linking two together through a shared latent infection process, and finally exposing what happens when data sources disagree.
This is the modelling idea behind tools such as `epidemia`, which jointly models cases and deaths from a shared renewal process [@scottEpidemiaRPackage2021].

## Slides

-   [Jointly fitting multiple data sources](slides/multiple-data-sources)

## Objectives

The aim of this session is to develop a joint model of several surveillance data sources the way a modeller might: question first, then a sketch of a model for it, then building it up one data source at a time, checking what each step buys us and what happens when data sources conflict.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/multiple-data-sources.qmd`.

## Libraries used

In this session we will use the `nfidd.nowcasting` package to access stan models and helper functions, the `dplyr` and `tidyr` packages for data wrangling, the `ggplot2` library for plotting, the `tidybayes` package for extracting results of the inference, and the `purrr` package for [functional programming](https://adv-r.hadley.nz/fp.html).

```{r libraries, message = FALSE}
library("nfidd.nowcasting")
library("dplyr")
library("tidyr")
library("ggplot2")
library("tidybayes")
library("purrr")
```

::: {.callout-tip}
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
We also set an option that makes `cmdstanr` show line numbers when printing model code.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(123)
options(cmdstanr_print_line_numbers = TRUE)
options(cmdstanr_warn_inits = FALSE)
```
:::

# The question we are trying to answer

Before reaching for a model we need to be explicit about what we want from it.
Our **research question and target estimands** are the same as in earlier sessions: we want to recover the latent infection trajectory and the reproduction number $R_t$ that drives it, as quickly and with as little bias as we can.

No single data source gives us a perfect view of those quantities: each is a delayed, scaled and noisy window on the same infections, and we look at what each one measures in the next section.
The promise of using them together is that the timely-but-noisy data sources sharpen our estimate of the recent trajectory while the stable-but-delayed data sources anchor the longer-term level.
The question for this session is therefore not just "can we fit several data sources?" but "what does each data source add, and what happens when they pull in different directions?".

# A workflow for building the model

Throughout this course we have been building models **iteratively** — each session added a piece and we checked it. That is, informally, the *Bayesian workflow*: the iterative loop of building, checking, and revising a model [@gelman2020bayesian].

Here we make that explicit. Fitting several data sources is one step in a wider, iterative workflow for building, checking and comparing infectious disease models [@abbottWorkflowInfectiousDisease2026] — a recently proposed, infectious-disease-specific instance of that loop. It is one *proposed* workflow rather than the only way to build a model, but it gives us a concrete order to work in, so it is worth looking at before we write any code.

![A workflow for infectious disease modelling. We move from a research question through a process model and per-source observation models to choices about how to integrate data sources and do inference, and back again -- model checking, identifiability and conflict between data sources all feed revisions. Figure by Abbott et al., reproduced from the `a-workflow-for-infectious-disease-modelling` repository [@abbottWorkflowInfectiousDisease2026] under the MIT licence.](figures/workflow-diagram.png)

There is a lot going on in this diagram, and we will not touch every box in one session.
The point is not to memorise it but to take from it an order of work: according to this approach, a model for our question is assembled from a handful of pieces, roughly as follows.

1.  **Research question and target estimands.** We fixed these above: latent infections and $R_t$.
2.  **Process model.** A single shared epidemic process, a geometric random walk on $R_t$ feeding the renewal equation, so there is *one* infection trajectory underneath everything (next section).
3.  **Data source selection.** Catalogue the data sources available to us (cases, deaths, wastewater) and what each one measures and how it is biased.
4.  **Observation model per source.** For *each* data source, a delay (infection to observation), a scaling (what fraction of infections, on what scale) and a likelihood. These are separate sub-models, one per data source.
5.  **Modularisation.** Break the model into those sub-models so each can be built and checked on its own.
6.  **Inference and computation choices.** How we fit the model (here, Hamiltonian Monte Carlo via Stan).
7.  **Implementation.** Write the model, mapping each DAG component to code.
8.  **Model validation.** Check each piece against data, fitting each data source on its own to see whether it recovers the shared process before we combine anything.
9.  **Data integration.** Only once the pieces work do we choose how to combine the per-source observation models. The diagram flags two things here: the *choice* of how to integrate sources, and a feedback arrow warning that *conflict between data sources reveals model misspecification*.

We work through these in the session, revising earlier steps as later ones send us back.

# The process model

The first modelling step, once the question is fixed, is the *process model*, the latent epidemic we want to recover before any data enter.
We reuse the process from the [renewal equation session](R-estimation-and-the-renewal-equation).
A geometric random walk on the reproduction number $R_t$ drives a single shared latent infection trajectory through the renewal equation, so one $R_t$ path and one infection series sit underneath everything that follows.

Fixing this shared process first is what lets several data sources inform the *same* infections.
Every data source we add later is a delayed, scaled view of this one trajectory, rather than a separate epidemic of its own.

::: {.callout-note collapse="true"}
## Mathematical definition

The shared process is a geometric random walk on $R_t$ feeding the renewal equation:
$$
\log R_t = \log R_{t-1} + \sigma_{rw}\,\epsilon_t, \quad \epsilon_t \sim \mathrm{Normal}(0, 1),
$$
$$
I_t = R_t \sum_{s \ge 1} I_{t-s}\, g(s),
$$
where $g$ is the generation-time PMF and $\sigma_{rw}$ the random-walk standard deviation.
Because we start mid-outbreak, the sum needs infections from *before* day one, so we seed the first `gen_time_max` days with an exponential-growth history $I_t = I_0^{\text{seed}}\,e^{\rho(t - t_{\text{seed}})}$ (estimated level $I_0^{\text{seed}}$ and initial growth rate $\rho$) rather than a single starting value.
:::

# Reviewing our data sources

The workflow's *data source selection* step asks us to catalogue the data sources we have and, for each, what it measures and how it is biased, before we commit to an observation model for it.
Each data source will be one of the branches in the DAG we design next, and reviewing them now tells us what delay, scaling and likelihood each branch will need.

-   **Confirmed cases** are timely, with a short infection-to-report delay, but only an ascertained *fraction* of infections is ever confirmed, and that fraction depends on testing capacity and policy.
    They pin down the *timing* of recent infections but carry reporting noise.
-   **Deaths** are a more stable, less noisy signal, but arrive with a long infection-to-death delay and capture only a small, severity-dependent fraction of infections (the infection fatality ratio).
    They say little about the present but anchor the longer-term level.
-   **Wastewater** measures shed pathogen concentration, so it does not depend on testing or care-seeking at all and picks up infections regardless of ascertainment.
    It is a continuous signal on its own scale, is relatively indirect and noisy, and needs a scaling before it can be compared to infections.

Each is a delayed, scaled and noisy window on the *same* infections, which is what makes combining them worthwhile: the timely data sources sharpen the recent trajectory while the stable ones anchor the level.

::: {.callout-tip}
## Take a few minutes

If you could only bring in these data sources *one at a time*, in what order would you investigate them, and why?
:::

::: {.callout-note collapse="true"}
## Solution

There is no single right answer: it depends on timeliness, how easily you can get the data, and what stakeholders will act on.

- **Deaths** are the most stable and least noisy, so they are often the safest first choice.
- **Cases** are the more traditional starting point and are timely, though noisier and dependent on testing.
- **Wastewater** is attractive but is not grounded to anything directly actionable for public health, so it often comes last.
:::

# Designing a DAG of the observation process

With the data sources catalogued, the next step is to draw those relationships as a directed acyclic graph (DAG): a picture of how each quantity generates the next, from the shared process at the top down to the observations at the bottom.
For our question the DAG is small.
A geometric random walk on $R_t$ drives a single shared latent infection series $I$ through the renewal equation.
Each data source then takes that *same* $I$, passes it through its own delay and scaling to give an expected signal, and observes that signal through its own likelihood: cases as overdispersed (negative binomial) counts, deaths as Poisson counts, wastewater as a log-normal concentration.

::: {.callout-note}
## What is a DAG?

A directed acyclic graph draws a model as a picture.
Each quantity, observed or latent, is a *node*, and a directed *arrow* runs from each quantity to the one it helps generate.
Reading the arrows off the graph shows how the data are generated and makes the model's conditional independence assumptions explicit.
A model DAG is, in effect, the TLDR of the model.
See @abbottWorkflowInfectiousDisease2026 for how DAGs structure process and observation models in a state-space framework.
:::

```{=html}
<figure style="margin:1.2em 0;">
<svg viewBox="0 0 880 470" width="100%" style="max-width:760px;display:block;margin:0 auto;font-family:inherit;" role="img" aria-labelledby="dag-title dag-desc">
  <title id="dag-title">Directed acyclic graph for the multi-source model</title>
  <desc id="dag-desc">A geometric random walk on R(t) drives a single shared latent infection series through the renewal equation. Each data source passes the same infections through its own delay and scaling to an expected signal, which is then observed: cases as overdispersed negative-binomial counts, deaths as Poisson counts, wastewater as a log-normal concentration.</desc>
  <defs>
    <marker id="dag-arrow" viewBox="0 0 10 10" refX="8.5" refY="5" markerWidth="7" markerHeight="7" orient="auto-start-reverse">
      <path d="M0,0 L10,5 L0,10 z" fill="#5b6b7b"/>
    </marker>
  </defs>
  <g stroke="#5b6b7b" stroke-width="2" fill="none" marker-end="url(#dag-arrow)">
    <path d="M440,64 L440,104"/>
    <path d="M330,158 C230,200 160,206 135,252"/>
    <path d="M440,158 L440,252"/>
    <path d="M550,158 C650,200 720,206 745,252"/>
    <path d="M135,306 L135,372"/>
    <path d="M440,306 L440,372"/>
    <path d="M745,306 L745,372"/>
  </g>
  <g font-size="13.5" fill="#374553" text-anchor="middle" paint-order="stroke" stroke="#ffffff" stroke-width="4" stroke-linejoin="round">
    <text x="205" y="214">delay + ascertainment</text>
    <text x="440" y="214">delay + IFR</text>
    <text x="665" y="214">delay + scaling</text>
  </g>
  <g text-anchor="middle">
    <rect x="310" y="14" width="260" height="50" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="440" y="45" font-size="16" fill="#1d2b3a">R<tspan baseline-shift="sub" font-size="11">t</tspan> (geometric random walk)</text>
    <rect x="300" y="104" width="280" height="54" rx="9" fill="#2c7fb8" stroke="#1d5a86" stroke-width="1.5"/>
    <text x="440" y="137" font-size="16.5" font-weight="600" fill="#ffffff">Shared latent infections I</text>
    <rect x="30" y="252" width="210" height="54" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="135" y="284" font-size="15" fill="#1d2b3a">Expected cases</text>
    <rect x="335" y="252" width="210" height="54" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="440" y="284" font-size="15" fill="#1d2b3a">Expected deaths</text>
    <rect x="640" y="252" width="210" height="54" rx="9" fill="#eef2f7" stroke="#33506e" stroke-width="1.5"/>
    <text x="745" y="284" font-size="15" fill="#1d2b3a">Expected wastewater</text>
    <rect x="30" y="372" width="210" height="54" rx="9" fill="#ffffff" stroke="#33506e" stroke-width="1.5"/>
    <text x="135" y="404" font-size="15" fill="#1d2b3a">cases ~ NegBinomial</text>
    <rect x="335" y="372" width="210" height="54" rx="9" fill="#ffffff" stroke="#33506e" stroke-width="1.5"/>
    <text x="440" y="404" font-size="15" fill="#1d2b3a">deaths ~ Poisson</text>
    <rect x="640" y="372" width="210" height="54" rx="9" fill="#ffffff" stroke="#33506e" stroke-width="1.5"/>
    <text x="745" y="404" font-size="15" fill="#1d2b3a">ww ~ LogNormal</text>
  </g>
</svg>
<figcaption style="text-align:center;font-size:0.9em;color:#5b6b7b;">A DAG for our model: one shared infection process, observed through a delay, a scaling and a likelihood for each data source.</figcaption>
</figure>
```

::: {.callout-note .optional collapse="true"}
## Parallel (joint) vs sequential modelling

The DAG above is *one* choice of how the data sources relate to each other; it is worth being explicit about the alternative.

In a **parallel** model -- the approach we use in this session -- every data source is its own convolution *of the shared latent infections*, and the data sources are *conditionally independent given those infections*.
Schematically:

$$
\text{cases}_t \sim f\big(\text{convolve}(I, p_\text{cases})\big), \qquad \text{deaths}_t \sim g\big(\text{convolve}(I, p_\text{deaths})\big)
$$

Both cases and deaths look *directly* at the same infections $I$, each through their own delay distribution $p$.
Once we condition on $I$, knowing the cases tells us nothing extra about the deaths: all the shared information flows through the infections.

In a **sequential** model, by contrast, one data source is modelled as a convolution *of another data source* rather than of the infections, for example

$$
\text{deaths}_t \sim g\big(\text{convolve}(\text{cases}, p)\big),
$$

chaining deaths off cases instead of off the shared infections.

**Which should you choose?**
The question is whether it is valid to treat one data source as generated *from* another.

-   A **parallel** model assumes only that each data source is a delayed, scaled view of the same infections, which is usually a safe assumption for cases, deaths and wastewater.
-   A **sequential** model additionally assumes one data source is generated *from* another.

These are two ends of a spectrum rather than a hard choice.
The naive extreme is chaining a data source off *reported* cases, which is usually hard to justify because reported cases are themselves an incomplete, delayed observation rather than the truth.
But there is useful ground in between.
You can chain one signal off a *latent, de-noised* version of another -- for example modelling deaths as arising from the estimated infections that the cases inform, rather than from raw case counts -- or pass the posterior from one model as an informative prior into the next (the pipeline approach).
These keep some sequential structure without pretending that a reported data source is the truth.

The trade-off is really about the validity of that generative assumption rather than about one formulation having uniquely good diagnostics or uncertainty; you can diagnose conflict and propagate uncertainty under any of them.
We use the **parallel** formulation throughout, which keeps the infection process as the one shared quantity that every data source informs and avoids privileging any single data source, but a sequential or intermediate linkage can be the right choice when the chain genuinely holds, for example when a downstream count really is a subset of an upstream one.
:::

## Breaking the DAG into components

The workflow's next step is to break that DAG into modular components, or sub-models.
Ours splits cleanly: the shared process (the $R_t$ random walk and the renewal equation) is one module, and each data source's delay-scaling-likelihood branch is another, self-contained module hanging off the shared infections.
That decomposition is exactly what lets us **build the model in parts** rather than in one leap: we can bring in one branch at a time and check it before adding the next.

::: {.callout-note}
## Top-down vs bottom-up

We are working *top-down* here: sketch the full model we would ideally like, then build it up in parts.
That is not the only order.
A perfectly valid *bottom-up* alternative is to start from the smallest subset of data you actually need to answer the question, design a DAG for just that, and add components as more data (or more need) arrives.
The two meet in the middle; which you prefer often comes down to whether you already know the full picture or are discovering it as you go.
:::

# Simulating a shared infection trajectory

As in previous sessions we start by simulating the data-generating process, so that we know the truth we are trying to recover.
Rather than invent a new epidemic we reuse the *same* simulated infections the course has used in the [renewal equation](R-estimation-and-the-renewal-equation) and [nowcasting](nowcasting) sessions, and treat them as the shared latent process that every data source observes.

```{r shared_infections}
gen_time_pmf <- make_gen_time_pmf()
inf_ts <- make_daily_infections(infection_times)
```

We take a short window around the epidemic peak so that this session, which fits several models, renders quickly while still spanning a clear rise, peak and decline.

```{r shared_infections_window}
infections <- inf_ts$infections[64:104]
n <- length(infections)
I0 <- infections[1]
inf_df <- tibble(day = seq_len(n), infections = infections)

ggplot(inf_df, aes(x = day, y = infections)) +
  geom_line() +
  labs(title = "Shared latent infections",
       subtitle = "The trajectory every data source observes",
       x = "Day", y = "Infections")
```

We will layer the observation models on top of these shared infections one data source at a time, simulating each data source only at the point we introduce it.

# Building the model in parts

We now build the model up piece by piece, following the build order the workflow suggests: fit each data source on its own, then integrate.

To do this without writing (and compiling) a different model for every combination of data sources, we use a *single* Stan model with a switch for each data source.
The model extends the random-walk renewal model from the previous session, and its blocks map directly onto the DAG above: one shared geometric random walk for $R_t$ drives one shared infection process, each data source adds its own convolution, scaling, and likelihood term, and integer flags (`use_cases`, `use_deaths`, `use_ww`) decide which data sources contribute to the likelihood.

```{r stan_model}
mod <- nfidd_cmdstan_model("estimate-inf-and-r-multi-stream")
```

Rather than read the whole model at once, we look at it a piece at a time, matching each piece to a component of the DAG.
The full model is on [GitHub](https://github.com/nfidd/sismid-nowcasting/blob/main/inst/stan/estimate-inf-and-r-multi-stream.stan) if you want to see it end to end.

::: {.callout-note}
## Stan and modular models

We use a single Stan file here because it is what the rest of the course uses, but Stan is not an ideal language for building models *modularly* like this.
Stan has no built-in way to package the shared process and each observation branch as separate, reusable components and then compose them: in practice you end up with one monolithic file, as we have.
You *can* get some way there by vendoring Stan functions between projects with tooling such as [`primarycensored`](https://primarycensored.epinowcast.org/), which packages reusable Stan functions (for censored and truncated delays) and copies them into your model at build time.
Other probabilistic programming languages support this modular, composable style more naturally; see the Julia prototype [`ComposableTuringIDModels.jl`](https://github.com/EpiAware/ComposableTuringIDModels.jl) for one take on building this kind of model from swappable components.
:::

The top of the DAG is the shared process: the $R_t$ random walk and the renewal equation that turns it into one latent infection series.
In the `transformed parameters` block this is:

```stan
// one shared infection / Rt process feeds every data source
array[n] real R = geometric_random_walk(init_R, rw_noise, rw_sd); // <1>
array[gen_time_max] real seed_infections;                         // <2>
for (t in 1:gen_time_max)
  seed_infections[t] = seed_base * exp(initial_growth * (t - gen_time_max));
array[n] real infections =
  renewal_seeded(seed_infections, R, gen_time_pmf);              // <3>
```
1. The geometric random walk on $R_t$ (the top node of the DAG), exactly as in the previous session.
2. A seeded initial history. We start our window mid-outbreak, so there are already infections *before* day one. We build a full `gen_time_max`-day history by exponential growth from an estimated level `seed_base` (near the observed starting level) and an estimated initial growth rate.
3. The renewal equation turns $R_t$ into the shared latent infections $I$. Unlike the plain `renewal(I0, ...)` used elsewhere, which starts from a single value and would collapse to near zero here, `renewal_seeded` starts from that full history, so the estimated infections begin at the right level.

Each data source is then a branch hanging off those shared infections: the same $I$, convolved with the data source's own delay and multiplied by its own scaling to give an expected signal.

```stan
array[n] real case_conv =
  convolve_with_delay(infections, case_delay_pmf); // <1>
for (i in 1:n) {
  exp_cases[i] = ascertainment * case_conv[i];     // <2>
  exp_deaths[i] = death_scale[i] * death_conv[i];
  exp_ww[i] = ww_scale * ww_conv[i];
}
```
1. The delay edge in the DAG: convolve the shared infections with this data source's delay PMF (deaths and wastewater have their own `..._conv` lines just above).
2. The scaling: multiply by `ascertainment`, `ifr`/`death_scale` or `ww_scale` to get each data source's expected signal.

::: {.callout-note collapse="true"}
## Mathematical definition

For data source $k$ the expected signal is the shared infections convolved with the data source's delay PMF $p_k$ and multiplied by its scaling $\phi_k$ (ascertainment, IFR/`death_scale`, or wastewater scale):
$$
\mu^{(k)}_t = \phi_k \sum_{d \ge 0} I_{t-d}\, p_k(d).
$$
:::

Finally each expected signal is observed through its own likelihood, and a `use_*` flag decides whether that data source enters the fit.

```stan
if (use_cases) {
  cases ~ neg_binomial_2(                         // <1>
    exp_cases, inv_square(cases_overdispersion)
  );
}
if (use_deaths) {
  deaths ~ poisson(exp_deaths);       // <2>
}
```
1. Cases: overdispersed (negative binomial) counts of the expected cases (the cases observation node). Reporting noise usually exceeds the Poisson, so cases get an extra overdispersion parameter; the reciprocal-overdispersion form near zero recovers the Poisson.
2. Deaths: Poisson counts of the expected deaths. Wastewater is a log-normal likelihood on the log concentration (in the maths below); we switch it on in Part 3.

::: {.callout-note collapse="true"}
## Mathematical definition

Each data source is observed through its own likelihood on the expected signal $\mu^{(k)}_t$:
$$
C_t \sim \mathrm{NegBinomial}\big(\mu^{(\text{cases})}_t,\ \phi\big), \qquad D_t \sim \mathrm{Poisson}\big(\mu^{(\text{deaths})}_t\big),
$$
$$
\log W_t \sim \mathrm{Normal}\big(\log \mu^{(\text{ww})}_t,\ \sigma_{ww}\big),
$$
for cases $C_t$, deaths $D_t$ and wastewater concentration $W_t$.
:::

Because the data sources are conditionally independent given the infections, whichever terms are switched on are simply *added* together to form the joint log-likelihood.

::: {.callout-tip}
## Take 5 minutes

Looking at the pieces above (and comparing with the random-walk model `estimate-inf-and-r-rw` from the [previous session](R-estimation-and-the-renewal-equation#estimating-r_t-with-a-geometric-random-walk-prior)): which parts are shared across all three data sources, and which are specific to each data source?
Map each `use_*` flag to a branch of the DAG.
:::

The Stan model always expects data for all three data sources; the `use_*` flags decide which ones actually enter the likelihood.
So that we can build up the data the same way we build up the model, we start with a small helper that assembles the data list for whichever data sources we have simulated so far, filling any data source we have not yet introduced with harmless placeholders (they never enter the likelihood while its `use_*` flag is `0`).
The helper also carries each data source's scaling prior as a `(mean, sd)` pair.
These default to *relaxed* priors we use for the joint fits; we pass *tight* priors when we fit a single data source on its own, for a reason we come to below.

```{r data_helper}
make_data <- function(cases, deaths, ww,
                      case_delay_pmf, death_delay_pmf, ww_delay_pmf,
                      ascertainment_p = c(0.4, 0.06),
                      ifr_p = c(0.1, 0.02),
                      ww_scale_p = c(2, 0.3),
                      use_nb_cases = 1) {
  list(
    n = length(cases),
    I0 = I0,
    gen_time_max = length(gen_time_pmf),
    gen_time_pmf = gen_time_pmf,
    cases = cases,
    case_delay_max = length(case_delay_pmf) - 1,
    case_delay_pmf = case_delay_pmf,
    deaths = deaths,
    death_delay_max = length(death_delay_pmf) - 1,
    death_delay_pmf = death_delay_pmf,
    ww = ww,
    ww_delay_max = length(ww_delay_pmf) - 1,
    ww_delay_pmf = ww_delay_pmf,
    ascertainment_p = ascertainment_p,
    ifr_p = ifr_p,
    ww_scale_p = ww_scale_p,
    use_nb_cases = use_nb_cases,
    tv_death_scale = 0
  )
}
```

A second helper switches the requested data sources on and fits the model.
We provide initial values to help the sampler, and expose `adapt_delta`: the multi-source *joint* fits have a more difficult posterior, so for those we raise `adapt_delta` to `0.95`, the same lever used in the [joint nowcasting session](joint-nowcasting).
The single-source diagnostic fits are easier, so they use the course-standard default.
To keep this session quick to run we fit **2 chains** and a shorter warmup (`iter_warmup = 250`, keeping `iter_sampling = 500`).

```{r fit_helper}
fit_streams <- function(data, use_cases, use_deaths, use_ww,
                        adapt_delta = 0.8, iter_warmup = 250) {
  data <- c(data, list(
    use_cases = use_cases, use_deaths = use_deaths, use_ww = use_ww
  ))
  nfidd_sample(
    mod, data = data,
    chains = 2, parallel_chains = 2,
    iter_warmup = iter_warmup, iter_sampling = 500,
    adapt_delta = adapt_delta,
    max_treedepth = 12,
    init = \() list(init_R = 1, rw_sd = 0.01)
  )
}
```

## Part 1: simulate and fit cases and deaths on their own

*Workflow stage: observation model per source.*
The first thing the workflow asks of us, once the shared process is fixed, is an observation model *per source*.
We start with the two count data sources.
Each is a convolution of the shared infections with its own delay distribution, scaled by a fraction of infections, observed as counts:

-   **Cases**: a short infection-to-report delay, scaled by an *ascertainment* fraction (most infections are never confirmed), and reported with noise beyond the Poisson, so we simulate them as *negative binomial*.
-   **Deaths**: a long infection-to-death delay, scaled by the *infection fatality ratio* (IFR), simulated as Poisson.

```{r simulate_cd}
case_delay_pmf  <- censored_delay_pmf(rgamma, max = 10, shape = 3, rate = 1)
death_delay_pmf <- censored_delay_pmf(rgamma, max = 12, shape = 8, rate = 1)

ascertainment <- 0.4
ifr <- 0.1
## negative-binomial size: smaller = noisier cases. At size 20 the case
## variance is a few times the Poisson variance at the peak, so the cases
## are clearly overdispersed but still carry the signal.
case_dispersion <- 20

exp_cases  <- ascertainment * convolve_with_delay(infections, case_delay_pmf)
exp_deaths <- ifr * convolve_with_delay(infections, death_delay_pmf)

obs <- inf_df |>
  mutate(
    cases  = rnbinom(n, mu = exp_cases, size = case_dispersion),
    deaths = rpois(n, exp_deaths)
  )
```

We can plot the two simulated count data sources, the same way we plotted simulated onsets in earlier sessions.
Each is a delayed, rescaled echo of the same infection curve: deaths peak later than cases.
Cases carry reporting noise beyond the Poisson, which the negative binomial adds, so the case series scatters further around its expectation than Poisson counts of the same magnitude would.

```{r plot_cd}
obs |>
  select(day, cases, deaths) |>
  pivot_longer(c(cases, deaths), names_to = "stream", values_to = "value") |>
  ggplot(aes(x = day, y = value)) +
  geom_col() +
  facet_grid(stream ~ ., scales = "free_y") +
  labs(title = "Simulated cases and deaths", x = "Day", y = "Count")
```

### Prior predictive check

*Workflow stage: prior predictive check.*
Before we fit anything, the workflow asks us to check that our *priors* alone can generate data in the right ballpark, not orders of magnitude too large or too small.
We sample from the model with every likelihood switched off (`use_* = 0`), so the draws come only from the priors, and compare the implied cases and deaths with the data.

```{r prior_predictive, results = 'hide', message = FALSE}
## every likelihood off: the draws come from the priors alone
ww_placeholder_pmf <- censored_delay_pmf(rgamma, max = 10, shape = 2, rate = 1)
prior_data <- make_data(
  cases = obs$cases, deaths = obs$deaths, ww = rep(0, n),
  case_delay_pmf = case_delay_pmf,
  death_delay_pmf = death_delay_pmf,
  ww_delay_pmf = ww_placeholder_pmf
)
fit_prior <- fit_streams(prior_data, use_cases = 0, use_deaths = 0, use_ww = 0)
```

```{r plot_prior_predictive}
prior_obs <- bind_rows(
  obs |> transmute(day, source = "cases", value = cases),
  obs |> transmute(day, source = "deaths", value = deaths)
)
## the geometric random-walk prior lets infections wander a long way, so the
## prior predictive 90% upper runs into the thousands; clip it to twice the
## observed maximum for display -- the band really extends far higher
caps <- prior_obs |> group_by(source) |> summarise(cap = 2 * max(value))

prior_pp <- bind_rows(
  fit_prior |> gather_draws(pp_cases[day])  |> mutate(source = "cases"),
  fit_prior |> gather_draws(pp_deaths[day]) |> mutate(source = "deaths")
) |>
  ungroup() |>
  ## the most explosive prior draws overflow the count RNG to NA; drop them
  filter(is.finite(.value)) |>
  group_by(source, day) |>
  median_qi(.value, .width = 0.9) |>
  left_join(caps, by = "source") |>
  mutate(across(c(.value, .lower, .upper), \(x) pmin(x, cap)))

ggplot(prior_pp, aes(x = day, y = .value)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), fill = "#9ecae1",
              alpha = 0.7) +
  geom_line(colour = "#08519c", linewidth = 0.8) +
  geom_point(data = prior_obs, aes(y = value), colour = "red", size = 0.8) +
  facet_wrap(~source, scales = "free_y") +
  labs(title = "Prior predictive check: cases and deaths",
       subtitle = paste("Prior median and 90% band (blue, clipped at top --",
                        "it extends far higher) vs simulated data (red)"),
       x = "Day", y = NULL)
```

The data sit comfortably inside the prior band, so the priors are not obviously mis-scaled.
The band is very wide because the random-walk prior on $R_t$ permits a lot of growth before any data constrain it, which is itself worth knowing before we fit.

::: {.callout-note}
## Do this at every step

A prior predictive check belongs at *every* step of the build, for each new sub-model.
We show it once here, for the two count sources, for brevity.
:::

### Checking each data source on its own

*Workflow stage: validate each sub-model independently.*
Before combining anything, the workflow tells us to check each sub-model on its own: can each data source, through its own delay and scaling, recover the shared infection process?
This is the modular pay-off from breaking the DAG into components: we validate one branch at a time before we link them.

There is a subtlety here worth being explicit about, because it drives how we fit these single data sources.
A single count data source only ever tells us about infections *multiplied by* its scaling: doubling the infections and halving the ascertainment gives exactly the same expected cases.
So a data source identifies the *trend* of infections (and hence $R_t$) but **not their absolute level**.
This is a property of the model rather than of having only one data source: as we will see, adding more data sources does not break the invariance, because each new data source arrives with its own free scaling.
To recover the level from one data source we have to inform its scaling.
We therefore give each single-source diagnostic fit a *tight* prior on that data source's scaling, centred near the truth, as a deliberate device: it lets us check that the data source recovers the infection level, not just its shape.

```{r fit_single, results = 'hide', message = FALSE}
## placeholder wastewater: never enters the likelihood while use_ww = 0
ww_placeholder_pmf <- censored_delay_pmf(rgamma, max = 10, shape = 2, rate = 1)
cd_data <- make_data(
  cases = obs$cases, deaths = obs$deaths, ww = rep(0, n),
  case_delay_pmf = case_delay_pmf,
  death_delay_pmf = death_delay_pmf,
  ww_delay_pmf = ww_placeholder_pmf
)

## tight, informed priors on each data source's own scaling anchor the level
cases_data <- cd_data
cases_data$ascertainment_p <- c(0.4, 0.03)
deaths_data <- cd_data
deaths_data$ifr_p <- c(0.1, 0.01)

fit_cases <- fit_streams(
  cases_data, use_cases = 1, use_deaths = 0, use_ww = 0
)
fit_deaths <- fit_streams(
  deaths_data, use_cases = 0, use_deaths = 1, use_ww = 0
)
```

We extract the estimated infections from each single-source fit and compare them to the truth.
With the scaling informed, each data source now recovers the infection *level*, not just the trend.

```{r plot_single}
single_inf <- bind_rows(
  fit_cases  |> gather_draws(infections[day]) |> mutate(model = "cases only"),
  fit_deaths |> gather_draws(infections[day]) |> mutate(model = "deaths only")
) |>
  ungroup() |>
  sample_draws(100)

ggplot(single_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  facet_wrap(~model) +
  labs(title = "Infections from single-source fits (grey) vs truth (red)",
       x = "Day", y = "Infections")
```

The reproduction number is recovered from each data source too.
Unlike the level, $R_t$ is a property of the *trend*, so each data source identifies it without any help from the scaling prior.
We plot each single-source $R_t$ estimate; we do not draw a "true" $R_t$ line, because the quantity we are recovering is the latent $R_t$ and we compare estimates to each other rather than to a known truth.

```{r plot_single_r}
single_r <- bind_rows(
  fit_cases  |> gather_draws(R[day]) |> mutate(model = "cases only"),
  fit_deaths |> gather_draws(R[day]) |> mutate(model = "deaths only")
) |>
  ungroup() |>
  sample_draws(100)

ggplot(single_r, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~model) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt from single-source fits (sampled trajectories)",
       x = "Day", y = "Rt")
```

::: {.callout-tip}
## Take 2 minutes

Compare the two single-source fits.
Which data source pins down the *recent* part of the infection curve (the right-hand end) more tightly, and why?
What does the long infection-to-death delay do to the deaths-only fit near the present?
:::

::: {.callout-note collapse="true"}
## Solution

The cases-only fit constrains the recent infections relatively well, because the infection-to-report delay is short, so recent infections still feed into observed cases.
The deaths-only fit is much more uncertain near the present: with a long infection-to-death delay, recent infections have barely begun to show up in deaths, so the most recent infections are informed almost entirely by the prior.
The longer, more dispersed death delay also blurs *sharp changes*, so turning points (when infections peak and turn over) are timed less precisely from deaths than from the timelier cases.
:::

## Part 2: link cases and deaths through the shared infections

*Workflow stage: data integration choice.*
Each data source works on its own, so we can now make the **data integration choice** from the workflow: link the two of them.
Both cases and deaths look directly at the *same* latent infections, each through its own delay and scaling, and their likelihood terms are added.
With two data sources tied to one shared infection level, two scaling priors now bear on that level instead of one, so we can drop the tight single-source priors and fall back to the *relaxed* scaling priors baked into `make_data` rather than pinning any single scaling tightly.

```{r fit_two, results = 'hide', message = FALSE}
fit_cd <- fit_streams(
  cd_data, use_cases = 1, use_deaths = 1, use_ww = 0, adapt_delta = 0.95
)
```

By linking the timely cases with the stable (if delayed) deaths, the two data sources together should constrain the infection trajectory better than either alone.

```{r plot_two}
two_inf <- fit_cd |>
  gather_draws(infections[day]) |>
  ungroup() |>
  sample_draws(100)

ggplot(two_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  labs(title = "Infections, cases + deaths jointly (grey) vs truth (red)",
       subtitle = "Two data sources sharing one latent infection process")
```

The two-source fit gives us $R_t$ as well, now informed by both data sources at once.

```{r plot_two_r}
two_r <- fit_cd |>
  gather_draws(R[day]) |>
  ungroup() |>
  sample_draws(100)

ggplot(two_r, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt, cases + deaths jointly (sampled trajectories)",
       x = "Day", y = "Rt")
```

## Part 3: introduce wastewater and fit it on its own

So far both data sources have been counts of people.
We now introduce a third, very different data source: **wastewater**.

Wastewater surveillance measures the concentration of pathogen genetic material (e.g. viral RNA) shed into the sewage system.
Because it samples whatever is in the sewer, it is a *population-level* signal that does not depend on people seeking testing or care, and it picks up infections regardless of whether they are ever ascertained as cases.
That makes it an attractive complement to the count data sources: it can lead case reports, captures asymptomatic infection, and is relatively cheap to collect at the scale of a whole catchment.

We model it in a deliberately simple way that mirrors the other data sources: the signal is taken to be **proportional to recent infections**, via a short infection-to-shedding delay and a single scaling factor, and observed on the **log scale with log-normal noise** (a single site, with no normalisation or site effects).

::: {.callout-note collapse="true"}
## What this wastewater model skips over

Our one-delay, one-scaling, log-normal model is a deliberate caricature.
A realistic wastewater model has to contend with at least the following, all of which we fold into that single delay and scale.

-   **Shedding kinetics.** How much RNA an infected person sheds changes over the course of their infection, so the infection-to-signal map is really a shedding *load curve* rather than a single delay.
-   **Catchment and connectivity.** Only part of the population drains to the sampled point, and that fraction is rarely known.
-   **Flow and dilution.** Concentration depends on wastewater volume, which shifts with rainfall and industrial input, so signals are usually normalised (e.g. by flow or by a faecal marker such as PMMoV).
-   **In-sewer decay and travel time.** RNA degrades and disperses between shedding and the sampling point, adding further delay and loss.
-   **Sampling and assay.** Grab versus composite sampling, extraction efficiency, PCR noise and a limit of detection all shape the measurement.
-   **Multiple sites.** Several catchments need site effects and spatial pooling rather than one shared scale.

Dedicated tools build these steps explicitly.
See `EpiSewer` [@lisonRobustRealtimeEstimation2025; @lisonEpiSewer2025] and the CDC `ww-inference-model` [@johnsonWwinference2026], a fuller version of this kind of model that jointly fits wastewater and hospital admissions from a hierarchical renewal process [@johnsonBayesianGenerative2026], to go further.
:::

This is the third data source's observation node.
It is the same delay-scaling-then-observe pattern as the count data sources, but with a log-normal likelihood rather than Poisson:

```stan
if (use_ww) {
  ww ~ normal(log(exp_ww), ww_sigma); // <1>
}
```
1. Wastewater: the log concentration is normal around the log of the expected signal, i.e. the signal itself is log-normal.

```{r simulate_ww}
ww_delay_pmf <- censored_delay_pmf(rgamma, max = 10, shape = 2, rate = 1)
ww_scale <- 2

exp_ww <- ww_scale * convolve_with_delay(infections, ww_delay_pmf)
ww_obs_sd <- 0.35   # log-scale observation noise: larger = noisier signal
obs <- obs |>
  mutate(ww = rnorm(n, log(exp_ww), ww_obs_sd))
```

::: {.callout-note}
## Take 2 minutes

The wastewater data source is modelled on the *log* scale with log-normal noise, whereas cases and deaths are counts, observed as negative-binomial and Poisson draws respectively.
Why might a log-normal observation model be a more natural choice for a continuous concentration measurement than a count distribution?
:::

::: {.callout-note collapse="true"}
## Solution

A wastewater concentration is a positive, continuous measurement rather than a count of discrete events, so a count distribution does not apply directly.
Measurement error in such assays tends to be multiplicative (proportional to the signal) rather than additive, which a log-normal observation model captures well.
Dedicated wastewater tools build a much richer shedding and measurement model than ours; see `EpiSewer` [@lisonRobustRealtimeEstimation2025; @lisonEpiSewer2025] and the CDC `ww-inference-model` [@johnsonWwinference2026] to go further.
:::

*Workflow stage: validate each sub-model independently.*
As with the other data sources, we first fit wastewater on its own to check it can recover the shared infections through its delay and scaling.
We rebuild the data, now with the real wastewater data source, and give the single-source fit the same treatment as the counts: a tight prior on its scaling so it recovers the infection level and not only the trend.

```{r fit_ww, results = 'hide', message = FALSE}
full_data <- make_data(
  cases = obs$cases, deaths = obs$deaths, ww = obs$ww,
  case_delay_pmf = case_delay_pmf,
  death_delay_pmf = death_delay_pmf,
  ww_delay_pmf = ww_delay_pmf
)
ww_data <- full_data
ww_data$ww_scale_p <- c(2, 0.15)   # tight, informed prior on the ww scaling
fit_ww <- fit_streams(ww_data, use_cases = 0, use_deaths = 0, use_ww = 1)
```

```{r plot_ww}
ww_inf <- fit_ww |>
  gather_draws(infections[day]) |>
  ungroup() |>
  sample_draws(100)

ggplot(ww_inf, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  labs(title = "Infections, wastewater only (grey) vs truth (red)",
       subtitle = "A short-delay data source on its own scale")
```

## Part 4: add wastewater to the model

*Workflow stage: data integration choice.*
Adding a data source is now just another integration step: switch wastewater on alongside cases and deaths.
We again use the relaxed scaling priors from `make_data`: with three data sources tied to one shared infection level, we no longer need to pin any *single* scaling tightly.

```{r fit_all, results = 'hide', message = FALSE}
fit <- fit_streams(
  full_data, use_cases = 1, use_deaths = 1, use_ww = 1, adapt_delta = 0.95
)
```

```{r fit_summary}
## focus on the scalings and a few Rt values so we can compare their precision
fit$summary(c("ascertainment", "ifr", "ww_scale", "R[1]", "R[20]", "R[41]"))
```

We extract the posterior for the shared infections and reproduction number from the full three-source fit.

```{r posteriors}
posteriors <- fit |>
  gather_draws(infections[day], R[day]) |>
  ungroup() |>
  sample_draws(100)
```

The infections are now informed by all three data sources at once:

```{r plot_inf}
inf_posterior <- posteriors |>
  filter(.variable == "infections")
ggplot(inf_posterior, aes(x = day)) +
  geom_line(aes(y = .value, group = .draw), alpha = 0.1) +
  geom_line(data = inf_df, aes(y = infections), colour = "red") +
  labs(title = "Infections, estimated (grey) and true (red)",
       subtitle = "Joint model fit to cases, deaths and wastewater")
```

and so is the reproduction number, which we come to next.

## Comparing $R_t$ across data sources

*Workflow stage: model checking.*
We have now fitted $R_t$ from each data source on its own and from all three together.
Filling in the two pairs we have not fitted yet lets us put every combination on one figure and see what each addition buys.

```{r fit_pairs, results = 'hide', message = FALSE}
fit_cw <- fit_streams(
  full_data, use_cases = 1, use_deaths = 0, use_ww = 1, adapt_delta = 0.95
)
fit_dw <- fit_streams(
  full_data, use_cases = 0, use_deaths = 1, use_ww = 1, adapt_delta = 0.95
)
```

Each single data source identifies $R_t$ (a trend) on its own, but with different strengths: cases and wastewater are timely and sharpen the recent end, deaths are stable but blur it, and the joint fits borrow from all of them.

```{r plot_r, fig.width = 10, fig.height = 7}
r_by_model <- bind_rows(
  fit_cases  |> gather_draws(R[day]) |> mutate(model = "cases only"),
  fit_deaths |> gather_draws(R[day]) |> mutate(model = "deaths only"),
  fit_ww     |> gather_draws(R[day]) |> mutate(model = "wastewater only"),
  fit_cd     |> gather_draws(R[day]) |> mutate(model = "cases + deaths"),
  fit_cw     |> gather_draws(R[day]) |> mutate(model = "cases + wastewater"),
  fit_dw     |> gather_draws(R[day]) |> mutate(model = "deaths + wastewater"),
  fit        |> gather_draws(R[day]) |> mutate(model = "all three")
) |>
  ungroup() |>
  sample_draws(100) |>
  mutate(model = factor(model, levels = c(
    "cases only", "deaths only", "wastewater only",
    "cases + deaths", "cases + wastewater", "deaths + wastewater",
    "all three"
  )))

ggplot(r_by_model, aes(x = day, y = .value, group = .draw)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~model, nrow = 3) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt from each single source, each pair, and the joint fit",
       subtitle = "Sampled trajectories from each fit",
       x = "Day", y = "Rt")
```

## Checking the joint fit against every data source

*Workflow stage: model checking.*
Before we trust the joint fit we should check it reproduces each data source it was fitted to.
The model gives us posterior predictive draws for each data source (`pp_cases`, `pp_deaths`, `pp_ww`): simulated observations *with* each data source's own observation noise, so we can compare them like for like with the data.
We overlay sampled predictive trajectories for the shared infections and all three data sources against the simulated data.

```{r ppc_streams}
levels_ppc <- c("infections", "cases", "deaths", "wastewater")
ppc_streams <- bind_rows(
  fit |> gather_draws(infections[day]) |> mutate(stream = "infections"),
  fit |> gather_draws(pp_cases[day])   |> mutate(stream = "cases"),
  fit |> gather_draws(pp_deaths[day])  |> mutate(stream = "deaths"),
  fit |> gather_draws(pp_ww[day])      |> mutate(stream = "wastewater")
) |>
  ungroup() |>
  group_by(stream) |>
  sample_draws(100) |>
  ungroup() |>
  mutate(stream = factor(stream, levels = levels_ppc))

observed_streams <- bind_rows(
  inf_df |> transmute(day, stream = "infections", value = infections),
  obs    |> transmute(day, stream = "cases", value = cases),
  obs    |> transmute(day, stream = "deaths", value = deaths),
  obs    |> transmute(day, stream = "wastewater", value = ww)
) |>
  mutate(stream = factor(stream, levels = levels_ppc))

ggplot(ppc_streams, aes(x = day, y = .value)) +
  geom_line(aes(group = .draw), alpha = 0.1) +
  geom_point(data = observed_streams, aes(y = value),
             colour = "red", size = 0.6) +
  facet_wrap(~stream, scales = "free_y") +
  labs(title = "Posterior predictive check: infections and each data source",
       subtitle = paste("Sampled trajectories with observation error (grey)",
                        "vs simulated data (red)"),
       x = "Day", y = NULL)
```

Each data source's predictive draws should bracket its data (the infections panel shows the latent trajectory against the truth we simulated), which tells us the shared infection process is compatible with all three data sources at once.

::: {.callout-note}
## Check as you build, not only at the end

We run this posterior predictive check once, on the full model, for brevity.
In a real workflow it belongs at *every* step, after each single-source fit and after linking cases + deaths, so a poor fit surfaces as early as possible.
A failed check is the workflow's feedback arrow in action, sending you *back* to revise that source's observation model before adding the next one.
:::

::: {.callout-tip}
## Take 2 minutes

Compare the scaling parameters (`ascertainment`, `ifr`, `ww_scale`) in the summary above with the priors we set for them (`c(0.4, 0.06)`, `c(0.1, 0.02)`, `c(2, 0.3)`).
How much has the data narrowed each one?
Why might the *absolute level* of infections be harder to pin down than the *trajectory* of $R_t$?
:::

::: {.callout-note collapse="true"}
## Solution

<!-- verified against the rendered fit_summary -->
The data narrows the scalings only modestly.
Their posteriors (`ascertainment` 0.41 ± 0.04, `ifr` 0.11 ± 0.01, `ww_scale` 1.86 ± 0.19) are roughly a third to a half tighter than the priors (sd 0.06, 0.02, 0.3), so they stay pinned largely by prior information rather than by the data.

- **Why the level is weak.** Each data source constrains only infections *times* its scaling. Multiply the infections by a constant $c$ and divide *every* scaling by $c$, and each data source's expected signal is unchanged, so the *shape* (and hence $R_t$) is identified but the *absolute level* is not.
- **What pins the level here.** Only prior information: the scaling priors, the prior on the seeded starting level `seed_base`, and the `<lower = 0, upper = 1>` bounds on `ascertainment` and `ifr`. Since $I \to cI$ is exactly `seed_base` $\to c \times$ `seed_base`, that seeding prior constrains the level directly.
- **Adding sources does not fix it.** Each new data source brings its own free scaling to absorb $c$, so the invariance survives. The joint fit buys *more prior information* on the level (three scaling priors, not one), not identification.
- **Why we still recover it here.** Every scaling prior is centred on its simulated value and `seed_base ~ normal(I0, I0)` on the true starting infections, so we are slightly cheating. On real data the level is only ever as good as the priors you can justify.
:::

# Real-time snapshots at day 20 and day 40

*Workflow stage: model checking (fitness for real-time use).*
Everything above used the *whole* outbreak.
In practice $R_t$ estimation happens *during* an outbreak, when the data only run up to today.
As in the [renewal equation session](R-estimation-and-the-renewal-equation#estimating-r_t-in-real-time) and the [nowcasting session](nowcasting), we can mimic this by truncating the data sources at a `cutoff` and refitting, so each fit sees only what would have been available at that point.

We take two snapshots: **day 20**, during the growth phase, and **day 40**, around the peak.
We treat the **day-40** estimate as our *retrospective goal*: with more of the outbreak observed it is the better-anchored estimate, so the question is how close each data source's **real-time day-20** estimate gets to that later, more complete view.
At each cutoff we fit every *single* data source on its own and every *pair* of data sources, each running only up to its own cutoff, so we can see how each combination behaves in real time.
Unlike the single-source *level* checks above, these snapshots use the standard (relaxed) scaling priors: $R_t$ is a trend and is identified regardless of the scaling, so we do not need to inform the scale here.
A small helper truncates the data and fits, reusing the same compiled model on the shorter series.

```{r realtime_helper}
fit_at_cutoff <- function(cutoff, use_cases, use_deaths, use_ww, ...) {
  keep <- seq_len(cutoff)
  d <- make_data(
    cases = obs$cases[keep], deaths = obs$deaths[keep], ww = obs$ww[keep],
    case_delay_pmf = case_delay_pmf,
    death_delay_pmf = death_delay_pmf,
    ww_delay_pmf = ww_delay_pmf
  )
  fit_streams(d, use_cases, use_deaths, use_ww, ...)
}
```

```{r realtime_fits, results = 'hide', message = FALSE}
cutoffs <- c(20, 40)
## three single sources and the three pairwise combinations
model_specs <- list(
  "cases only"          = c(1, 0, 0),
  "deaths only"         = c(0, 1, 0),
  "wastewater only"     = c(0, 0, 1),
  "cases + deaths"      = c(1, 1, 0),
  "cases + wastewater"  = c(1, 0, 1),
  "deaths + wastewater" = c(0, 1, 1)
)

rt_snapshots <- map_dfr(cutoffs, function(ct) {
  map_dfr(names(model_specs), function(nm) {
    f <- model_specs[[nm]]
    ad <- if (sum(f) > 1) 0.95 else 0.8
    fit_at_cutoff(ct, f[1], f[2], f[3], adapt_delta = ad) |>
      gather_draws(R[day]) |>
      ungroup() |>
      group_by(day) |>
      median_qi(.value, .width = 0.9) |>
      mutate(model = nm)
  }) |>
    mutate(cutoff = ct)
}) |>
  mutate(
    model = factor(model, levels = names(model_specs)),
    cutoff_label = factor(
      if_else(cutoff == 20, "real-time (day 20)", "retrospective (day 40)"),
      levels = c("real-time (day 20)", "retrospective (day 40)")
    )
  )
```

We put both snapshots on one figure, coloured by cutoff and faceted into a 2 x 3 grid of single sources (top row) and pairs (bottom row), so that within each panel we can read the real-time (day 20) estimate against the retrospective (day 40) goal.

```{r realtime_plot, fig.width = 10, fig.height = 6}
ggplot(rt_snapshots, aes(x = day, y = .value,
                         colour = cutoff_label, fill = cutoff_label)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~model, nrow = 2) +
  scale_colour_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Real-time (day 20) vs retrospective (day 40) Rt",
       subtitle = paste("Median and 90% CI; each fit uses only data up to its",
                        "cutoff, with day 40 the retrospective goal"),
       x = "Day", y = "Rt", colour = NULL, fill = NULL) +
  theme(legend.position = "bottom")
```

The grid leaves out the full model, so we fit all three data sources together at each cutoff and show it on its own.

```{r realtime_joint, results = 'hide', message = FALSE}
rt_joint <- map_dfr(cutoffs, function(ct) {
  fit_at_cutoff(ct, 1, 1, 1, adapt_delta = 0.95) |>
    gather_draws(R[day]) |>
    ungroup() |>
    group_by(day) |>
    median_qi(.value, .width = 0.9) |>
    mutate(cutoff = ct)
}) |>
  mutate(cutoff_label = factor(
    if_else(cutoff == 20, "real-time (day 20)", "retrospective (day 40)"),
    levels = c("real-time (day 20)", "retrospective (day 40)")
  ))
```

```{r realtime_joint_plot}
ggplot(rt_joint, aes(x = day, y = .value,
                     colour = cutoff_label, fill = cutoff_label)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Real-time vs retrospective Rt: full three-source model",
       x = "Day", y = "Rt", colour = NULL, fill = NULL) +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
## Take 5 minutes

The grid shows six models, three single sources (top row) and three pairs (bottom row), at both cutoffs, and the plot below adds the full three-source model.
In each panel compare the real-time (day 20) band with the retrospective (day 40) band over their overlapping days.

- Which *single* source's day-20 estimate lands closest to its day-40 goal?
- Which *pair* does best, and what does adding a second source do to the day-20 estimate?
- Does the full three-source model beat the best pair in real time?
:::

::: {.callout-note collapse="true"}
## Solution

<!-- verified against the rendered 2x3 figure -->
The deaths do carry signal at day 20, but it is *old* signal.
Over the first ten days or so the deaths-only day-20 estimate sits on top of its day-40 goal.
It is the *recent* end that the deaths cannot see: with a long infection-to-death delay, infections from the last week have barely begun to cause deaths, so the estimate keeps $R_t$ near the level the earlier data implied instead of turning it over.

- **Single sources.** Cases and wastewater have short delays, so their day-20 estimates track the day-40 goal over almost the whole overlapping window. The deaths-only day-20 estimate stays around 1.4 to 1.5 and flat, while the day-40 fit shows $R_t$ already back near 1 by day 20, so deaths miss the turn.
- **Pairs.** Every pair here contains a timely source, so all three track the goal reasonably. Cases + wastewater is the closest match and the tightest, which is what we would expect from two short-delay sources.
- **What a second source buys.** Pairing deaths with a timely source (cases + deaths, deaths + wastewater) pulls the day-20 estimate off the deaths-only plateau and back onto the retrospective trajectory. The gain is largest where the single source was weakest, at the recent end.
- **Does the full model beat the best pair?** Not by much. The three-source day-20 estimate sits on its day-40 goal about as closely as cases + wastewater does, and it dips a little below the goal over the last few days. Once two timely sources are in the model, adding the delayed one has little left to say about the present, which is worth remembering before paying to add a data source.
:::

# When data sources conflict

Combining data sources works well when they tell a consistent story, but real surveillance data sources can *disagree*.
This is the feedback arrow in the workflow that we singled out earlier: *conflict between data sources reveals model misspecification*.

## What does it mean for data sources to conflict?

Because every data source is a delayed, scaled view of the *same* latent infections, each implicitly carries its own answer to the question "what must the infection trajectory have looked like?".
Two data sources **conflict** when those implied trajectories are incompatible: the cases imply infections that are falling while the deaths imply infections that are still rising, say, and no single $R_t$ path can produce both at once.

Why does this happen in practice?
The model assumes each data source is the shared infections convolved with a *fixed* delay and multiplied by a *constant* scaling.
Conflict is what we see when one of those assumptions is wrong for at least one data source, for example:

- **Changing ascertainment.** A change in testing policy or test availability bends the case curve without any change in transmission, so the constant-ascertainment assumption breaks.
- **Mis-specified delays.** If the true infection-to-death delay is longer or more dispersed than the PMF we supplied, deaths will appear shifted relative to the infections the other data sources imply.
- **Mis-specified scalings.** A drifting infection fatality ratio (e.g. with changing age structure or treatment) or a reporting artefact pushes a data source off the level the others support.
- **Truly different populations.** Streams that sample different groups (e.g. hospital deaths vs community cases) need not share one infection process at all.

To see conflict in action we give it a concrete **mechanism**: a *drifting infection fatality ratio*.
We re-simulate the deaths from the *same* shared infections as before, but let the death scaling rise over the second half of the outbreak (as it might if severity increased with age structure or a new variant).
Cases and wastewater are unchanged.
Under our constant-scaling model the rising deaths therefore *look like* a rising epidemic, which the flat-scaling cases and wastewater flatly contradict.

```{r simulate_conflict}
## death scaling drifts up over the second half of the outbreak
ifr_drift <- ifr * c(rep(1, 20), seq(1, 4, length.out = n - 20))
exp_deaths_conflict <-
  ifr_drift * convolve_with_delay(infections, death_delay_pmf)

obs_conflict <- obs |>
  mutate(deaths = rpois(n, exp_deaths_conflict))
```

## Fitting the constant-scaling model to conflicting data sources

We refit the same joint, three-source model -- still assuming a *constant* death scaling -- to this conflicting data, reusing the helpers and just swapping in the conflicting deaths.

```{r fit_conflict, results = 'hide', message = FALSE}
data_conflict <- full_data
data_conflict$deaths <- obs_conflict$deaths
fit_conflict <- fit_streams(
  data_conflict, use_cases = 1, use_deaths = 1, use_ww = 1, adapt_delta = 0.95
)
```

We compare the $R_t$ estimates from the consistent and conflicting fits.

```{r plot_conflict}
r_consistent <- fit |>
  gather_draws(R[day]) |>
  ungroup() |>
  sample_draws(100) |>
  mutate(data = "consistent")

r_conflict <- fit_conflict |>
  gather_draws(R[day]) |>
  ungroup() |>
  sample_draws(100) |>
  mutate(data = "conflicting deaths")

ggplot(
  bind_rows(r_consistent, r_conflict),
  aes(x = day, y = .value, group = interaction(.draw, data), colour = data)
) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt under consistent vs conflicting data sources") +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
## Take 2 minutes

How does the $R_t$ estimate change when the deaths data source conflicts with the others?
Is the result closer to the "true" declining $R_t$, the rising signal implied by deaths, or somewhere in between?
:::

::: {.callout-note collapse="true"}
## Solution

<!-- verified against the rendered consistent-vs-conflicting figure -->

- **Direction.** Through the *early* part of the second half the conflicting-deaths fit pulls $R_t$ *upwards*, away from the true declining trajectory, as it tries to accommodate the rising deaths. Later it over-compensates and dips *below* the consistent fit.
- **Why.** One shared infection process cannot be both declining (to match cases and wastewater) and rising (to match deaths), so the model splits the difference and the compromise wanders either side of the truth.
- **Result.** The joint $R_t$ is faithful to no single data source: neither the true declining path nor the rising one the deaths imply.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: when data sources conflict

```{=html}
<iframe class="nfidd-embed" src="../interactive/data-conflict.html?embed=conflict" title="when surveillance data sources conflict interactive" loading="lazy" style="width:100%;height:640px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Drag the death-scaling drift to open a conflict, then let the scaling vary to snap the estimate back onto the truth.
:::

## How the joint model surfaces conflict

The $R_t$ comparison above already hints at trouble, but in a real analysis we would not know the truth to compare against.
So how would conflict actually announce itself?
It shows up in three linked ways, and a principled workflow looks for all of them.

**1. Poor fit to one data source.**
The clearest symptom is a *posterior predictive check*: simulate each data source from the fitted model, *with* its observation noise, and overlay it on the data.
Because the shared infections are dragged towards a compromise, at least one data source ends up fitted badly.
We draw the posterior predictive counts for each data source (`pp_cases`, `pp_deaths`, from each data source's own likelihood) and compare sampled trajectories to the conflicting data.

```{r ppc_conflict}
pp_conflict <- fit_conflict |>
  gather_draws(pp_cases[day], pp_deaths[day]) |>
  ungroup() |>
  group_by(.variable) |>
  sample_draws(100) |>
  ungroup()

obs_long <- obs_conflict |>
  select(day, cases, deaths) |>
  pivot_longer(-day, names_to = "stream", values_to = "value") |>
  mutate(.variable = paste0("pp_", stream))

ggplot(pp_conflict, aes(x = day, y = .value)) +
  geom_line(aes(group = .draw), alpha = 0.1) +
  geom_point(data = obs_long, aes(y = value), colour = "red", size = 0.6) +
  facet_wrap(~.variable, scales = "free_y") +
  labs(title = "Posterior predictive fit under conflicting data sources",
       subtitle = "Sampled predictive trajectories (grey) vs observed (red)",
       x = "Day", y = NULL)
```

The compromise infection trajectory cannot match both data sources at once.
Here the rising deaths pull the shared infections up, so the predictive draws for the *cases* run above the observed cases through the second half of the outbreak while the deaths are still fitted.
A run of points sitting on one side of the predictive draws for a single data source is the sign that that data source is losing the argument with the others.

**2. Tension in the shared latent infections.**
Conflict does not stay neatly inside one data source.
Because all data sources pull on the *same* infections, an irreconcilable data source distorts the shared trajectory itself, and hence $R_t$, as we saw, and inflates its uncertainty as the sampler is pulled between two incompatible explanations.
A latent quantity that suddenly becomes much more uncertain, or that no longer matches *any* data source when you propagate it back through the observation models, is a second symptom of conflict.

**3. Degraded sampling.**
When the geometry of the posterior is contorted by trying to satisfy contradictory data, the sampler can struggle.
That surfaces as divergent transitions, low effective sample sizes, or high `rhat` in the fit summary.
It is the least reliable of the three symptoms, as we are about to see.
It is always worth scanning the diagnostics of the conflicting fit (`fit_conflict`) and comparing them to the clean fit, because sampler complaints are not just a computational nuisance and can be a signature of a misspecified model.

```{r conflict_diagnostics}
diagnostics <- function(f, label) {
  s <- f$summary(c("R", "infections", "ascertainment", "ifr", "ww_scale"))
  tibble(
    fit = label,
    max_rhat = round(max(s$rhat), 3),
    min_ess_bulk = round(min(s$ess_bulk)),
    divergences = sum(f$diagnostic_summary(quiet = TRUE)$num_divergent)
  )
}

bind_rows(
  diagnostics(fit, "consistent"),
  diagnostics(fit_conflict, "conflicting deaths")
)
```

::: {.callout-tip}
## Take 5 minutes

Look at the posterior predictive plot and at the diagnostics printed above.
Which data source is fitted worst, and in which direction does the model miss?
Do the sampler diagnostics look worse than for the consistent fit?
:::

::: {.callout-note collapse="true"}
## Solution

<!-- verified against the rendered ppc_conflict figure and diagnostics table -->

**Cases are fitted worst.**
From about day 20 the observed cases (red) sit *below* the bulk of the predictive draws, and the gap widens through the second half of the outbreak: the model over-predicts the cases.
That is the shared infections being dragged *up* by the rising deaths, which the constant death scaling can only explain as more infections.
The deaths panel, by contrast, is fitted reasonably well, with the latest points sitting at the top edge of the predictive draws rather than outside them.
This is the asymmetry to expect: the source that drives the misspecification gets fitted at the expense of the sources that disagree with it.

The sampler diagnostics barely move.
`rhat` is 1.01 for both fits and neither has any divergent transitions.
The only sign of strain is the effective sample size, which drops from around 830 for the consistent fit to around 550 for the conflicting one.
So it is the posterior predictive check, not the sampler, that exposes this conflict.
Degraded sampling is a useful symptom when it appears, but a model can be badly misspecified and still sample cleanly, which is why the checks in this session do not rely on it.
:::

## What should a modeller do about conflict?

It is worth being clear about what the shared latent process does and does not do here.
Sharing one infection process across data sources *surfaces* a conflict (the symptoms above) and yields a single compromise estimate, but it does not *resolve* the disagreement: the compromise is faithful to no data source and, as we have seen, can be actively misleading about $R_t$.
Real resolution comes from modelling the *reason* the data sources disagree, not from averaging the data sources together.

::: {.callout-note}
## Options for resolving conflict

When a principled model check flags conflict between data sources, there are a handful of standard responses [@presanis2013conflict; @deangelis2018analysing; @gelman2020bayesian]:

- **Model the differing mechanism.** Extend the model so the suspected source of divergence is explicit -- a time-varying ascertainment or IFR, a re-estimated or more dispersed delay, or data source-specific reporting effects. This is the principled default when the data sources really do observe the same infections [@deangelis2018analysing; @nicholson2022interoperability].
- **Expand or relax the observation model.** Allow overdispersion, or weaken an over-confident likelihood, so a data source is no longer forced onto a level it cannot support.
- **Down-weight or drop a data source.** If a data source measures a truly different population, or carries an artefact you cannot model, it may be right to exclude it -- but only once you understand *why* [@presanis2013conflict].
- **Revisit the data.** Conflict is sometimes a data problem (a definition change, a reporting glitch) rather than a model problem, and is best fixed upstream.

Conflict diagnostics in Bayesian evidence synthesis make this formal: they quantify how much each source disagrees with the rest before deciding what to do [@presanis2013conflict].
:::

We will *do* the first of these: model the mechanism behind the divergence.
We suspected a drifting death scaling, so we relax the constant-IFR assumption and let the death scaling follow its own random walk over time.
The model already has a switch for this, `tv_death_scale`; setting it to `1` turns the scalar `ifr` into a geometric random walk.

```{r fit_resolved, results = 'hide', message = FALSE}
data_resolved <- data_conflict
data_resolved$tv_death_scale <- 1
fit_resolved <- fit_streams(
  data_resolved, use_cases = 1, use_deaths = 1, use_ww = 1, adapt_delta = 0.95
)
```

First, does it now fit the deaths?
We repeat the posterior predictive check, comparing the constant-scaling and time-varying-scaling fits.

```{r ppc_resolved}
ppc_for <- function(fit, label) {
  fit |>
    gather_draws(pp_cases[day], pp_deaths[day]) |>
    ungroup() |>
    group_by(.variable) |>
    sample_draws(100) |>
    ungroup() |>
    mutate(model = label)
}

resolved_draws <- bind_rows(
  ppc_for(fit_conflict, "constant scaling"),
  ppc_for(fit_resolved, "time-varying death scaling")
)

ggplot(resolved_draws, aes(x = day, y = .value)) +
  geom_line(aes(group = .draw), alpha = 0.1) +
  geom_point(data = obs_long, aes(y = value), colour = "red", size = 0.6) +
  facet_grid(.variable ~ model, scales = "free_y") +
  labs(title = "Posterior predictive fit: before and after modelling the drift",
       subtitle = "Sampled predictive trajectories (grey) vs observed (red)",
       x = "Day", y = NULL)
```

Letting the death scaling drift lets the deaths panel track the rising data without dragging the shared infections up, so both data sources are fitted well at once.
We can confirm that $R_t$ no longer chases the deaths, by comparing it to the true declining trajectory the cases and wastewater came from.

```{r r_resolved}
r_resolved <- bind_rows(
  fit_conflict |> gather_draws(R[day]) |> mutate(model = "constant scaling"),
  fit_resolved |>
    gather_draws(R[day]) |>
    mutate(model = "time-varying death scaling")
) |>
  ungroup() |>
  sample_draws(100)

ggplot(r_resolved, aes(x = day, y = .value,
                       group = interaction(.draw, model), colour = model)) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(title = "Rt before and after modelling the drifting death scaling") +
  theme(legend.position = "bottom")
```

```{r recovered_scale}
fit_resolved |>
  gather_draws(death_scale[day]) |>
  median_qi(.value, .width = 0.9) |>
  ggplot(aes(x = day, y = .value)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.3) +
  geom_line() +
  geom_line(
    data = tibble(day = seq_len(n), .value = ifr_drift),
    colour = "red"
  ) +
  labs(title = "Recovered death scaling (black) vs truth (red)",
       x = "Day", y = "Death scaling")
```

We can also compare the sampler diagnostics of the two fits.

```{r resolved_diagnostics}
bind_rows(
  diagnostics(fit_conflict, "constant scaling"),
  diagnostics(fit_resolved, "time-varying death scaling")
)
```

::: {.callout-tip}
## Take 5 minutes

Compare the two fits.
Does the time-varying-scaling model fit *both* the cases and the deaths?
Has $R_t$ returned to the true declining trajectory, and are the sampler diagnostics of `fit_resolved` better than `fit_conflict`?
Does the recovered death scaling track the drift we built in?
:::

::: {.callout-note collapse="true"}
## Solution

Modelling the mechanism reconciles the data sources.
With a time-varying death scaling the deaths are fitted well *and* the cases are no longer over-predicted, because the rising deaths are now explained by a rising scaling rather than by rising infections.
The case predictive draws tighten around the data, which is the clearest sign that the compromise has gone.
$R_t$ returns to the declining trajectory the cases and wastewater imply, and the recovered death scaling tracks the drift we built in.
The sampler diagnostics improve, but modestly.
The effective sample size recovers, from around 550 for the constant-scaling fit to around 840 once the drift is modelled, while `rhat` stays at 1.01 and neither fit has divergent transitions.
That is the pattern to expect here: the constant-scaling fit already sampled acceptably, so the clearest evidence that the conflict has gone is the fit to the data, not the sampler.
:::

This loop of building in parts, checking each addition, and letting conflict and identifiability send us back to revise is the core of this modelling workflow [@abbottWorkflowInfectiousDisease2026].

Everything here has been on *simulated* data, where we know the truth and can check that the model recovers it.
That is the point of building and testing on simulated data first.
Once the model is tested and checked this way, the same workflow repeats on real data: the same steps of designing and breaking up the DAG, building in parts, and using model checks to send us back to revise, but now without a known truth to compare against, so the posterior predictive checks and conflict diagnostics from this session do the work of telling us whether the model is adequate.

# Going further

## Scope

This session extended the **renewal-with-delays** model to multiple parallel data sources.
We kept it to renewal plus convolutions and did *not* bring in nowcasting or the reporting triangle: each data source here is assumed fully observed up to the present.
Combining this multi-source approach with the nowcasting models from the [joint nowcasting session](joint-nowcasting) is a natural next step.

## Challenge

-   We used a negative binomial for cases (with an overdispersion parameter) but kept deaths Poisson. Following the same pattern, how would you also make the *deaths* overdispersed, and when might that matter?
-   We resolved the conflict by letting the *death* scaling drift. Following the same pattern as `tv_death_scale`, how would you add a time-varying *ascertainment* to absorb a conflict driven by a change in testing policy? Which data source conflicts would that be able to absorb, and which would it not?

## Methods in practice

-   `epidemia` implements joint renewal-equation models of cases and deaths of the kind introduced here [@scottEpidemiaRPackage2021].
-   `EpiSewer` is a Bayesian generative model for estimating transmission dynamics from a wastewater signal, with a carefully specified shedding and measurement model [@lisonRobustRealtimeEstimation2025; @lisonEpiSewer2025].
-   The CDC's wastewater-informed modelling work jointly fits wastewater concentration data and epidemiological count data to produce nowcasts and forecasts [@johnsonWwinference2026; @johnsonBayesianGenerative2026]. Unlike our deliberately single-site example, it combines *multiple* wastewater sites with hierarchical structure -- a multi-site, multi-resolution extension of the ideas in this session.
-   [`epinowcast`](https://package.epinowcast.org/) is a flexible Bayesian nowcasting framework that can also jointly fit *multiple* datasets (for example several reporting streams, sites or strata), sharing structure across them through partially pooled modules [@epinowcastPackage2024].
-   [`ComposableTuringIDModels.jl`](https://github.com/EpiAware/ComposableTuringIDModels.jl) is a Julia prototype that builds this kind of multi-source model from swappable, composable components, an example of tooling designed to make the modular workflow in this session easier than it is in a single Stan file.
-   The workflow paper itself works through schematic case studies (no code) that apply these steps to real modelling problems, a useful set of worked examples of the workflow we followed here [@abbottWorkflowInfectiousDisease2026].

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
-   Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
